### 2017-11-01 manuscript/mmc_manuscript.tex
Initialize manuscript.

\documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
%\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}

%SetFonts

\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Eq}[1]{Equation~(\ref{#1})}
\newcommand{\fig}[1]{Fig.~\ref{#1}}
\newcommand{\Fig}[1]{Figure~\ref{#1}}

\title{Distinguishing among coalescent models using two-site joint allele frequency spectra}
\author{Daniel P. Rice}
\date{\today}

\begin{document}
\maketitle

\abstract{The genetic diversity of a population reflects its demographic and
evolutionary history. Methods for inferring this history typically
assume that the ancestry of a sample can be modeled by the Kingman
coalescent process. A defining feature of the Kingman coalescent is
that it generates genealogies that are binary trees: no more than two
ancestral lineages may coalesce at the same time. However, this
assumption breaks down under several scenarios. For example, pervasive
natural selection, rapid spatial range expansion, and extreme
variation in offspring number can all generate genealogies with
"multiple-merger" events in which more than two lineages coalesce
instantaneously. Therefore, detecting multiple mergers is important
both for understanding which forces have shaped the diversity of a
population and for avoiding fitting misspecified models to data.
Current methods to detect multiple mergers rely on the average site
frequency spectrum (SFS). However, the signatures of multiple
mergers in the average SFS are also consistent with a Kingman
coalescent process with a time-varying population size. Here, I
present a new method for detecting multiple mergers based on
the mutual information of the joint site frequency spectrum at pairs of linked sites. Unlike the average SFS,
the mutual information depends mostly on the topologies of genealogies
rather than their branch lengths and is therefore robust to most
demographic effects.}

\section*{Introduction}
A typical goal of population genetics is to infer the evolutionary and demographic history of a population from its contemporary genetic diversity. The genetic differences among individuals reflect their genealogical history, which, in turn, reflects the history of various population parameters such as the population size, spatial structure, and the influence of natural selection. By modeling how the (unobserved) genealogical process depends on these parameters and how patterns of genetic diversity depend on genealogy, we can develop inference procedures to learn about the past.

The most commonly-used model is the Kingman coalescent, which arises in many models of neutral evolution.
The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in (continuous) time.
As a result, the genealogies generated by the Kingman coalescent are randomly bifurcating trees: each node subtends exactly two branches and the number of leaves subtended by one of these branches is uniformly distributed.

Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population .
This timescale, which is inversely proportional to the rate of coalescence between pairs of lineages, determines the branch lengths of genealogies and thus determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population. Therefore, it is commonly referred to as the \emph{effective population size}

%The Kingman coalescent generates genealogies with several distinguishing characteristics. First, the genealogies are binary trees: each node has only two children. A second and related topological feature is that for a node subtending $n$ leaves, the number of leaves subtended by one of its children is uniformly distributed between 1 and $n-1$. Finally, under a neutral model of evolution, the rate of coalescence (and thus the branch lengths of the genealogy) is inversely proportional to the population size.

The site frequency spectrum (SFS) is a useful summary statistic of genetic data. Defined as the fraction of mutations as a function of allele frequency, the SFS contains a lot of information about the distribution of underlying genealogies. The number of mutations in $i$ samples is proportional to the lengths of branches subtending $i$ leaves of the genealogy. This quantity depends on the topology and branch lengths of trees. Assuming the Kingman model is true, one can infer the history of the population size from the SFS because the number of mutations at a particular frequency is related to the coalescent rate as a function of time. For example, high-frequency alleles are the result of mutations on deep branches and thus reflect the population size in the distant past.

A serious problem for this inference procedure is that alternative models of evolution generate different coalescent models. For example, models of pervasive weak selection, skewed offspring number distributions, and periodic strong bottlenecks can all generate genealogies that differ from the Kingman both topologically and in terms of branch length.
Collectively, these models are known as \emph{multiple mergers coalescents} because, unlike the Kingman coalescent, they permit multiple lineages to coalesce instantaneously. As a result, multiple-mergers genealogies contain nodes with more than two children and lack the uniform branching structure of the Kingman. Even more important for inference is the fact that the coalescent timescales---and thus the branch lengths and levels of diversity---are determined by parameters other than the population size.
Thus, the results of a Kingman-based inference procedure applied to data generated by a non-Kingman process will be both quantitatively and qualitatively wrong.
Clearly, it would be useful to be able to distinguish among different coalescent models, both to check that popular inferences of historical population sizes are valid and to understand which evolutionary forces are dominant in various populations.

An obvious question is whether we can use the average site frequency spectrum to reject the Kingman? There are two distinguishing features of the SFS generated by multiple mergers coalescents. First, low-frequency mutations are enriched relative to the time-homogeneous Kingman coalescent. Unfortunately, this is also a feature of the Kingman coalescent in growing populations, so it is not useful for model selection. Second, the multiple-merger SFS is non-monotonic: it has a positive slope for high frequencies. Unlike the excess of rare alleles, this feature cannot be reproduced by an exchangeable Kingman coalescent model. However, it is difficult to identify the non-monotonicity in real data. The problem is that it necessary to know which allele is ancestral to distinguish frequency $x$ mutations from frequency $1-x$ mutations. Various methods for identifying the ancestral allele exist, but in practice even a very low rate of error can generate a spurious signal.

Ideally, we would like a statistic (or statistics) that:
\begin{enumerate}
\item is qualitatively different between models of genealogies,
\item is robust to demographic changes, and 
\item requires minimal information beyond genetic diversity.
\end{enumerate}
The first two criteria are important for making qualitative distinctions between models. The last one would maximize its applicability.
Here, I propose that the correlations between elements of the site frequency spectrum are such a statistic.

\end{document}

### 2017-11-03 manuscript/mmc_manuscript.tex
Update manuscript.

@@
 \newcommand{\Fig}[1]{Figure~\ref{#1}}
+\newcommand{\E}[1]{\left<{#1}\right>}
 
@@
 
-\section*{Introduction}
+\section*{Background}
 A typical goal of population genetics is to infer the evolutionary and demographic history of a population from its contemporary genetic diversity.
@@
 
-The most commonly-used model is the Kingman coalescent, which arises in many models of neutral evolution.
+The most commonly used genealogical model is the Kingman coalescent, which arises in many models of neutral evolution.
 The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in (continuous) time.
@@
 
-Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population .
-This timescale, which is inversely proportional to the rate of coalescence between pairs of lineages, determines the branch lengths of genealogies and thus determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
+Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population.
+This timescale, which we will call $T_2$, is inversely proportional to the rate of coalescence between pairs of lineages and sets the branch lengths of genealogies.
+Thus, $T_2$ determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
 Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
 In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population.
-Therefore, it is commonly referred to as the \emph{effective population size}
+Therefore, it is commonly referred to as the effective population size.
 
@@
 
-The site frequency spectrum (SFS) is a useful summary statistic of genetic data.
-Defined as the fraction of mutations as a function of allele frequency, the SFS contains a lot of information about the distribution of underlying genealogies.
-The number of mutations in $i$ samples is proportional to the lengths of branches subtending $i$ leaves of the genealogy.
-This quantity depends on the topology and branch lengths of trees.
-Assuming the Kingman model is true, one can infer the history of the population size from the SFS because the number of mutations at a particular frequency is related to the coalescent rate as a function of time.
-For example, high-frequency alleles are the result of mutations on deep branches and thus reflect the population size in the distant past.
+Because computing the full likelihood of the data is generally intractable, population genetic inference is typically done on informative summary statistics.
+One informative statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in a sample.
+The expected number of mutations in $i$ sampled chromosomes is proportional to the lengths of branches subtending $i$ leaves of the genealogy, and thus depends the distributions of topologies as well as branch lengths.
+By assuming the Kingman model, one can marginalize over the unobserved tree topologies and extract information about the distribution of branch lengths and thus about the rate of coalescence as a function of time.
+If this assumption is good, the site frequency spectrum thus reflects the history of population size changes.
+For example, high-frequency alleles are the result of mutations on deep branches and their number in a sample reflects the population size in the distant past.
 
@@
 Even more important for inference is the fact that the coalescent timescales---and thus the branch lengths and levels of diversity---are determined by parameters other than the population size.
+In this case, the term `effective population size' is misleading because $T_2$ is no longer proportional to the number of individuals in the population.
 Thus, the results of a Kingman-based inference procedure applied to data generated by a non-Kingman process will be both quantitatively and qualitatively wrong.
@@
 
-An obvious question is whether we can use the average site frequency spectrum to reject the Kingman?
+An obvious question is whether we can use the average site frequency spectrum to reject the Kingman model?
 There are two distinguishing features of the SFS generated by multiple mergers coalescents.
 First, low-frequency mutations are enriched relative to the time-homogeneous Kingman coalescent.
-Unfortunately, this is also a feature of the Kingman coalescent in growing populations, so it is not useful for model selection.
+Unfortunately, this is also a feature of the Kingman coalescent in growing populations, so it is not well-suited for model selection (but see \cite{}).
 Second, the multiple-merger SFS is non-monotonic: it has a positive slope for high frequencies.
@@
 However, it is difficult to identify the non-monotonicity in real data.
-The problem is that it necessary to know which allele is ancestral to distinguish frequency $x$ mutations from frequency $1-x$ mutations.
+The problem is that it necessary to know which allele is ancestral to distinguish derived mutations at frequency $x$ from those at frequency $1-x$.
 Various methods for identifying the ancestral allele exist, but in practice even a very low rate of error can generate a spurious signal.
@@
 The first two criteria are important for making qualitative distinctions between models. The last one would maximize its applicability.
-Here, I propose that the correlations between elements of the site frequency spectrum are such a statistic.
+Here, I propose that the joint site frequency spectrum at pairs of linked sites is such a statistic.
+
+\section*{Theoretical motivation}
+
+Consider a sample of $n$ chromosomes.
+Denote the site frequency spectrum, the fraction of sites with a mutation in $i$ of $n$ samples, by $\phi_i$.
+We define the joint site frequency spectrum, $\phi_{ij}(d)$ for pairs of sites separated by a genomic distance of $d$ bases to be the fraction of such pairs with a mutation in $i$ samples at the left site and a mutation in $j$ samples at the right site.
+As described above, the expected site frequency spectrum is related to the expected branch lengths of the genealogies by:
+\begin{equation}
+\E{\phi_i} = \mu \E{\tau_i},
+\end{equation}
+where $\mu$ is the per--base pair mutation rate, $\tau_i$, is the total branch length subtending $i$ leaves, and expectations are taken over the distribution of genealogies.
+Similarly, the joint site frequency spectrum is proportional to the second moments of the joint branch length distribution:
+\begin{equation}
+\E{\phi_{ij}} = \mu^2 \E{\tau_i^l \tau_j^r},
+\end{equation}
+where $\tau_i^l$ and $\tau_j^r$ are the branch lengths at the left and right site respectively.
+There are several previous results that suggest that the correlation structure of branch lengths contains information that can distinguish multiple mergers from Kingman coalescent processes.
+If this is the case, then the joint site frequency spectrum will be a useful summary statistic for coalscent model selection.
+
+\subsubsection*{Tightly-linked sites}
+
+For tightly-linked sites (i.e.
+$rdT_2\ll1$ for per--base pair recombination rate $r$), $\E{\tau_i^l \tau_j^r} \approx \E{\tau_i \tau_j}$, the branch length correlation matrix for a non-recombining locus.
+This correlation matrix depends on correlations among the timing of coalescent events as well as correlations between clade sizes in tree topologies.
+Fu \cite{} calculated the first and second moments of $\tau_i$ for the time-homogeneous Kingman coalescent.
+His results show that $\E{\tau_i \tau_j} - \E{\tau_i}\E{\tau_j} < 0$ for $i \neq j$ and $i \neq n-j$.
+This implies that the off-diagonal elements of the joint site frequency spectrum will be depleted relative to the product of the marginal site frequency spectra for tightly-linked sites.
+Recently, Birkner \textit{et al.} \cite{}, extended Fu's calculation to the class of multiple-mergers coalescent models known as time-homogeneous $\Lambda$-coalescents.
+In contrast to the Kingman case, Birkner \textit{et al.} found that, in the $\Lambda$-coalescent, $\E{\tau_i \tau_j} - \E{\tau_i}\E{\tau_j}$ may be positive for $i \neq j$.
+These positive associations are largest for $1 < i - j \ll n$.
+Together these results suggest that the joint site frequency spectrum of tightly-linked sites can distinguish qualitatively between the constant-sized Kingman and multiple-mergers coalescent.
+
+We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{1}).
+Furthermore, the total branch length is independent of the topology.
+As a result, there is a trade-off between the branch length leading to singleton/tripleton mutations and the branch length leading to doubletons: loci with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
+Conditional on observing a doubleton at a locus, it is thus more likely that the locus has topology (B) and so the expected number of singletons is lower.
+In terms of the joint site frequency spectrum, we have $\E{\phi_{12}} < \E{\phi_{1}} \E{\phi_{2}}$.
+
+On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
+For example, topology (C) has less overall opportunity for singletons as well as doubletons than (A) or (B), even though the expected proportion of singletons is higher.
+Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies more likely.
+If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) and lead to $\E{\phi_{12}} > \E{\phi_{1}} \E{\phi_{2}}$.
+
+There are two limitations of these existing results.
+First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
+Second, they assume a non-recombining locus and may break down for sites separated by a finite distance.
+
+\subsubsection*{Loosely-linked sites}
+
+Previous work on the multiple-mergers coalescent with recombination is limited relative to the literature on multiple-mergers models in non-recombining loci.
+Nonetheless, existing results suggest that the two-site joint site frequency spectrum will behave differently as a function of genetic distance in multiple-mergers versus Kingman models.
+
+Eldon and Wakeley \cite{} calculated the expected linkage disequilibrium in a coalescent model of a population with multiple-mergers induced by occasional ``jackpot'' reproductive events.
+The found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations.
+This effect is similar to the impact of population growth on LD.
+In contrast, Eldon and Wakeley found that LD
+
+LD in the MMC.
+
+Correlations in coalescence times.
+
+\section*{Summary statistics}
+
+\section*{The jSFS-PMI distinguishes multiple-mergers from Kingman coalescent}
+
+\section*{The signal is robust to population size change}
+
+\section*{Low-recombination rate regions show signals of multiple mergers in \textit{D. melanogaster}}
+
+\section*{Humans}
 

### 2017-11-07 manuscript/mmc_manuscript.tex
Run weaker step-change simulations and update simulation notebook.

@@
 
-\title{Distinguishing among coalescent models using two-site joint allele frequency spectra}
+\title{Distinguishing among coalescent models using two-site allele frequency spectra}
 \author{Daniel P. Rice}
@@
 present a new method for detecting multiple mergers based on
-the mutual information of the joint site frequency spectrum at pairs of linked sites. Unlike the average SFS,
+the mutual information of the joint allele frequency spectrum at pairs of linked sites. Unlike the average SFS,
 the mutual information depends mostly on the topologies of genealogies
@@
 First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
-Second, they assume a non-recombining locus and may break down for sites separated by a finite distance.
+Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.
 
-\subsubsection*{Loosely-linked sites}
+\subsubsection*{Weakly-linked sites}
 
@@
 
-Eldon and Wakeley \cite{} calculated the expected linkage disequilibrium in a coalescent model of a population with multiple-mergers induced by occasional ``jackpot'' reproductive events.
-The found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations.
+Eldon and Wakeley \cite{} calculated the expected linkage disequilibrium (LD) in a modified Wright-Fisher model with occasional ``jackpot'' reproductive events.
+In these events, one individual is selected to be the ancestor of a fixed finite fraction of the population.
+As a result, even in the $N \to \infty$ limit, more than two lineages may share the same ancestor in the same generation.
+Eldon and Wakely found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations in this model.
 This effect is similar to the impact of population growth on LD.
-In contrast, Eldon and Wakeley found that LD
+In contrast, the authors found that pairwise coalescent times at two unlinked sites may be correlated due to jackpot events: there is a finite probability that both lineages at both sites are descended from the ``lucky'' ancestor.
 
-LD in the MMC.
+Birkner et al. \cite{} extended the results of Eldon and Wakeley \cite{} by deriving the ancestral recombination graph for a diploid population with jackpot reproductive events.
+They show that in this model, the limiting marginal coalescent process is a simultaneous multiple-mergers coalescent with up to four simultaneous mergers permitted.
+(The four possible mergers are due to the four chromosomes of the two diploid parents of the jackpot event.)
+Birkner et al. find that the correlations in pairwise coalescent times found by Eldon and Wakeley holds for correlations between coalescent times in samples larger than two.
+This results suggests that the two-site frequency spectrum considered as a function of genetic distance will be informative for identifying multiple-mergers coalescent processes.
 
-Correlations in coalescence times.
+\subsubsection*{Natural selection and multiple mergers}
+
+It is well-known that natural selection distorts sample genealogies (reviewed in \cite{neher:2013}.)
+For example, Durrett and Schweinsberg \cite{durrett:schweinsberg:2005} and Coop and Ralph \cite{coop:ralph:2013} have studied the effects of recurrent selective sweeps on the genealogical process.
+They find that, under some conditions, the rapid increase in the frequency of a beneficial allele generates multiple-mergers--like coalescent processes at linked sites.
+Others \cite{} have shown that rapidly-adapting populations, in which multiple beneficial mutations segregate at the same time and interfere with one another, also generate genealogies with apparent multiple mergers and unbalanced branching patterns.
+(These models converge to a modified version of the Bolthausen-Sznitman $\Lambda$-coalesecent \cite{neher:hallatscheck:2013}.)
+
+The genealogical effects of selection are not restricted to adaptive evolution; purifying selection at many linked loci can also distort the genealogies at neutral sites.
+While most work on this ``background selection'' \cite{charlesworth:etal:1993} focuses on the reduction of pairwise diversity (e.g.
+\cite{mcvicker}), recent theoretical studies have examined how it distorts genealogies.
 

### 2017-12-02 manuscript/mmc_manuscript.tex

@@
 \usepackage{amsmath}
+\usepackage{url}
 
@@
 \newcommand{\Fig}[1]{Figure~\ref{#1}}
+
 \newcommand{\E}[1]{\left<{#1}\right>}
+\DeclareMathOperator{\pmi}{PMI}
+\DeclareMathOperator{\wpmi}{wPMI}
 
@@
 variation in offspring number can all generate genealogies with
-"multiple-merger" events in which more than two lineages coalesce
+``multiple-merger'' events in which more than two lineages coalesce
 instantaneously. Therefore, detecting multiple mergers is important
@@
 
-\section*{Background}
+\section{Background}
 A typical goal of population genetics is to infer the evolutionary and demographic history of a population from its contemporary genetic diversity.
@@
 For example, models of pervasive weak selection, skewed offspring number distributions, and periodic strong bottlenecks can all generate genealogies that differ from the Kingman both topologically and in terms of branch length.
+[Review of MMC literature here?
+Neutral models (Eldon, Birkner, etc.), sweeps (Durett Schweinsberg, Coop and Ralph), rapid adaptation (Neher, Desai), background selection (Seger, Good, Nicholaisen), Range expansion (??)]
 Collectively, these models are known as \emph{multiple mergers coalescents} because, unlike the Kingman coalescent, they permit multiple lineages to coalesce instantaneously.
@@
 Clearly, it would be useful to be able to distinguish among different coalescent models, both to check that popular inferences of historical population sizes are valid and to understand which evolutionary forces are dominant in various populations.
+
+%\subsection{Natural selection and multiple mergers}
+%
+%It is well-known that natural selection distorts sample genealogies and thus patterns of genetic diversity at neutral sites (reviewed in \cite{neher:2013}.)
+%Selective sweeps of beneficial mutations reduces the genetic diversity at linked sites because the subpopulation marked by the beneficial allele expands rapidly compared to the neutral coalescent timescale \cite{maynard-smith:haig:1974, kaplan:etal:1989}.
+%Durrett and Schweinsberg \cite{durrett:schweinsberg:2005} showed that recurring selective sweeps at linked sites 
+%
+%
+%For example, Durrett and Schweinsberg \cite{durrett:schweinsberg:2005} and Coop and Ralph \cite{coop:ralph:2013} have studied the effects of recurrent selective sweeps on the genealogical process.
+%They find that, under some conditions, the rapid increase in the frequency of a beneficial allele generates multiple-mergers--like coalescent processes at linked sites.
+%Others \cite{} have shown that rapidly-adapting populations, in which multiple beneficial mutations segregate at the same time and interfere with one another, also generate genealogies with apparent multiple mergers and unbalanced branching patterns.
+%(These models converge to a modified version of the Bolthausen-Sznitman $\Lambda$-coalesecent \cite{neher:hallatscheck:2013}.)
+%
+%The genealogical effects of selection are not restricted to adaptive evolution; purifying selection at many linked loci can also distort the genealogies at neutral sites.
+%While most work on this ``background selection'' \cite{charlesworth:etal:1993} focuses on the reduction of pairwise diversity (e.g.
+\cite{mcvicker}), recent theoretical studies have examined how it distorts genealogies.
+%
+
+[Mention the distinction between single-locus models (e.g.
+Seger on Whale lice, Neher et al.
+with viruses) where you can explicitly infer trees, and multi-locus methods, where you average over many latent genealogies.
+This method is in the latter group.]
 
@@
 
-\section*{Theoretical motivation}
+\section{Theoretical motivation}
 
@@
 
-\subsubsection*{Tightly-linked sites}
+\subsubsection{Tightly-linked sites}
 
@@
 In contrast to the Kingman case, Birkner \textit{et al.} found that, in the $\Lambda$-coalescent, $\E{\tau_i \tau_j} - \E{\tau_i}\E{\tau_j}$ may be positive for $i \neq j$.
-These positive associations are largest for $1 < i - j \ll n$.
+These positive associations are largest for $|i - j| \ll n$.
 Together these results suggest that the joint site frequency spectrum of tightly-linked sites can distinguish qualitatively between the constant-sized Kingman and multiple-mergers coalescent.
 
-We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{1}).
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/trees.png}
+\caption{Possible genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue.  \label{fig:trees}}
+\end{figure}
+
+We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
 Furthermore, the total branch length is independent of the topology.
@@
 
-\subsubsection*{Weakly-linked sites}
+\subsubsection{Weakly-linked sites}
 
@@
 As a result, even in the $N \to \infty$ limit, more than two lineages may share the same ancestor in the same generation.
-Eldon and Wakely found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations in this model.
+Eldon and Wakeley found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations in this model.
 This effect is similar to the impact of population growth on LD.
@@
 (The four possible mergers are due to the four chromosomes of the two diploid parents of the jackpot event.)
-Birkner et al. find that the correlations in pairwise coalescent times found by Eldon and Wakeley holds for correlations between coalescent times in samples larger than two.
+Birkner et al. found that the correlations in pairwise coalescent times found by Eldon and Wakeley holds for correlations between coalescent times in samples larger than two.
 This results suggests that the two-site frequency spectrum considered as a function of genetic distance will be informative for identifying multiple-mergers coalescent processes.
 
-\subsubsection*{Natural selection and multiple mergers}
+\section{Methods}
 
-It is well-known that natural selection distorts sample genealogies (reviewed in \cite{neher:2013}.)
-For example, Durrett and Schweinsberg \cite{durrett:schweinsberg:2005} and Coop and Ralph \cite{coop:ralph:2013} have studied the effects of recurrent selective sweeps on the genealogical process.
-They find that, under some conditions, the rapid increase in the frequency of a beneficial allele generates multiple-mergers--like coalescent processes at linked sites.
-Others \cite{} have shown that rapidly-adapting populations, in which multiple beneficial mutations segregate at the same time and interfere with one another, also generate genealogies with apparent multiple mergers and unbalanced branching patterns.
-(These models converge to a modified version of the Bolthausen-Sznitman $\Lambda$-coalesecent \cite{neher:hallatscheck:2013}.)
+\subsection{Summary statistics}
 
-The genealogical effects of selection are not restricted to adaptive evolution; purifying selection at many linked loci can also distort the genealogies at neutral sites.
-While most work on this ``background selection'' \cite{charlesworth:etal:1993} focuses on the reduction of pairwise diversity (e.g.
-\cite{mcvicker}), recent theoretical studies have examined how it distorts genealogies.
+We would like to detect signals of non-Kingman coalescent processes in polymorphism data from a sample of size $n$ where the ancestral state is unknown.
+Thus, we will calculate statistics of the expected \textit{folded} site frequency spectrum, $\E{\eta_i }= \E{\phi_i + \phi_{n-i}}$, which is symmetric under the exchange of the ancestral and derived alleles.
+This quantity is the fraction of all sites (not only polymorphic sites) with minor allele frequency $i \in \{0, \ldots, \left \lfloor{n/2}\right \rfloor  \}$, and can be considered as a probability distribution.
+We can therefore think of the folded two-site frequency spectrum as a joint probability mass function, $\E{\eta_{ij}}$ over  $\{0, \ldots, \left \lfloor{n/2}\right \rfloor  \}^2$, the set of minor allele frequencies at the left and right positions.
 
-\section*{Summary statistics}
+We are interested in the signatures of genealogical distortions that affect the two-site frequency spectrum beyond their effects on the single-site frequency spectrum.
+A natural transformation of the data is the pointwise mutual information:
+\begin{equation}
+\pmi(i,j) = \log_2 \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right).
+\label{eq:pmi}
+\end{equation}
+Pointwise mutual information is a commonly-used measure of the strength of relationship between two random variables and is equal to the log ratio of the joint probability to the product of the marginal probabilities.
+(The more familiar mutual information is equal to the expectation of the pointwise mutual information.)
+An advantage of the pointwise mutual information in our setting is that for $i,j \geq 1$ both the numerator and denominator in \eq{eq:pmi} scale like $(T_2 \mu)^2$, so that PMI is insensitive to the mutation rate and pairwise coalescent time, and only measures the shapes of genealogies.
 
-\section*{The jSFS-PMI distinguishes multiple-mergers from Kingman coalescent}
+In order to highlight the portions of the joint site frequency spectrum that are most salient and easiest to estimate from finite data, we will also calculate the weighted pointwise mutual information:
+\begin{equation}
+\wpmi(i,j) = \frac{\E{\eta_{ij}}}{\bar{\pi}^2} \log_2 \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right),
+\label{eq:pmi}
+\end{equation}
+for $i, j \geq 1$, where $\bar{\pi} = 2T_2 \mu$ is the average pairwise diversity and preserves the invariance to changes in mutation rate.
 
-\section*{The signal is robust to population size change}
+\subsection{Simulations}
 
-\section*{Low-recombination rate regions show signals of multiple mergers in \textit{D. melanogaster}}
+For time-homogeneous $\Lambda$-coalescents without recombination, the expected joint site frequency spectrum may be calculated numerically using the recursion in \cite{birkner:etal:2013}.
+However, this algorithm is $\mathcal{O}(n^4)$ in the size of the sample and is thus impractical to calculate for large samples.
+Furthermore, there are no analytical results for recombining loci, time-varying coalescent rates, or models with natural selectuion.
+We therefore turn to simulations.
 
-\section*{Humans}
+We ran coalescent simulations using a version of the program \texttt{msprime} \cite{msprime} modified to allow for multiple-mergers according to the Beta coalescent model.
+[NOTE: it probably would not be too complicated to also do simultaneous multiple-mergers as suggested by Matthias Steinrucken.]
+Because we are interested in summary statistics that are independent of the rate of neutral mutations, we run the coalescent simulations without mutations and calculate expected site frequency spectra directly from branch lengths.
+This approach allows us to extract the maximum amount of information from each simulation run.
+
+We are primarily interested in distinguishing multiple-mergers coalescents from Kingman coalescent models of growing populations because these models generate similar single-site frequency spectra.
+First, we consider a growing population where the population size (inverse coalescence rate) is given by
+\begin{equation} N(t) = N_0 \exp (- gt),\end{equation}
+ where $t$ is the time before the present.
+In this model, the control parameter is $gN_0$, the population growth rate in coalescent time units.
+[NOTE: I could also use a model where the population was constant up until a time $T_g$ in the past and then began to grow exponentially.
+I suspect this won't change anything except adding another parameter, but I probably should test it out in at least a few cases.]
+
+Recent evidence suggests that some populations, including humans, may be growing at a faster-than-exponential rate \cite{} and that such growth may distort genealogies \cite{}.
+As a model for faster-than-exponential growth, we simulate piecewise-constant population sizes with two epochs:
+\begin{equation}
+N(t) = 
+\begin{cases}
+      N_0, &  t < T_g \\
+      N_0 / \gamma, & t \geq T_g.
+\end{cases}
+\end{equation}
+Patterns of diversity in this model are determined by $T_g / N$, the time of the growth, in coalescent units and $\gamma > 1$, the instantaneous growth factor.
+
+For simulations of recombining loci, we varied $r$, the per-generation recombination rate between the focal positions.
+These simulations have an additional control parameter $rT_2$, where $T_2$ is the average pairwise coalescent time measured from the simulations.
+
+[TO-DO: I've implemented bottleneck, 2-deme, and many-deme stepping-stone population simulations, but need to re-run them with the current approach.]
+
+[TO-DO: Simulations of linked-selection using SLiM \cite{slim:paper}.
+It's installed on the cluster and locally.
+I've learned how use it and run a few tests.
+I need to pick parameter values and run simulations at scale.]
+
+\subsection{Data analysis}
+
+\subsubsection{Zambian \textit{Drosophila melanogaster}}
+
+We calculated the 2-site frequency spectrum of single nucleotide polymorphisms in a sample of whole genome sequences ($\sim40x$ coverage) of $\sim200$ \textit{Drosophila melanogaster} from Zambia \cite{}.
+This population is believed to be from the ancestral range of the species, is very diverse ($\pi \approx 1\%$), and has limited global admixture \cite{}.
+We downloaded the individual consensus sequences available from the Drosophila Genome Nexus at \url{http://www.johnpool.net/genomes.html}, removed chromosome arms with known inversions, and downsampled to one hundred samples for each chromosome arm.
+In order to focus on high-quality sites and to simplify the interpretation of the allele frequency spectrum, we considered only sites with confident genotypes called in all one hundred samples (XX\% of sites.)
+[TO-DO: Look at only 4-fold degenerate sites. I've already identified them]
+
+To estimate the recombination rate between sites, we used the recombination map published by Comeron \textit{et al.} \cite{}, and available at \url{https://petrov.stanford.edu/cgi-bin/recombination-rates_updateR5.pl}.
+This map has a resolution of 100KB and was calculated by identifying individual cross-over events in XX trios.
+While the scale of this experiment was immense, the average number of observed cross-over events per window was only XX and so much of the window-to-window variation appears to be attributable to counting noise.
+To reduce the effect of this noise, we applied a standard wavelet de-noising procedure \cite{} to the recombination rate as a function of genomic position, yielding a smoother map [refer to an appendix].
+This recombination map yields a stronger relationship between recombination rate and diversity than the original map, a well-known feature of the \textit{D.
+melanogaster} genome \cite{begun:aquadro, etc}, suggesting that the wavelet de-noising was successful.
+
+\subsubsection{European humans}
+[TO-DO: EVE consortium European humans. deCODE genetic map]
+
+\section{Results}
+
+\subsection{Simulations}
+\Fig{fig:msfs} shows the average single-site frequency spectra for calculated from coalescent simulations as described above.
+Time-homogeneous Beta coalescent simulations range from the Kingman coalescent ($\alpha=2$) through increasing intensities of multiple-mergers as $\alpha \to 1$.
+(The $\alpha  \to 1$ limit is the Bolthausen-Sznitman coalescent.)
+The site frequency spectrum shows an excess of rare alleles relative to the Kingman expectation, and this excess is stronger for smaller values of $\alpha$.
+
+As expected, Kingman coalescent simulations of growing populations also show an excess of rare alleles.
+This distortion is more extreme for larger growth rates and the precise shape of the site frequency spectrum depends on the growth rate or timing of instantaneous growth.
+Note that our goal here is not to find population growth models that precisely reproduce the site frequency spectra of Beta coalescents, but to show that the qualitative behavior of the two classes of models is similar.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/mSFS.pdf}
+\caption{The single-site frequency spectrum compared to time-homogeneous Kingman expectation (dashed line) for multiple-mergers (Beta) coalescent, exponentially-growing populations, and instantaneously-growing populations (bottom row) as described in Methods.
+Spectra are normalized by the average pairwise diversity $\pi$.
+\label{fig:msfs}}
+\end{figure}
+
+\subsubsection{Two-site frequency spectra of non-recombining loci}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{figures/pmi_beta_r0.pdf}
+\caption{PMI (left) and wPMI (right) for time-homogeneous Beta coalescent simulations without recombination.
+Positive values are in orange, negative values are in purple.
+[TODO: colorbar]\label{fig:pmi_beta_r0}}
+\end{figure}
+
+\Fig{fig:pmi_beta_r0} shows the pointwise mutual information and weighted pointwise mutual information for time-homogeneous Beta coalescent simulations without recombination.
+As expected, the Kingman coalescent has only negative values away from the diagonal, while increasing the intensity of multiple mergers leads to positive off-diagonal values.
+In particular, multiple mergers induce positive correlations between the numbers of alleles at similar but not identical frequencies, as shown by the positive values near the diagonals.
+Finally, the weighted pointwise information reveals that multiple mergers lead to positive correlations between the numbers of low-frequency and high-frequency mutations at linked sites.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/pmi_growth_r0.pdf}
+\caption{PMI in Kingman coalescent simulations of growing populations without recombination.
+Left column shows exponential growth.
+Other columns show instantaneous growth.
+Positive values are in orange, negative values are in purple.
+[TODO: colorbar]\label{fig:pmi_growth_r0}}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/wpmi_growth_r0.pdf}
+\caption{wPMI in Kingman coalescent simulations of growing populations without recombination.
+Left column shows exponential growth.
+Other columns show instantaneous growth.
+Positive values are in orange, negative values are in purple.
+[TODO: colorbar]\label{fig:wpmi_growth_r0}}
+\end{figure}
+
+\Fig{fig:pmi_growth_r0} and \Fig{fig:wpmi_growth_r0} show the pointwise mutual information and weighted pointwise mutual information respectively for models of population growth without recombination.
+Together they show that, unlike the single-site frequency spectrum, these transformations of the two-site frequency spectrum of linked sites are relatively insensitive to population growth.
+All of the parameters considered generate PMI values that are negative for alleles at different frequencies.
+This suggests that the two-site frequency spectrum contains information that can qualitatively distinguish between Kingman and multiple-mergers coalescent processes.
+
+\subsubsection{Two-site frequency spectra of recombining loci}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/wpmi_beta_r.pdf}
+\caption{wPMI in Beta coalescent simulations of loci with recombination. Positive values are in orange, negative values are in purple. [TODO: colorbar]\label{fig:wpmi_beta_r}}
+\end{figure}
+
+\Fig{fig:wpmi_beta_r} shows the weighted pointwise mutual information for time-homogeneous Beta coalescent simulations with recombination.
+The beta coalescent simulations have overall more positive PMI values than the Kingman coalescent at all recombination rates
+On the other hand, at intermediate recombination rates $rT_2 \sim 1$, the Kingman simulations have positive associations between alleles at similar frequencies, muddying this feature as a signature of multiple mergers.
+(These positive correlations are due to the partial breakdown of size-$i$ clades into size-$i-\delta_i$ clades at moderately-linked sites.)
+
+However, there are two signals that appear to be robust to recombination.
+First, the positive association between low-frequency and high-frequency alleles is not present in the Kingman simulations at any recombination rate.
+Second, this positive association persists even at large genetic distances in the Beta coalescent simulations.
+In contrast, in the Kingman simulations, all mutual information between the two sites decays to zero at genetic distances greater than $rT_2 \sim 1$.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/wpmi_exp_r.pdf}
+\caption{wPMI in Kingman coalescent simulations of exponential growth with recombination. Positive values are in orange, negative values are in purple. [TODO: colorbar]\label{fig:wpmi_exp_r}}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{figures/wpmi_step_r.pdf}
+\caption{wPMI in Kingman coalescent simulations of instantaneous growth with recombination. Positive values are in orange, negative values are in purple. [TODO: colorbar]\label{fig:wpmi_step_r}}
+\end{figure}
+
+\Fig{fig:wpmi_exp_r} and \Fig{fig:wpmi_step_r} show the weighted pointwise mutual information for models of population growth without recombination.
+While the different results vary in detail, they all share the same key features with the time-homogeneous Kingman model:
+(1) recombination generates positive associations between alleles at similar frequencies, but not between high- and low-frequency alleles, and
+(2) mutual information decays to zero for $rT_2 \gg 1$.
+
+These simulation results suggest a simple new data summary that can qualitatively distinguish between genealogical distortions due to multiple mergers and those due to population growth.
+We reduce the site frequency spectrum to two elements by defining the expected non-singleton frequency:
+\begin{equation}
+\E{\eta_{>1}} = \sum_{i=2}^{\left \lfloor n/2 \right \rfloor} \E{\eta_i},
+\end{equation}
+and similarly define the reduced two-site frequency spectrum.
+The association between high- and low- frequency alleles is captured by the pointwise mutual information between singletons and non-singletons:
+\begin{equation}
+\log_2 \left( \frac{\E{\eta_{1,>1}}}{\E{\eta_1} \E{\eta_{>1}}} \right).
+\label{eq:singleton_pmi}
+\end{equation}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/singleton_pmi_all.pdf}
+\caption{Singleton/non-singleton PMI in all coalescent simulations recombination.
+Colored curves show time-homogeneous Beta coalescent (including Kingman in blue.) Grey curves show all population growth simulations.
+\label{fig:singleton_pmi}}
+\end{figure}
+
+\Fig{fig:singleton_pmi} shows \eq{eq:singleton_pmi} for all simulations.
+The left panel shows that, for each coalescent process, the singleton/non-singleton PMI is not very different between perfectly loci, $rT_c=0$, and moderately-linked loci $rT_c \sim 1$.
+The right panel shows that singleton/non-singleton PMI decays to zero at long genetic distances for Kingman models, but not the Beta coalescent. 
+At no point do any of the Kingman coalescent models generate positive singleton/non-singleton PMI.
+
+\subsection{Data analysis}
+
+\subsubsection{Zambian \textit{Drosophila melanogaster}}
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/chr2R_summary.pdf}
+\caption{\textit{Drosophila} chromosome arm 2R summary.
+(A) Recombination rate in 100kb windows.
+We grouped the data into four evenly-spaced bins by recombination rate (colors).
+(B) Average pairwise diversity is higher in bins with higher recombination rate.
+(C) All recombination-rate bins show an excess of low-frequency alleles relative to the Kingman expectation.
+The lowest-recombination bin shows the most distortion.
+(D) Singleton/non-singleton PMI is positive and persists at long genetic distances.
+\label{fig:drosophila}}
+\end{figure}
+
+We partitioned chromosome arm 2R into four bins by our de-noised estimates of the recombination rate in 100KB windows (\fig{fig:drosophila}A, see methods).
+The local recombination rate is a good predictor of the local average pairwise diversity, as has been shown before for \emph{D. melanogaster} (\fig{fig:drosophila}B).
+This relationship is putatively due to the action of background selection or selective sweeps \cite{}.
+If this hypothesis is correct, we expect to see more signatures of non-Kingman coalescent processes in the regions of low recombination.
+In fact, we see that the single-site frequency spectrum deviates from the time-homogeneous Kingman expectation in all windows, but especially in the lowest-recombination window (\fig{fig:drosophila}C).
+We also find positive singleton/non-singleton PMI in all windows (\fig{fig:drosophila}D).
+Furthermore, this positive association does not decay to zero for $rT_2\gg1$, as expected from the multiple-mergers simulations.
+This signal appears to be stronger in lower-recombination rate windows than in high-recombination rate windows.
+
+On the other hand, the decay of singleton/non-singleton PMI at short genetic distances ($rT_2<1$) is not predicted from our beta coalescent simulations.
+[Is this because selection generates different ARGs from $\Lambda$-coalescents? Is it because of some genomic property or data artifact we're not modeling?]
+
+[TO-DO: Alternative explanations. e.g. variation in mutation rate]
+
+[TO-DO: Other chromosome arms.]
+
+[TO-DO: Replace binning by $r$ with something continuous.]
+
+\subsubsection{European \textit{Homo sapiens}}
+[TODO]
 

### 2017-12-05 src/helpers.py
Add a python module containing commonly-used helper functions.

@@ -1,0 +1,48 @@
+import numpy as np
+
+def pairwise_diversity(allele_counts, sample_size):
+    f = allele_counts / sample_size
+    return 2*f*(1-f) / (1 - (1/sample_size))
+
+def folded_sfs(minor_allele_counts, sample_size):
+    bins=np.arange(1, (sample_size+1)//2 + 2)
+    return np.histogram(minor_allele_counts, bins=bins)[0]
+
+def sfs2pi(sfs, sample_size):
+    counts = np.arange(1, (sample_size+1)//2 + 1)
+    pi_weights = pairwise_diversity(counts, sample_size)
+    return pi_weights.dot(sfs)
+
+def import_slim_output(filename):
+    positions = []
+    allele_counts = []
+    with open(filename) as infile:
+        line = ''
+        # Scan to the begining of the sample output
+        while not line.startswith('#OUT'):
+            line = infile.readline()
+        sample_size = int(line.split()[-1])
+        while not line.startswith('Mutations:'):
+            line = infile.readline()
+        # Get the segregating mutations and allele counts
+        while infile:
+            line = infile.readline()
+            if line.startswith('Genomes:'):
+                break
+            sline = line.split()
+            ac = int(sline[8])
+            if ac < sample_size:
+                positions.append(int(sline[3]))
+                allele_counts.append(ac)
+    return sample_size, np.array(positions), np.array(allele_counts)
+
+def cross_correlation(X, Y, lim):
+    return np.correlate(X, Y[:-lim], 'valid') / (X.shape[0] - lim)
+
+def windowed_sfs(positions, minor_allele_counts, sample_size, L, window_size):
+    sfs_w = np.zeros(((sample_size+1)//2, L//window_size), dtype=int)
+    pos_w = positions // window_size
+    bins = np.arange(L//window_size + 1)
+    for i in range(sfs_w.shape[0]):
+        sfs_w[i,:] = np.histogram(pos_w[minor_allele_counts == i+1], bins=bins)[0]
+    return sfs_w

### 2017-12-05 src/parse_slim.py
Script for parsing slim output to windowed site frequency spectra.

@@ -1,0 +1,30 @@
+import sys
+import numpy as np
+import helpers as h
+
+filename = sys.argv[1]
+L_str = sys.argv[2]
+w_str = sys.argv[3]
+
+if 'e' in L_str:
+    L = int(float(L_str))
+else:
+    L = int(L_str)
+
+if 'e' in w_str:
+    window_size = int(float(w_str))
+else:
+    window_size = int(w_str)
+
+fn_split = filename.split('.')
+outfn = '.'.join(fn_split[:-1]) + '.wsfs.' + fn_split[-1] + '.gz'
+
+sys.stderr.write('Importing data...\n')
+sample_size, positions, allele_counts = h.import_slim_output(filename)
+minor_allele_counts = np.minimum(allele_counts, sample_size - allele_counts)
+
+sys.stderr.write('Computing windowed statistics...\n')
+sfs_w = h.windowed_sfs(positions, minor_allele_counts, sample_size, L, window_size)
+
+sys.stderr.write('Saving to file...\n')
+np.savetxt(outfn, sfs_w, fmt='%d', header='window_size={}'.format(window_size))

### 2017-12-05 src/parse_slim.py
Add slim output parsing to pipeline

@@ -19,12 +19,12 @@ else:
 fn_split = filename.split('.')
 outfn = '.'.join(fn_split[:-1]) + '.wsfs.' + fn_split[-1] + '.gz'
 
-sys.stderr.write('Importing data...\n')
+# sys.stderr.write('Importing data...\n')
 sample_size, positions, allele_counts = h.import_slim_output(filename)
 minor_allele_counts = np.minimum(allele_counts, sample_size - allele_counts)
 
-sys.stderr.write('Computing windowed statistics...\n')
+# sys.stderr.write('Computing windowed statistics...\n')
 sfs_w = h.windowed_sfs(positions, minor_allele_counts, sample_size, L, window_size)
 
-sys.stderr.write('Saving to file...\n')
+# sys.stderr.write('Saving to file...\n')
 np.savetxt(outfn, sfs_w, fmt='%d', header='window_size={}'.format(window_size))

### 2018-07-31 src/helpers.py
Minor changes to helper files.

@@ -8,12 +8,12 @@ def pairwise_diversity(allele_counts, sample_size):
 
 
 def folded_sfs(minor_allele_counts, sample_size):
-    bins=np.arange(1, (sample_size+1)//2 + 2)
+    bins=np.arange(1, sample_size//2 + 2)
     return np.histogram(minor_allele_counts, bins=bins)[0]
 
 
 def sfs2pi(sfs, sample_size):
-    counts = np.arange(1, (sample_size+1)//2 + 1)
+    counts = np.arange(1, sample_size//2 + 1)
     pi_weights = pairwise_diversity(counts, sample_size)
     return pi_weights.dot(sfs)
 

### 2018-07-31 src/zivkovic.py

@@ -260,7 +260,7 @@ def sigma_ij(n, mode, **params):
     return Sigma_ij
 
 def main():
-    n = 4
+    n = 5
 
     print("Constant")
     mode = "constant"

### 2018-08-03 manuscript/manuscript.tex
Begin draft

--- 2017-11-03 manuscript/mmc_manuscript.tex
@@
 \documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-%\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
 \usepackage{amssymb}
 \usepackage{amsmath}
+\usepackage{url}
+\usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
+% Fix biblatex behavior that writes "In: " before journal name.
+\renewbibmacro{in:}{%
+  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
 
-%SetFonts
+\bibliography{references}
 
-\newcommand{\eq}[1]{Eq.~(\ref{#1})}
-\newcommand{\Eq}[1]{Equation~(\ref{#1})}
-\newcommand{\fig}[1]{Fig.~\ref{#1}}
-\newcommand{\Fig}[1]{Figure~\ref{#1}}
-\newcommand{\E}[1]{\left<{#1}\right>}
-
-\title{Distinguishing among coalescent models using two-site joint allele frequency spectra}
+\title{Distinguishing among coalescent models using two-site allele frequency spectra}
 \author{Daniel P. Rice}
@@
 
-\abstract{The genetic diversity of a population reflects its demographic and
+\abstract{[NOTE: OLD]
+The genetic diversity of a population reflects its demographic and
 evolutionary history. Methods for inferring this history typically
@@
 
-\section*{Background}
-A typical goal of population genetics is to infer the evolutionary and demographic history of a population from its contemporary genetic diversity.
+\section*{Introduction}
+The genetic diversity of a population reflects its demographic and evolutionary history.
+Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
+The fundamental tools of the trade are toy models, which show how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
+However, populations are complicated and, moreover, varied in their complications.
+No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
+It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.
+
+One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
+The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
+Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
+The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{ms})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
+Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
+
+In its simplest form, the Kingman coalescent has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
+Under a variety of conditions, the coalescent rate is inversely proportional to the size of the population~\autocite{Kingman1982b}.
+Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
+This observation is the basis for a variety of methods to infer the population size history from genetic data.
+
+With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
+One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
+The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
+Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
+Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
+Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
+(Another class of SFS-based methods are based on forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)
+
+[Question: where do I cite the SMC models?]
+
+Model misspecification. Schrider, others on problems of inference. Evidence for widespread selection.
+
+Class of non-Kingman coalescents: multiple mergers. Where they arise. What they mean for inference (e.g. distort SFS, long-range linkage, timescale issues)
+
+Tests for multiple mergers and non-neutrality in general.
+General neutrality tests (Tradition of Taj's D. Achaz)
+Problem with confounding (compare to neutral background)
+Local tests for sweeps (haplotype tests)
+Specific tests for multiple mergers
+
+Our test, based on the 2-site frequency spectrum:
+\begin{itemize}
+\item Genome-wide. Not comparing loci to putatively neutral background.
+\item Distinguishes time-varying N from multiple mergers.
+\item SNP data.
+\item Doesn't require phasing or ancestral states.
+\item Easy to compute.
+\item Works in principle with small samples.
+\end{itemize}
+
+Outline:
+\begin{itemize}
+\item Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)
+\item Coalescent simulations with recombination (it still works).
+\item Forward-time simulations of linked sweeps (still works).
+\item Application to D. mel (it detects non-Kingman)
+\end{itemize}
+
+Modern population genetics makes it possible to learn the evolutionary and demographic history of a population from its contemporary genetic diversity~\autocite{Hahn2018}.
 The genetic differences among individuals reflect their genealogical history, which, in turn, reflects the history of various population parameters such as the population size, spatial structure, and the influence of natural selection.
@@
 
-The most commonly used genealogical model is the Kingman coalescent, which arises in many models of neutral evolution.
-The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in (continuous) time.
-As a result, the genealogies generated by the Kingman coalescent are randomly bifurcating trees: each node subtends exactly two branches and the number of leaves subtended by one of these branches is uniformly distributed.
+The most commonly used genealogical model is the Kingman coalescent, which arises in many models of neutral evolution~\autocite{Kingman1982, Hudson1983, Tajima1983}.
+The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in time.
+Moreover, the genealogies generated by the Kingman coalescent are randomly bifurcating trees: each node subtends exactly two branches and the number of leaves subtended by one of these branches is uniformly distributed~\autocite{Tajima1983}.
 
-Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population.
-This timescale, which we will call $T_2$, is inversely proportional to the rate of coalescence between pairs of lineages and sets the branch lengths of genealogies.
-Thus, $T_2$ determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
-Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
-In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population.
+% Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population.
+% This timescale, which we will call $T_2$, is inversely proportional to the rate of coalescence between pairs of lineages and sets the branch lengths of genealogies.
+% Thus, $T_2$ determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
+% Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
+% In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population.
 Therefore, it is commonly referred to as the effective population size.
@@
 
-An obvious question is whether we can use the average site frequency spectrum to reject the Kingman model?
+[Mention the distinction between single-locus models (e.g.
+Seger on Whale lice, Neher et al.
+with viruses) where you can explicitly infer trees, and multi-locus methods, where you average over many latent genealogies.
+This method is in the latter group.]
+
+An obvious question is whether we can use the average site frequency spectrum to reject multiple-mergers.
 There are two distinguishing features of the SFS generated by multiple mergers coalescents.
@@
 
-\section*{Theoretical motivation}
+\section*{Single-locus theory}
 
-Consider a sample of $n$ chromosomes.
-Denote the site frequency spectrum, the fraction of sites with a mutation in $i$ of $n$ samples, by $\phi_i$.
-We define the joint site frequency spectrum, $\phi_{ij}(d)$ for pairs of sites separated by a genomic distance of $d$ bases to be the fraction of such pairs with a mutation in $i$ samples at the left site and a mutation in $j$ samples at the right site.
-As described above, the expected site frequency spectrum is related to the expected branch lengths of the genealogies by:
-\begin{equation}
-\E{\phi_i} = \mu \E{\tau_i},
-\end{equation}
-where $\mu$ is the per--base pair mutation rate, $\tau_i$, is the total branch length subtending $i$ leaves, and expectations are taken over the distribution of genealogies.
-Similarly, the joint site frequency spectrum is proportional to the second moments of the joint branch length distribution:
-\begin{equation}
-\E{\phi_{ij}} = \mu^2 \E{\tau_i^l \tau_j^r},
-\end{equation}
-where $\tau_i^l$ and $\tau_j^r$ are the branch lengths at the left and right site respectively.
-There are several previous results that suggest that the correlation structure of branch lengths contains information that can distinguish multiple mergers from Kingman coalescent processes.
-If this is the case, then the joint site frequency spectrum will be a useful summary statistic for coalscent model selection.
+\section*{Recombination}
 
-\subsubsection*{Tightly-linked sites}
+\section*{Application to \textit{Drosophila melanogaster}}
 
-For tightly-linked sites (i.e.
-$rdT_2\ll1$ for per--base pair recombination rate $r$), $\E{\tau_i^l \tau_j^r} \approx \E{\tau_i \tau_j}$, the branch length correlation matrix for a non-recombining locus.
-This correlation matrix depends on correlations among the timing of coalescent events as well as correlations between clade sizes in tree topologies.
-Fu \cite{} calculated the first and second moments of $\tau_i$ for the time-homogeneous Kingman coalescent.
-His results show that $\E{\tau_i \tau_j} - \E{\tau_i}\E{\tau_j} < 0$ for $i \neq j$ and $i \neq n-j$.
-This implies that the off-diagonal elements of the joint site frequency spectrum will be depleted relative to the product of the marginal site frequency spectra for tightly-linked sites.
-Recently, Birkner \textit{et al.} \cite{}, extended Fu's calculation to the class of multiple-mergers coalescent models known as time-homogeneous $\Lambda$-coalescents.
-In contrast to the Kingman case, Birkner \textit{et al.} found that, in the $\Lambda$-coalescent, $\E{\tau_i \tau_j} - \E{\tau_i}\E{\tau_j}$ may be positive for $i \neq j$.
-These positive associations are largest for $1 < i - j \ll n$.
-Together these results suggest that the joint site frequency spectrum of tightly-linked sites can distinguish qualitatively between the constant-sized Kingman and multiple-mergers coalescent.
-
-We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{1}).
-Furthermore, the total branch length is independent of the topology.
-As a result, there is a trade-off between the branch length leading to singleton/tripleton mutations and the branch length leading to doubletons: loci with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
-Conditional on observing a doubleton at a locus, it is thus more likely that the locus has topology (B) and so the expected number of singletons is lower.
-In terms of the joint site frequency spectrum, we have $\E{\phi_{12}} < \E{\phi_{1}} \E{\phi_{2}}$.
-
-On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
-For example, topology (C) has less overall opportunity for singletons as well as doubletons than (A) or (B), even though the expected proportion of singletons is higher.
-Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies more likely.
-If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) and lead to $\E{\phi_{12}} > \E{\phi_{1}} \E{\phi_{2}}$.
-
-There are two limitations of these existing results.
-First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
-Second, they assume a non-recombining locus and may break down for sites separated by a finite distance.
-
-\subsubsection*{Loosely-linked sites}
-
-Previous work on the multiple-mergers coalescent with recombination is limited relative to the literature on multiple-mergers models in non-recombining loci.
-Nonetheless, existing results suggest that the two-site joint site frequency spectrum will behave differently as a function of genetic distance in multiple-mergers versus Kingman models.
-
-Eldon and Wakeley \cite{} calculated the expected linkage disequilibrium in a coalescent model of a population with multiple-mergers induced by occasional ``jackpot'' reproductive events.
-The found that LD may be reduced by multiple mergers at short genetic distances because of the elevated proportion of singleton mutations.
-This effect is similar to the impact of population growth on LD.
-In contrast, Eldon and Wakeley found that LD
-
-LD in the MMC.
-
-Correlations in coalescence times.
-
-\section*{Summary statistics}
-
-\section*{The jSFS-PMI distinguishes multiple-mergers from Kingman coalescent}
-
-\section*{The signal is robust to population size change}
-
-\section*{Low-recombination rate regions show signals of multiple mergers in \textit{D. melanogaster}}
-
-\section*{Humans}
+\printbibliography
 
 \end{document}
+
+TODO:
+- Review and re-write abstract.
+-

### 2018-08-03 manuscript/manuscript.tex
Finish introduction.

@@
 Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
-The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{ms})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
+The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
 Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
@@
 Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
-This observation is the basis for a variety of methods to infer the population size history from genetic data.
+This observation is the basis for a variety of methods to infer historical population sizes from genetic data.
 
@@
 Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
-(Another class of SFS-based methods are based on forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)
+(Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)
 
-[Question: where do I cite the SMC models?]
+A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
+One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
+When this assumption is violated, Kingman-based inference methods are misspecified.
+For example, \cite{SchriderEtAl2016} recently demonstrated how population size inference can be distorted by selective sweeps.
+This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
+In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
+Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
 
-Model misspecification. Schrider, others on problems of inference. Evidence for widespread selection.
+An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of neutral and non-neutral contexts.
+Whereas in the Kingman coalescent lineages may only coalesce pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
+The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
+These models are relevant for species with
+``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
+fat-tailed offspring number distributions \autocite{Schweinsberg2003},
+recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
+rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
+and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
+In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
+For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
+In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
+It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
 
-Class of non-Kingman coalescents: multiple mergers. Where they arise. What they mean for inference (e.g. distort SFS, long-range linkage, timescale issues)
+One possible approach to identifying multiple mergers in genomic data to use the SFS as a summary statistic.
+(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but as with other ``exact'' inference methods it does not scale to genomic data.)
+\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
+\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merge coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
+\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.
 
-Tests for multiple mergers and non-neutrality in general.
-General neutrality tests (Tradition of Taj's D. Achaz)
-Problem with confounding (compare to neutral background)
-Local tests for sweeps (haplotype tests)
-Specific tests for multiple mergers
+However, existing methods have limitations for distinguishing multiple mergers from general models of population-size change.
+The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
+\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
+The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but detecting it in data requires knowledge of the ancestral allele at each site.
+High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS at high frequencies.
 
-Our test, based on the 2-site frequency spectrum:
+Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby site---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
+These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
+Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
+Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
+Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
+
+\section*{Definitions and single-locus theory}
+[Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)]
+
+\section*{Simulations with recombination}
+[Coalescent simulations with recombination (it still works).]
+
+\section*{Linked selective sweeps}
+[Forward-time simulations of linked sweeps (still works).]
+
+\section*{Application to \textit{Drosophila melanogaster}}
+
+\section*{Discussion}
 \begin{itemize}
-\item Genome-wide. Not comparing loci to putatively neutral background.
-\item Distinguishes time-varying N from multiple mergers.
-\item SNP data.
-\item Doesn't require phasing or ancestral states.
-\item Easy to compute.
-\item Works in principle with small samples.
+    \item Focus here on coalescent, but also relevant for diffusion methods
+    \item Single-locus tests that rely on trees (Seger whale lice, Neher with viruses)
+    \item Multi-locus tests that compare vs genomic background
+    \item Could extend to a local method to look for variation in ``multiple-merger-ness''
+    \item Could also extend
 \end{itemize}
 
-Outline:
-\begin{itemize}
-\item Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)
-\item Coalescent simulations with recombination (it still works).
-\item Forward-time simulations of linked sweeps (still works).
-\item Application to D. mel (it detects non-Kingman)
-\end{itemize}
-
-Modern population genetics makes it possible to learn the evolutionary and demographic history of a population from its contemporary genetic diversity~\autocite{Hahn2018}.
-The genetic differences among individuals reflect their genealogical history, which, in turn, reflects the history of various population parameters such as the population size, spatial structure, and the influence of natural selection.
-By modeling how the (unobserved) genealogical process depends on these parameters and how patterns of genetic diversity depend on genealogy, we can develop inference procedures to learn about the past.
-
-The most commonly used genealogical model is the Kingman coalescent, which arises in many models of neutral evolution~\autocite{Kingman1982, Hudson1983, Tajima1983}.
-The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in time.
-Moreover, the genealogies generated by the Kingman coalescent are randomly bifurcating trees: each node subtends exactly two branches and the number of leaves subtended by one of these branches is uniformly distributed~\autocite{Tajima1983}.
-
-% Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population.
-% This timescale, which we will call $T_2$, is inversely proportional to the rate of coalescence between pairs of lineages and sets the branch lengths of genealogies.
-% Thus, $T_2$ determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
-% Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
-% In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population.
-Therefore, it is commonly referred to as the effective population size.
-
-%The Kingman coalescent generates genealogies with several distinguishing characteristics.
-First, the genealogies are binary trees: each node has only two children.
-A second and related topological feature is that for a node subtending $n$ leaves, the number of leaves subtended by one of its children is uniformly distributed between 1 and $n-1$.
-Finally, under a neutral model of evolution, the rate of coalescence (and thus the branch lengths of the genealogy) is inversely proportional to the population size.
-
-Because computing the full likelihood of the data is generally intractable, population genetic inference is typically done on informative summary statistics.
-One informative statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in a sample.
-The expected number of mutations in $i$ sampled chromosomes is proportional to the lengths of branches subtending $i$ leaves of the genealogy, and thus depends the distributions of topologies as well as branch lengths.
-By assuming the Kingman model, one can marginalize over the unobserved tree topologies and extract information about the distribution of branch lengths and thus about the rate of coalescence as a function of time.
-If this assumption is good, the site frequency spectrum thus reflects the history of population size changes.
-For example, high-frequency alleles are the result of mutations on deep branches and their number in a sample reflects the population size in the distant past.
-
-A serious problem for this inference procedure is that alternative models of evolution generate different coalescent models.
-For example, models of pervasive weak selection, skewed offspring number distributions, and periodic strong bottlenecks can all generate genealogies that differ from the Kingman both topologically and in terms of branch length.
-[Review of MMC literature here?
-Neutral models (Eldon, Birkner, etc.), sweeps (Durett Schweinsberg, Coop and Ralph), rapid adaptation (Neher, Desai), background selection (Seger, Good, Nicholaisen), Range expansion (??)]
-Collectively, these models are known as \emph{multiple mergers coalescents} because, unlike the Kingman coalescent, they permit multiple lineages to coalesce instantaneously.
-As a result, multiple-mergers genealogies contain nodes with more than two children and lack the uniform branching structure of the Kingman.
-Even more important for inference is the fact that the coalescent timescales---and thus the branch lengths and levels of diversity---are determined by parameters other than the population size.
-In this case, the term `effective population size' is misleading because $T_2$ is no longer proportional to the number of individuals in the population.
-Thus, the results of a Kingman-based inference procedure applied to data generated by a non-Kingman process will be both quantitatively and qualitatively wrong.
-Clearly, it would be useful to be able to distinguish among different coalescent models, both to check that popular inferences of historical population sizes are valid and to understand which evolutionary forces are dominant in various populations.
-
-[Mention the distinction between single-locus models (e.g.
-Seger on Whale lice, Neher et al.
-with viruses) where you can explicitly infer trees, and multi-locus methods, where you average over many latent genealogies.
-This method is in the latter group.]
-
-An obvious question is whether we can use the average site frequency spectrum to reject multiple-mergers.
-There are two distinguishing features of the SFS generated by multiple mergers coalescents.
-First, low-frequency mutations are enriched relative to the time-homogeneous Kingman coalescent.
-Unfortunately, this is also a feature of the Kingman coalescent in growing populations, so it is not well-suited for model selection (but see \cite{}).
-Second, the multiple-merger SFS is non-monotonic: it has a positive slope for high frequencies.
-Unlike the excess of rare alleles, this feature cannot be reproduced by an exchangeable Kingman coalescent model.
-However, it is difficult to identify the non-monotonicity in real data.
-The problem is that it necessary to know which allele is ancestral to distinguish derived mutations at frequency $x$ from those at frequency $1-x$.
-Various methods for identifying the ancestral allele exist, but in practice even a very low rate of error can generate a spurious signal.
-
-Ideally, we would like a statistic (or statistics) that:
-\begin{enumerate}
-\item is qualitatively different between models of genealogies,
-\item is robust to demographic changes, and
-\item requires minimal information beyond genetic diversity.
-\end{enumerate}
-The first two criteria are important for making qualitative distinctions between models. The last one would maximize its applicability.
-Here, I propose that the joint site frequency spectrum at pairs of linked sites is such a statistic.
-
-\section*{Single-locus theory}
-
-\section*{Recombination}
-
-\section*{Application to \textit{Drosophila melanogaster}}
+Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
 

### 2018-08-03 manuscript/manuscript.tex
Begin definitions and non-recombining theory.

@@
   \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
+
+\newcommand{\floor}[1]{\lfloor #1 \rfloor}
+\newcommand{\E}[1]{\left< #1 \right>}
 
@@
 Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
-The fundamental tools of the trade are toy models, which show how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
+The fundamental tools of the trade are toy models, which are used to study how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
 However, populations are complicated and, moreover, varied in their complications.
@@
 One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
-The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
+The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
 Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
-The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
+The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
 Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
@@
 When this assumption is violated, Kingman-based inference methods are misspecified.
-For example, \cite{SchriderEtAl2016} recently demonstrated how population size inference can be distorted by selective sweeps.
+For example, \cite{SchriderEtAl2016} recently demonstrated how population-size inference can be distorted by selective sweeps.
 This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
@@
 
-An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of neutral and non-neutral contexts.
-Whereas in the Kingman coalescent lineages may only coalesce pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
+An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
+Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
 The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
@@
 
-One possible approach to identifying multiple mergers in genomic data to use the SFS as a summary statistic.
-(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but as with other ``exact'' inference methods it does not scale to genomic data.)
+One possible approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
+(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but, as with other ``exact'' inference methods, it does not scale to genomic data.)
 \cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
-\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merge coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
+\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
 \cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.
@@
 
-Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby site---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby sites---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
 We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
@@
 
-\section*{Definitions and single-locus theory}
-[Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)]
+\section*{Definitions and non-recombining theory}
+
+\subsection*{The SFS and 2-SFS}
+Consider a sample of $n$ chromosomes taking from a population of size $N$ with mutation rate $\mu$ per site per generation.
+Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : i \in \{1,\ldots,n-1\} \right\}$, where $\xi_i$ is the fraction of sites at which there is a mutation with derived allele count $i$ in the sample.
+In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
+Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}: i \in \{1, \ldots, \floor{n/2} \right\}$.
+%, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
+
+The SFS and folded SFS are single-site statistics.
+As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
+We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
+\begin{align*}
+    \left\{ \xi_{ij}(d) : i \in \{1, \ldots, n-1\}, j \in \{i, \ldots, n-1\} \right\},
+\end{align*}
+as the fraction of pairs of sites separated by $d$ bases with a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
+(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
+This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
+By analogy to the folded SFS, we define the folded 2-SFS, as
+\begin{align*}
+    \eta_{ij}(d) &= \xi_{ij}(d) \\
+                 &+ \delta_{n-2i} \xi_{(n-i),j}(d) \\
+                 &+ \delta_{n-2j} \xi_{i,(n-j)}(d) \\
+                 &+ \delta_{n-2i} \delta_{n-2j} \xi_{(n-i),(n-j)}(d).
+\end{align*}
+In this section, we will be considering non-recombining sites and so will suppress the distance $d$.
+
+In the limit of low per-site mutation rate, the expected SFS and 2-SFS are related to moments of the branch length distribution through,
+\begin{align}
+    \E{\xi_i} &= \mu \E{\tau_i} \\
+    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
+\end{align}
+where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
+Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.
+
+\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
+He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
+This result implies a negative association between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
+(There are positive associations between mutations at complimentary frequencies induced by trees whose root node partitions the tree into subtrees of size $i$ and $n-i$.)
+
+Later, \cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
+This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
+Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003} and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \cite{NeherHallatscheck2013}.
+Like Fu, Birkner et al.
+were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}, and thus mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
+However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
+Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.
+
+Our goal is to demonstrate that this distinction in the two-site frequency spectrum between multiple mergers and Kingman coalescent
+(1) applies to Kingman coalescents with time-varying coalescent rates,
+(2) is robust to recombination between the sites,
+(3) is also a feature of forward-time models with selection,
+and (4) can form the basis for a model-checking procedure for demographic inference methods.
+To this end, we will make use of the fact that the 2-SFS may be seen as a joint probability distribution to define a useful data transformation.
+
+\subsection*{Point-wise mutual information}
+\begin{itemize}
+    \item the sfs (with zeros) is a probability distribution
+    \item the 2-sfs (with zeros) is a joint probability distribution
+    \item Mutual information is a functional on joint pmfs that measures the association between two variables $\text{MI}[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$
+    \item The contributions to MI from each variable are ``pointwise mutual information'' $\text{PMI}(x,y) = \log \frac{p(x,y)}{p(x)p(y)}$.
+    \item Thus, we calculate PMI of the expected (folded) 2-SFS as $\text{PMI}(i,j) = \log \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right)$
+    \item The advantages of folding are (1) don't need ancestral state and (2) all elements are negative for standard Kingman.
+    \item also, note that it doesn't depend on either the mutation rate or the average pairwise coalescent time $T_2$.
+\end{itemize}
+
+\subsection*{Population growth vs. the beta coalescent}
+
+\begin{itemize}
+    \item Want to calculate PMI for 2-SFS for various models.
+    \item implemented Fu and Birkner formulas
+    \item \cite{ZivkovicWiehe2008} extended Fu to variable population size.
+    \item We implemented their formula for exponential growth and piecewise constant, but in principle could be extended to other models.
+    \item Figure 1 shows weighted PMI $\text{wPMI}(i,j) = \E{\eta_{ij}} \text{PMI}(i,j)$.
+    \item weighted to highlight regions of the 2-SFS with the most mutations.
+    \item Observations:
+    \begin{enumerate}
+        \item population growth does not qualitatively change the PMI, even though it distorts the SFS a lot.
+        \item more multiple mergers leads more positive elements.
+        \item Positive associations off-diagonal. Presumably due to uneven branching. Going to ignore because turns out to be sensitive to recombination.
+        \item Positive associations between high- and low-frequency mutations. This is the signal we'll focus on.
+    \end{enumerate}
+    \item Suggests a coarse-graining of the SFS into high- and low-frequency mutations.
+    \item Definition of hi-lo corr given cutoff and distance.
+\end{itemize}
+
+[Next: Intuition and $n=4$ calculations.]
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=0.8\textwidth]{figures/trees.png}
+% \caption{Possible genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue.  \label{fig:trees}}
+% \end{figure}
+%
+% We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
+% Furthermore, the total branch length is independent of the topology.
+% As a result, there is a trade-off between the branch length leading to singleton/tripleton mutations and the branch length leading to doubletons: loci with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
+% Conditional on observing a doubleton at a locus, it is thus more likely that the locus has topology (B) and so the expected number of singletons is lower.
+% In terms of the joint site frequency spectrum, we have $\E{\phi_{12}} < \E{\phi_{1}} \E{\phi_{2}}$.
+%
+% On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
+% For example, topology (C) has less overall opportunity for singletons as well as doubletons than (A) or (B), even though the expected proportion of singletons is higher.
+% Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies more likely.
+% If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) and lead to $\E{\phi_{12}} > \E{\phi_{1}} \E{\phi_{2}}$.
+%
+% There are two limitations of these existing results.
+% First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
+% Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.
 
@@
 
+\section*{Methods}
+\begin{itemize}
+    \item Code to compute moments
+    \item Simulation code extending msprime
+    \item Forward-time simulation with SLiM
+    \item Drosophila
+    \begin{itemize}
+        \item data pre-processing
+        \item fitting demographic model with fastNeutrino
+        \item simulating demographic model (could maybe compute exactly)
+    \end{itemize}
+\end{itemize}
+
 \printbibliography
@@
 \end{document}
-
-TODO:
-- Review and re-write abstract.
--

### 2018-08-05 manuscript/manuscript.tex
Finish definitions section.

@@
 \usepackage{url}
+% Allow inline lists
+\usepackage[inline]{enumitem}
 \usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
@@
 \newcommand{\E}[1]{\left< #1 \right>}
+\newcommand{\eq}[1]{Eq.~(\ref{#1})}
+\newcommand{\Eq}[1]{Equation~(\ref{#1})}
+\newcommand{\eqs}[2]{Eq.~(\ref{#1})~and~(\ref{#2})}
+\DeclareMathOperator{\pmi}{pmi}
+\DeclareMathOperator{\fpmi}{fpmi}
+\DeclareMathOperator{\hilopmi}{hilopmi}
 
@@
 
-\section*{Definitions and non-recombining theory}
+\section*{Definitions and previously known results}
 
 \subsection*{The SFS and 2-SFS}
-Consider a sample of $n$ chromosomes taking from a population of size $N$ with mutation rate $\mu$ per site per generation.
-Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : i \in \{1,\ldots,n-1\} \right\}$, where $\xi_i$ is the fraction of sites at which there is a mutation with derived allele count $i$ in the sample.
+Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
+Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
 In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
-Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}: i \in \{1, \ldots, \floor{n/2} \right\}$.
+Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta.
 %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
@@
 We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
-\begin{align*}
-    \left\{ \xi_{ij}(d) : i \in \{1, \ldots, n-1\}, j \in \{i, \ldots, n-1\} \right\},
-\end{align*}
-as the fraction of pairs of sites separated by $d$ bases with a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
+$\left\{ \xi_{ij}(d) : 1 \leq i, j < n\right\}$,
+as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
 (Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
@@
     \eta_{ij}(d) &= \xi_{ij}(d) \\
-                 &+ \delta_{n-2i} \xi_{(n-i),j}(d) \\
-                 &+ \delta_{n-2j} \xi_{i,(n-j)}(d) \\
-                 &+ \delta_{n-2i} \delta_{n-2j} \xi_{(n-i),(n-j)}(d).
+                 &+ (1-\delta_{i,n-i}) \xi_{n-i,j}(d) \\
+                 &+ (1-\delta_{j,n-j}) \xi_{i,n-j}(d) \\
+                 &+ (1-\delta_{i,n-i}) (1-\delta_{j,n-j}) \xi_{n-i,n-j}(d).
 \end{align*}
@@
 
-In the limit of low per-site mutation rate, the expected SFS and 2-SFS are related to moments of the branch length distribution through,
+In the limit of low per-site mutation rate ($\mu\to0$), all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
 \begin{align}
-    \E{\xi_i} &= \mu \E{\tau_i} \\
+    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
     \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
+    \label{eq:expected_2sfs}
 \end{align}
@@
 He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
-This result implies a negative association between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
-(There are positive associations between mutations at complimentary frequencies induced by trees whose root node partitions the tree into subtrees of size $i$ and $n-i$.)
+This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
+(There are positive correlations between mutations at complimentary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)
 
-Later, \cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
+\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
 This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
-Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003} and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \cite{NeherHallatscheck2013}.
-Like Fu, Birkner et al.
-were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}, and thus mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
+Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
+Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}
+As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
 However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
@@
 
-Our goal is to demonstrate that this distinction in the two-site frequency spectrum between multiple mergers and Kingman coalescent
-(1) applies to Kingman coalescents with time-varying coalescent rates,
-(2) is robust to recombination between the sites,
-(3) is also a feature of forward-time models with selection,
-and (4) can form the basis for a model-checking procedure for demographic inference methods.
-To this end, we will make use of the fact that the 2-SFS may be seen as a joint probability distribution to define a useful data transformation.
+Our goal is to demonstrate that this distinction in the two-site frequency spectrum between the multiple merger and Kingman coalescents:
+\begin{enumerate*}
+    \item applies to Kingman coalescents with time-varying coalescent rates,
+    \item is robust to recombination between the sites,
+    \item is also a feature of forward-time models with selection, and
+    \item can form the basis of a model-checking procedure for demographic inference methods.
+\end{enumerate*}
+To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following useful transformation.
 
-\subsection*{Point-wise mutual information}
-\begin{itemize}
-    \item the sfs (with zeros) is a probability distribution
-    \item the 2-sfs (with zeros) is a joint probability distribution
-    \item Mutual information is a functional on joint pmfs that measures the association between two variables $\text{MI}[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$
-    \item The contributions to MI from each variable are ``pointwise mutual information'' $\text{PMI}(x,y) = \log \frac{p(x,y)}{p(x)p(y)}$.
-    \item Thus, we calculate PMI of the expected (folded) 2-SFS as $\text{PMI}(i,j) = \log \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right)$
-    \item The advantages of folding are (1) don't need ancestral state and (2) all elements are negative for standard Kingman.
-    \item also, note that it doesn't depend on either the mutation rate or the average pairwise coalescent time $T_2$.
-\end{itemize}
+\subsection*{Frequency pointwise mutual information}
 
-\subsection*{Population growth vs. the beta coalescent}
+Given a coalescent model, the minor allele count at an arbitrary site is a random variable with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
+Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
+We are interested in quantifying the dependence between the minor allele counts at these sites, particularly for $i\neq j$.
+
+A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
+The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
+\begin{align}
+    I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
+    \label{eq:mutual_information}
+\end{align}
+where, with slight abuse of notation, $p(x) = \sum_y p(x,y)$ and vice versa.
+Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of independent random variables with the same marginal distributions \autocite{CoverThomas1991}.
+Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.
+
+We would like to measure the association between alleles at particular frequencies, rather than the overall dependence between allele frequencies at nearby sites (which would be a measure of linkage disequilibrium).
+\Eq{eq:mutual_information} shows that the $I$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
+\begin{align}
+    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)} = \log \frac{p(y|x)}{p(y)}
+\end{align}
+\autocite{ChurchHanks1990}.
+Pointwise mutual information measures the change in the probability that $X=x$ given knowledge that $Y=y$.
+When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
+On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
+
+As we have shown above, the folded 2-SFS is a joint probability distribution over minor allele counts.
+We may thus define the \textit{frequency pointwise mutual information} as
+\begin{align}
+    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
+\end{align}
+This transformation of the 2-SFS has several useful properties.
+First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
+Moreover, the results of \cite{Fu1995} show that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
+
+The second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the SFS.
+Population-size variation distorts the SFS by changing the distribution of waiting times.
+However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
+On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
+These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
+Our hope is that the 2-SFS contains additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
+
+\subsection*{Coarse-grained frequencies}
+
+With finite data, estimates of the 2-SFS will be noisy.
+This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
+We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
+This suggests a coarse-grained SFS and 2-SFS:
+\begin{align}
+    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
+    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
+    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
+\end{align}
+where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
+Coarse-graining allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
+
+We can also compute the pointwise mutual information in this coarse-grained distribution as
+\begin{align}
+    \hilopmi(d; i_c) = \log \frac
+                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
+                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
+\end{align}
+One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g the PMI between monomorphic and high minor allele count sites), but we will not use these statistics here.
+
+\section*{Results \label{sec:results}}
+
+\subsection*{Population growth vs. the beta coalescent in non-recombining loci}
 
@@
 
-\section*{Simulations with recombination}
+\subsection*{Pointwise mutual information between recombining sites}
 [Coalescent simulations with recombination (it still works).]
 
-\section*{Linked selective sweeps}
+\subsection*{Linked selective sweeps}
 [Forward-time simulations of linked sweeps (still works).]
 
-\section*{Application to \textit{Drosophila melanogaster}}
+\subsection*{Application to \textit{Drosophila melanogaster}}
 

### 2018-08-05 manuscript/manuscript.tex
Edit Definitions section.

@@
 Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
-Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}
+Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
 As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
@@
 
-Our goal is to demonstrate that this distinction in the two-site frequency spectrum between the multiple merger and Kingman coalescents:
-\begin{enumerate*}
-    \item applies to Kingman coalescents with time-varying coalescent rates,
-    \item is robust to recombination between the sites,
-    \item is also a feature of forward-time models with selection, and
-    \item can form the basis of a model-checking procedure for demographic inference methods.
-\end{enumerate*}
+Our goal is to demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
+\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
+    \item applies to Kingman coalescents with time-varying coalescent rates
+    \item is robust to recombination between the sites
+    \item is also a feature of forward-time models with selection
+    \item can form the basis of a model-checking procedure for demographic inference methods
+\end{enumerate*}.
 To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following useful transformation.
@@
 Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
-We are interested in quantifying the dependence between the minor allele counts at these sites, particularly for $i\neq j$.
+We are interested in quantifying the dependence between minor allele counts at a given pair of sites, particularly for $i\neq j$.
 
@@
 where, with slight abuse of notation, $p(x) = \sum_y p(x,y)$ and vice versa.
-Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of independent random variables with the same marginal distributions \autocite{CoverThomas1991}.
+Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of two independent random variables with the same marginal distributions as $X$ and $Y$ \autocite{CoverThomas1991}.
 Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.
@@
 We would like to measure the association between alleles at particular frequencies, rather than the overall dependence between allele frequencies at nearby sites (which would be a measure of linkage disequilibrium).
-\Eq{eq:mutual_information} shows that the $I$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
+\Eq{eq:mutual_information} shows that $I[p(x,y)]$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
 \begin{align}
@@
 
-As we have shown above, the folded 2-SFS is a joint probability distribution over minor allele counts.
+As we have shown above, the expected folded 2-SFS is a joint probability mass function over minor allele counts.
 We may thus define the \textit{frequency pointwise mutual information} as
@@
 \end{align}
-One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g the PMI between monomorphic and high minor allele count sites), but we will not use these statistics here.
+One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
 

### 2018-08-05 manuscript/manuscript.tex
Begin Results section.

@@
 
-\newcommand{\floor}[1]{\lfloor #1 \rfloor}
-\newcommand{\E}[1]{\left< #1 \right>}
 \newcommand{\eq}[1]{Eq.~(\ref{#1})}
@@
 \newcommand{\eqs}[2]{Eq.~(\ref{#1})~and~(\ref{#2})}
+\newcommand{\fig}[1]{Fig.~\ref{#1}}
+\newcommand{\Fig}[1]{Figure~\ref{#1}}
+\newcommand{\figs}[2]{Fig.~\ref{#1}~and~\ref{#2}}
+
+\newcommand{\floor}[1]{\lfloor #1 \rfloor}
+\newcommand{\E}[1]{\left< #1 \right>}
+
 \DeclareMathOperator{\pmi}{pmi}
 \DeclareMathOperator{\fpmi}{fpmi}
+\DeclareMathOperator{\wfpmi}{wfpmi}
 \DeclareMathOperator{\hilopmi}{hilopmi}
@@
 
+For plotting purposes, we will also use a weighted version of the $\fpmi$:
+\begin{align}
+    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
+\end{align}
+where $T_2$ is the coalescence time for a sample of size two.
+This weighting emphasizes the most common pairs of minor allele counts, while maintaining invariance to the mutation and pairwise coalescence rates.
+
 \subsection*{Coarse-grained frequencies}
@@
 
-\begin{itemize}
-    \item Want to calculate PMI for 2-SFS for various models.
-    \item implemented Fu and Birkner formulas
-    \item \cite{ZivkovicWiehe2008} extended Fu to variable population size.
-    \item We implemented their formula for exponential growth and piecewise constant, but in principle could be extended to other models.
-    \item Figure 1 shows weighted PMI $\text{wPMI}(i,j) = \E{\eta_{ij}} \text{PMI}(i,j)$.
-    \item weighted to highlight regions of the 2-SFS with the most mutations.
-    \item Observations:
-    \begin{enumerate}
-        \item population growth does not qualitatively change the PMI, even though it distorts the SFS a lot.
-        \item more multiple mergers leads more positive elements.
-        \item Positive associations off-diagonal. Presumably due to uneven branching. Going to ignore because turns out to be sensitive to recombination.
-        \item Positive associations between high- and low-frequency mutations. This is the signal we'll focus on.
-    \end{enumerate}
-    \item Suggests a coarse-graining of the SFS into high- and low-frequency mutations.
-    \item Definition of hi-lo corr given cutoff and distance.
-\end{itemize}
+In this section, we compare the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
+We wrote a \texttt{python} package to compute the SFS, 2-SFS, and $\fpmi$ numerically.
+This package uses the results of \cite{Fu1995} for the time-homogeneous Kingman coalescent, \cite{ZivkovicWiehe2008} for the time-inhomogeneous Kingman coalescent, and \cite{BirknerEtAl2013} for the beta coalescent.
+(See Methods for details.)
 
-[Next: Intuition and $n=4$ calculations.]
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure1.pdf}
+\caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$).
+For all models $n=39$.
+Growth and $\alpha$ parameters were chosen to generate average SFS with similar Tajima's D.
+The diagonal $i=j$ is masked.
+\label{fig:nonrecombining_pmi}}
+\end{figure}
 
-% \begin{figure}
-% \centering
-% \includegraphics[width=0.8\textwidth]{figures/trees.png}
-% \caption{Possible genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue.  \label{fig:trees}}
-% \end{figure}
-%
-% We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
-% Furthermore, the total branch length is independent of the topology.
-% As a result, there is a trade-off between the branch length leading to singleton/tripleton mutations and the branch length leading to doubletons: loci with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
-% Conditional on observing a doubleton at a locus, it is thus more likely that the locus has topology (B) and so the expected number of singletons is lower.
-% In terms of the joint site frequency spectrum, we have $\E{\phi_{12}} < \E{\phi_{1}} \E{\phi_{2}}$.
-%
-% On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
-% For example, topology (C) has less overall opportunity for singletons as well as doubletons than (A) or (B), even though the expected proportion of singletons is higher.
-% Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies more likely.
-% If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) and lead to $\E{\phi_{12}} > \E{\phi_{1}} \E{\phi_{2}}$.
-%
-% There are two limitations of these existing results.
-% First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
-% Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.
+\Fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
+We find that while exponential growth with $g=4$ has a substantial effect on the SFS (Tajima's D = $-0.39 \pi$, where $\pi$ is the average pairwise diversity), it does not qualitatively change the $\fpmi$.
+In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, just as in the constant-$N$ Kingman coalescent.
+
+On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, Tajima's D = $-0.42 \pi$) also generates positive $\fpmi$.
+The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
+This justifies coarse-graining the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.
+
+[fpmi vs tajima's D]
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure2.pdf}
+\caption{[CAPTION] \label{fig:pmi_vs_tajimasd}}
+\end{figure}
+
+\subsection*{Samples of four chromosomes}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/trees.png}
+\caption{Genealogies with a sample size of $n=4$.
+Opportunities for singleton/tripleton mutations are in orange.
+Opportunities for doubletons are in blue.
+(Note, there is a third multiple-merger topology, not shown.) \label{fig:trees}}
+\end{figure}
+
+We can get an intuitive understanding for the result above by considering a sample four chromosomes.
+In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
+Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
+As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
+Genealogies with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
+Conditional on observing a doubleton at a site, it is thus more likely that the genealogy has topology (B) and so the expected number of singletons at sites with the same genealogy is lower than average.
+In terms of the 2-SFS, we have $\E{\eta_{12}} < \E{\eta_{1}} \E{\eta_{2}}$.
+
+On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
+For example, topology (C) has less opportunity for singletons \emph{and} less opportunity for doubletons than (A) or (B), even though the expected proportion of singletons is higher.
+Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies higher.
+If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) so that $\E{\eta_{12}} > \E{\eta_{1}} \E{\eta_{2}}$.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
+\caption{[CAPTION] \label{fig:sdpmi_vs_sdratio}}
+\end{figure}
+
+[Singleton/doubleton fpmi versus singleton/doubleton ratio]
 
 \subsection*{Pointwise mutual information between recombining sites}
+
 [Coalescent simulations with recombination (it still works).]
 
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure4.pdf}
+\caption{[CAPTION] \label{fig:wpmi_recombination}}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure5.pdf}
+\caption{[CAPTION] \label{fig:hilo_vs_d}}
+\end{figure}
+
 \subsection*{Linked selective sweeps}
+
 [Forward-time simulations of linked sweeps (still works).]
@@
 \subsection*{Application to \textit{Drosophila melanogaster}}
+
+[Data processing (see methods). Reduce to $n=100$ per chromosome. Treat chromosomes separately. Get 4-D sites.]
+
+[Compute SFS. Use fastNeutrino to fit a demographic model. Model is:]
+
+[Use msprime to simulate under the model. Compute 2-SFS]
+
+[Compute data hi-lo correlation. Compare to simulations]
 
@@
     \item Could extend to a local method to look for variation in ``multiple-merger-ness''
-    \item Could also extend
+    \item Could also develop a formal hypothesis testing procedure.
+    \item Otherwise, should hilo should be a standard check. Possibly incorporate into model fitting packages.
+    \item Interesting to test across species.
 \end{itemize}
@@
 \begin{itemize}
-    \item Code to compute moments
+    \item Code to compute moments (currently piecewise-constant and exp. growth, but could be extended easily)
     \item Simulation code extending msprime
@@
 \end{document}
+
+% There are two limitations of these existing results.
+% First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
+% Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.

### 2018-08-06 manuscript/manuscript.tex
Finish non-recombining sites.

@@
 
-[fpmi vs tajima's D]
+Accordingly, we computed the coarse-grained $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
+For the same parameters, we also calculated a normalized Tajima's D statistic:
+$(\hat{\theta}_W - \pi) / \pi$, where $\hat{\theta}_W$ is Watterson's estimator of the coalescent scaled mutation rate \autocite{Watterson19??}.
+The latter is a measure of the distortions in the SFS induced by population growth or multiple mergers and equals zero for the constant-$N$ Kingman coalescent.
+\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
+On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
+Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.
 
@@
 \includegraphics[width=0.8\textwidth]{figures/figure2.pdf}
-\caption{[CAPTION] \label{fig:pmi_vs_tajimasd}}
+\caption{Pointwise mutual information between singletons and non-singletons for Kingman coalescents with exponential growth and beta coalescents.
+The exponential growth rate, $g$, ranges from 0.25 to 8.0 in coalescent time units.
+The beta coalescent parameter, $\alpha$, ranges from 1.75 (nearly Kingman) to 1.25 (nearly Bolthausen-Sznitman).
+\label{fig:fpmi_vs_tajimasd}}
 \end{figure}
@@
 Opportunities for doubletons are in blue.
-(Note, there is a third multiple-merger topology, not shown.) \label{fig:trees}}
+(Note: there is a third possible multiple-merger topology, not shown.) \label{fig:trees}}
 \end{figure}
@@
 \includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
-\caption{[CAPTION] \label{fig:sdpmi_vs_sdratio}}
+\caption{Distortions in $\fpmi$ vs.
+distortions in the SFS for $n=4$.
+Parameters for beta coalescents and exponential growth are as in \fig{fig:fpmi_vs_tajimasd}.
+For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^-4$ to 2 in coalescent time units.
+\label{fig:sdpmi_vs_sdratio}}
 \end{figure}
 
-[Singleton/doubleton fpmi versus singleton/doubleton ratio]
+As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
+\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
+\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
+In the latter model, we vary both the fold-change in $N$ and the time of the change.
+Like Tajima's D in the previous section, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
+(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
+As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
+Moreover, the two different models of growth collapse onto one another.
+This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
 

### 2018-08-06 manuscript/manuscript.tex
Add manuscript to git repo.
Add more results.

@@
 
-[Coalescent simulations with recombination (it still works).]
+The previous sections have shown that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
+However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
+In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy exactly and calculate tree imbalance statistics directly (see e.g., \cite{Seger2010}).
+The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
+Therefore, it is important to assess the robustness of our approach to recombination between sites.
+
+To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KehellerEtAl201?} modified to allow for multiple mergers.
+In particular, we simulated a model where coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
+In this model, marginal genealogies will follow the beta coalescent distribution, and the average SFS will be given by the formula in \cite{BirknerEtAl2013}.
+We also simulated data from two models of population growth: exponential growth and a two-epoch piecewise-constant model.
+
+For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
+We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
+For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
+We then calculated the expectations $\{\E{\tau_i\}}$ and $\{\E{\tau_ij}\}$ by averaging over independent loci.
+These expectations allow us to calculate all of the 2-SFS statistics defined above.
 
@@
 \includegraphics[width=0.8\textwidth]{figures/figure4.pdf}
-\caption{[CAPTION] \label{fig:wpmi_recombination}}
+\caption{Weighted frequency pointwise information for three recombination rates with sample size $n=100$.
+As in \Fig{fig:nonrecombining_pmi}, the diagonal elements are masked, and the growth and multiple-merger parameters were chosen to generate average SFS with similar Tajima's D.
+\label{fig:wpmi_recombination}}
 \end{figure}
+
+\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
+As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
+In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
+This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
+These positive correlations, however, do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
+On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
 
@@
 \includegraphics[width=0.8\textwidth]{figures/figure5.pdf}
-\caption{[CAPTION] \label{fig:hilo_vs_d}}
+\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus genetic distance.
+The first three rows show coalescent simulations in \texttt{msprime}.
+The fourth row shows forward-time \texttt{SLiM} simulations with selection at linked sites.
+Lines are colored by the distortion in the SFS, as measured by Tajima's D.
+[List parameters here?] \label{fig:hilo_vs_d}}
 \end{figure}
+
+\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d=r\E{T_2}$ and three different choices of $i_c$.
+Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
+At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
+When $r\E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
+In all Kingman models, $\hilopmi$ decays to zero for $r\E{T_2}\gg 1$.
+
+In contrast, the beta coalescent generates $\hilopmi$ that is consistently greater than the constant-$N$ Kingman.
+As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model then in models of population growth than generate similar distortions in the SFS.
+[IDEA: A plot of $\hilopmi$ at a fixed r vs Tajima's D for all the models together. Could make several plots, varying $r$ and $i_c$. This might be more convincing than the current figure.]
+This is true across recombination rates and high/low cutoff minor allele counts.
+These results demonstrate that $\hilopmi$ is capable of discriminating between coalescent models even when there is recombination between sites.
+
+There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
+This is related to the results of \cite{EldonWakeley20??}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
+They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
+Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
+We do not expect this behavior to be universal to multiple mergers coalescents with recombination.
 
@@
 
-[Forward-time simulations of linked sweeps (still works).]
+Various authors have shown that natural selection can generate multiple merger coalescents at linked neutral sites (e.g., \cite{DurrettSchweinsberg2005, CoopRalph, NeherHallatscheck2013, DesaiEtAl, Seger}).
+However, our simulations of the beta coalescent with recombination are of an explicitly neutral model.
+Thus, they are at best an approximation to the selective models cited above.
+It is therefore important to verify that selection at linked sites can, in fact, generate the sorts of signals in $\fpmi$ that we have detected in the beta coalescent.
 
-\subsection*{Application to \textit{Drosophila melanogaster}}
+To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
+In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with fixed selection coefficient $s$.
+Both types of mutations occurred at random, uniformly distributed across the length of the chromosome.
+By varying the recombination rate, mutation rates, population size, and selection coefficient, we were able to vary the rate of selective sweeps linked to neutral sites.
+After the populations reached steady-state, we sampled individuals and calculated $\{\eta_i\}$ and $\{\eta_{ij}(d)\}$ for the neutral mutations.
+(See Methods for details.)
 
-[Data processing (see methods). Reduce to $n=100$ per chromosome. Treat chromosomes separately. Get 4-D sites.]
+The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
+For intermediate genetic distances, $d r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
+That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
+The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d r \E{T_2} \gg 1$.
+This is reasonable because the effect of a particular sweep on genealogies should be localized around the beneficial mutation.
 
-[Compute SFS. Use fastNeutrino to fit a demographic model. Model is:]
+\subsection*{Coalescent model checking: application to \textit{Drosophila melanogaster}}
 
-[Use msprime to simulate under the model. Compute 2-SFS]
+Our results above show that $\fpmi$ and its coarse-grained analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
+This is true even when the population growth generates similar distortions in the SFS.
+We therefore propose the following model-checking procedure for demographic inference methods
+\begin{enumerate}[label=(\roman*), before=\unskip{: }]
+    \item Fit a demographic model to data using any relevant method.
+    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
+    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
+    \item Compare true to predicted statistics. Evaluate model fit.
+\end{enumerate}
+This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
+Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.
 
-[Compute data hi-lo correlation. Compare to simulations]
+In this section, we illustrate the procedure just outlined by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
+The DPGP3 data consists of haploid consensus sequences from $\sim 200$ wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
+Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
+On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{Elyashiv??201?, GarudPetrov2016, others?})
+Thus, this data is a good candidate for demonstrating the utility of $\fpmi$ for assessing coalescent model fit.
+
+After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
+Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
+We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtal2015}.
+We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
+We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{table:DPGP_params} [Make table].
+We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure6.pdf}
+\caption{DPGP3 data and coalescent model predictions.
+First row: observed (blue) and expected (orange) site frequency spectrum compared to constant-$N$ Kingman coalescent (dotted lines).
+Second row: Expected (upper triangle) and observed (lower triangle) weighted $\fpmi$ averaged over all pairs of sites less than $15 d_c$ apart.
+Third row: Observed (blue) and expected (orange) $\hilopmi$ versus genetic distance.
+Cutoff minor allele count, $i_c, = 5$.
+Solid blue curves show cubic spline fit.
+Dotted lines show the expectation under the constant-$N$ Kingman coalescent.
+[Middle row needs colorbar] \label{fig:dpgp3}}
+\end{figure}
+
+\begin{itemize}
+    \item Expected and observed SFS match (because we fit the model to that).
+    \item Both show excess of low-frequency SNPs.
+    \item At intermediate distance, can see that the $\fpmi$ is strongly positive in the data, but negative in the model.
+    \item This pattern extends to other genetic distances.
+    \item Correlations in the data decay at long distances. This is like the sweeps simulations, but we don't want to draw too strong a conclusion.
+\end{itemize}
 
@@
 \begin{itemize}
-    \item Focus here on coalescent, but also relevant for diffusion methods
-    \item Single-locus tests that rely on trees (Seger whale lice, Neher with viruses)
-    \item Multi-locus tests that compare vs genomic background
-    \item Could extend to a local method to look for variation in ``multiple-merger-ness''
-    \item Could also develop a formal hypothesis testing procedure.
-    \item Otherwise, should hilo should be a standard check. Possibly incorporate into model fitting packages.
-    \item Interesting to test across species.
+    \item Review results.
+    \item Availability of code: models for computing moments numerically, scripts for running simulations. Notebooks for data analysis.
+    \item We focus here on coalescent-based methods, but also relevant for diffusion methods (because they imply a coalescent model)
+    \item Mention methods to fit multiple mergers coalescent to data (computationally difficult).
+Not really what we're trying to do here.
+Our goal is that you can exploit this extra dimension of the data you already have to check your model and make sure you're not fitting a signal of selection, etc with a demographic model.
+    \item Could also develop a formal hypothesis testing procedure, using the bootstrap, etc.
+    \item Compare with: Single-locus tests that rely on building trees (Seger whale lice, Neher with viruses)
+    \item Compare with: Multi-locus tests that compare a particular region to genomic background. (What if the background isn't neutral/Kingman?)
+    \item But: we could extend our method to a local method that looks for variation in ``multiple-merger-ness''
+    \item Multiple-mergers coalescent and recombination: open area of research
+    \item $\fpmi(d)$ should be a standard posterior check for demographic models. Possibly incorporate into model fitting packages.
+    \item Interesting to compare across species.
 \end{itemize}
-
-Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
 
@@
     \begin{itemize}
-        \item data pre-processing
-        \item fitting demographic model with fastNeutrino
-        \item simulating demographic model (could maybe compute exactly)
+        \item data pre-processing: throw out inversions, downsampling to 100 chromosome, 4-D sites, filter for coverage
+        \item picking the central regions
+        \item fitting demographic model with fastNeutrino: fit 3-epoch piecewise constant to SFS. Report model parameters
+        \item simulating demographic model
+        \item computing statistics
     \end{itemize}
@@
 
+% Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
+
 % There are two limitations of these existing results.

### 2018-08-07 manuscript/manuscript.tex
Finish results. Begin methods.

@@
 We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
-We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{table:DPGP_params} [Make table].
+We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{tab:DPGP_params} [Make table].
 We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}).
@@
 
-\begin{itemize}
-    \item Expected and observed SFS match (because we fit the model to that).
-    \item Both show excess of low-frequency SNPs.
-    \item At intermediate distance, can see that the $\fpmi$ is strongly positive in the data, but negative in the model.
-    \item This pattern extends to other genetic distances.
-    \item Correlations in the data decay at long distances. This is like the sweeps simulations, but we don't want to draw too strong a conclusion.
-\end{itemize}
+The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
+In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
+Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
+This pattern extends across a range of genomic distances, $d/d_c \in (10^{-1}, 10^2)$ (\fig{fig:dpgp3}, third row).
+As a result, we may conclude that the data is not well explained by the Kingman coalescent with population growth.
+
+Note that the $\hilopmi$ decays toward zero at large distances.
+This matches the expectation from simulations with selective sweeps rather than the beta coalescent.
+However, we caution against concluding that sweeps are necessarily responsible for the deviations from the Kingman expectation.
 
@@
 \section*{Methods}
-\begin{itemize}
-    \item Code to compute moments (currently piecewise-constant and exp. growth, but could be extended easily)
-    \item Simulation code extending msprime
-    \item Forward-time simulation with SLiM
-    \item Drosophila
+
+\subsection*{Computing branch-length moments}
+
+We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$ using the \texttt{python} numerical package \texttt{numpy} \autocite{numpy}.
+For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
+To calculate the second moments of the coalescent time distribution for exponentially growing populations, we used Gaussian quadrature.
+For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.
+
+Functions for compute the branch-length moments are available [GITHUB].
+The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
+The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
+They are therefore only practical for sample sizes up to $n \approx 40$.
+The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ [CHECK THIS] and is also only practical for samples up to $n \approx 50$.
+
+\subsection*{Coalescent simulations}
+
+Our beta coalescent simulation code is based on modifications to the \texttt{msprime} package made by Joe Zhu [link to github] to allow for multiple mergers.
+This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
+
+[Possibly move some of the description here]
+
+\subsection*{Forward-time simulations of selective sweeps}
+
+We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
+In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
+Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
+We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
+With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
+Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
+Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.
+
+In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
+For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.
+
+\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].
+
+\subsection*{Analysis of \textit{D. melanogaster} data}
+
+The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
+We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
+These files contain the sequence alignments of all flies in the sample on all chromosome arms.
+We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
+For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order of sample name.
+(This means that the data for each chromosome arm is from a slightly different subset of the individuals.)
+
 \begin{itemize}
@@
 \end{itemize}
-\end{itemize}
 
@@
 
+\section*{Tables}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{[Table of parameters underlying Figure 2]}
+    \label{tab:nonrec_params}
+    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
+      % $\alpha$ & $\beta$ & $\gamma$ \\
+      % \hline
+      % 1 & 1110.1 & a\\
+      % 2 & 10.1 & b\\
+      % 3 & 23.113231 & c\\
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{[Table of parameters underlying Figure 4]}
+    \label{tab:n4_params}
+    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
+      % $\alpha$ & $\beta$ & $\gamma$ \\
+      % \hline
+      % 1 & 1110.1 & a\\
+      % 2 & 10.1 & b\\
+      % 3 & 23.113231 & c\\
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{[Table of parameters underlying Figure 6]}
+    \label{tab:rec_params}
+    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
+      % $\alpha$ & $\beta$ & $\gamma$ \\
+      % \hline
+      % 1 & 1110.1 & a\\
+      % 2 & 10.1 & b\\
+      % 3 & 23.113231 & c\\
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{[DPGP3 model fit]}
+    \label{tab:DPGP_params}
+    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
+      % $\alpha$ & $\beta$ & $\gamma$ \\
+      % \hline
+      % 1 & 1110.1 & a\\
+      % 2 & 10.1 & b\\
+      % 3 & 23.113231 & c\\
+    \end{tabular}
+  \end{center}
+\end{table}
+
 \end{document}
 
+TODO:
+- Abstract
+- Drosophila methods
+- Discussion
+- Parameter tables:
+    -
+- bibtex file
+- SI figures:
+    - Drosophila diversity
+    - Drosophila # called sites
+- Code
 % Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).

### 2018-08-07 manuscript/manuscript.tex
Finish methods.

@@
 The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
+The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
 We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
@@
 We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
-For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order of sample name.
-(This means that the data for each chromosome arm is from a slightly different subset of the individuals.)
+For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
+(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)
 
-\begin{itemize}
-    \item data pre-processing: throw out inversions, downsampling to 100 chromosome, 4-D sites, filter for coverage
-    \item picking the central regions
-    \item fitting demographic model with fastNeutrino: fit 3-epoch piecewise constant to SFS. Report model parameters
-    \item simulating demographic model
-    \item computing statistics
-\end{itemize}
+We calculated the average pairwise diversity, $\pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
+Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
+Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
+Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\pi$ for further analysis.
+The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
+
+In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites where fewer than 90 of the 100 samples have genotype calls.
+This leaves over 90\% of all sites and does not substantially alter the fraction of sites that are polymorphic in the sample (Table~\ref{tab:called_sites})
+
+We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
+We fit a three-epoch piecewise constant--$N$ model:
+\begin{align}
+    N(t) = \begin{cases}
+                N_1 & 0 \leq t \leq t_1 \\
+                N_2 & t_1 <  t \leq t_2 \\
+                N_{\text{anc}} & t > t_2,
+            \end{cases}
+    \label{eq:piecewise}
+\end{align}
+estimating both change-points ($t_1$ < $t_2$) and both population sizes ($N_1$, $N_2$).
+We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
+For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
+Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
+We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.
+
+In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
+Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
+A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available [GITHUB].
 
@@
 
-\section*{Tables}
+\clearpage
+
+\section*{Supplementary Figures and Tables}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/pi_vs_position.pdf}
+\caption{Average pairwise diversity, $\pi$, of 100 haploid samples from DPGP3 panel.
+Each point represents the average over a 10~KB window.
+Vertical lines show the boundaries of the central regions defined in Table~\ref{tab:central_regions}.
+\label{fig:dpgp_pi}}
+\end{figure}
 
@@
   \begin{center}
-    \caption{[DPGP3 model fit]}
+    \caption{Boundaries of the homogeneous central regions of chromosomes (shown in \ref{fig:dpgp_pi}).}
+    \label{tab:central_regions}
+    \begin{tabular}{l|r|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      \textbf{Chr.} & \textbf{Start position (MB)} & \textbf{End position (MB)}\\
+      \hline
+      2L & 1 & 17\\
+      2R & 6 & 19\\
+      3L & 1 & 17\\
+      3R & 10 & 26
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{Fraction of sites with at least 90 genotyped samples}
+    \label{tab:called_sites}
+    \begin{tabular}{l|r|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      \textbf{Chr.} & \textbf{All sites} & \textbf{Polymorphic sites}\\
+      \hline
+      2L & 0.913 & 0.924 \\
+      2R & 0.922 & 0.925 \\
+      3L & 0.919 & 0.920 \\
+      3R & 0.933 & 0.936
+    \end{tabular}
+  \end{center}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \caption{Piecewise-constant model parameters fit to the SFS of each chromosome arm (see \eq{eq:piecewise}, $N_{\text{anc}}=3 \times 10^5$).}
     \label{tab:DPGP_params}
-    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
-      % $\alpha$ & $\beta$ & $\gamma$ \\
-      % \hline
-      % 1 & 1110.1 & a\\
-      % 2 & 10.1 & b\\
-      % 3 & 23.113231 & c\\
+    \begin{tabular}{l|r|r|r|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+        \textbf{Chr.} & \textbf{$N_1 (\times 10^5)$} & \textbf{$N_2 (\times 10^5)$} & \textbf{$t_1 (\times 10^4)$} & \textbf{$t_2 (\times 10^4)$}\\
+        \hline
+        2L & 10.7 & 4.6 & 2.8 & 39.9 \\
+        2R &  9.1 & 3.9 & 1.6 &  8.8 \\
+        3L &  7.3 & 3.8 & 1.3 &  6.4 \\
+        3R &  9.7 & 5.1 & 2.2 &  8.6
     \end{tabular}
@@
 - Abstract
-- Drosophila methods
 - Discussion
+- Edit figures
 - Parameter tables:
-    -
-- bibtex file
-- SI figures:
-    - Drosophila diversity
-    - Drosophila # called sites
-- Code
+    - fig 2
+    - fig 4
+    - fig 6
+- Get rest of citations
+- Code:
+    - organize
+    - clean
+    - comment
+    - post on Github
+
+Things that were done but not included in this ms:
+- Background selection (didn't run the right parameters to see the effect. Might be computationally challenging.)
+- 2-Deme model with migration. (Could include in SI.)
+- 1-D stepping-stone.
+
+OLD TEXT:
 % Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).

### 2018-08-07 manuscript/manuscript.tex

@@
 Parameters for beta coalescents and exponential growth are as in \fig{fig:fpmi_vs_tajimasd}.
-For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^-4$ to 2 in coalescent time units.
+For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^{-4}$ to 2 in coalescent time units.
 \label{fig:sdpmi_vs_sdratio}}
@@
 The first three rows show coalescent simulations in \texttt{msprime}.
-The fourth row shows forward-time \texttt{SLiM} simulations with selection at linked sites.
+Exponential growth rate $g$ ranges from 0.01 to 8.0 in coalescent time units.
+Piecewise constant change-points, range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2.
+The fourth row shows forward-time \texttt{SLiM} simulations with selective sweeps at linked sites (beneficial mutation fraction $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$, see Methods).
 Lines are colored by the distortion in the SFS, as measured by Tajima's D.
-[List parameters here?] \label{fig:hilo_vs_d}}
+\label{fig:hilo_vs_d}}
 \end{figure}
@@
 We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
-We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{tab:DPGP_params} [Make table].
-We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}).
+We report our fitted parameters in Table~\ref{tab:DPGP_params}.
+We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}, See methods for details.).
 
@@
 Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
+For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
+We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{dmelrec}.
+We estimated $\E{T_2}$ by $\pi / 2 \mu$.
+For $\pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
+For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{dmelmu}.
+These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.
+
 A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available [GITHUB].
@@
 \end{figure}
-
-\begin{table}[h!]
-  \begin{center}
-    \caption{[Table of parameters underlying Figure 2]}
-    \label{tab:nonrec_params}
-    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
-      % $\alpha$ & $\beta$ & $\gamma$ \\
-      % \hline
-      % 1 & 1110.1 & a\\
-      % 2 & 10.1 & b\\
-      % 3 & 23.113231 & c\\
-    \end{tabular}
-  \end{center}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \caption{[Table of parameters underlying Figure 4]}
-    \label{tab:n4_params}
-    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
-      % $\alpha$ & $\beta$ & $\gamma$ \\
-      % \hline
-      % 1 & 1110.1 & a\\
-      % 2 & 10.1 & b\\
-      % 3 & 23.113231 & c\\
-    \end{tabular}
-  \end{center}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \caption{[Table of parameters underlying Figure 6]}
-    \label{tab:rec_params}
-    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
-      % $\alpha$ & $\beta$ & $\gamma$ \\
-      % \hline
-      % 1 & 1110.1 & a\\
-      % 2 & 10.1 & b\\
-      % 3 & 23.113231 & c\\
-    \end{tabular}
-  \end{center}
-\end{table}
 
@@
 - Edit figures
-- Parameter tables:
-    - fig 2
-    - fig 4
-    - fig 6
+- Consistency of nomenclature, labeling, etc.
+- Check drosophila numbers. Get citations or remake figure.
+- Check references to Methods. (Should all be there and all have corresponding methods section.)
 - Get rest of citations

### 2018-08-13 manuscript/manuscript.tex
Add some references.

@@
 For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
-We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{dmelrec}.
+We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
 We estimated $\E{T_2}$ by $\pi / 2 \mu$.
 For $\pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
-For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{dmelmu}.
+For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
 These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.

### 2018-08-15 manuscript/manuscript.tex
Rearrange intro. (Need to re-write to condense/smooth)

@@
 Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
-The fundamental tools of the trade are toy models, which are used to study how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
+The fundamental tools of the trade are simplified mathematical models, which connect historical quantities such as the population size to observable features of genetic data, and thus form the basis of parametric inference methods.
 However, populations are complicated and, moreover, varied in their complications.
@@
 
-In its simplest form, the Kingman coalescent has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
-Under a variety of conditions, the coalescent rate is inversely proportional to the size of the population~\autocite{Kingman1982b}.
+The Kingman coalescent underlies a variety of methods for inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
+In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
+Under a variety of conditions, the coalescent rate is inversely proportional to the population size, $N$~\autocite{Kingman1982b}.
 Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
-This observation is the basis for a variety of methods to infer historical population sizes from genetic data.
+Observable properties of genetic data depend on the ratio of the coalescent rate to other evolutionary rate parameters.
+For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
+Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains some features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or pattern of heterozygocity along the chromosome (e.g., \cite{LiDurbin2011}).
 
-With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
-One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
-The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
-Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
-Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
-Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
-(Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)
-
-A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
+A serious problem for this class of inference procedures is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
 One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
 When this assumption is violated, Kingman-based inference methods are misspecified.
-For example, \cite{SchriderEtAl2016} recently demonstrated how population-size inference can be distorted by selective sweeps.
-This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
-In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
-Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
+For example, when a beneficial mutation increases rapidly in frequency, it distorts the gene genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
+When such selective sweeps are common, a large fraction of coalescence events may occur during sweeps.
+In this case, the average coalescent rate is proportional to the beneficial mutation supply rate, which is itself \emph{directly} rather than inversely proportional to $N$.
+Thus, the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.
+
+While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
+For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
+In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
+Moreover, genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
 
@@
 fat-tailed offspring number distributions \autocite{Schweinsberg2003},
-recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
+recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
 rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
 and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
+
 In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
-For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
+For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
 In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
+
 It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
+This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{SelectionScanReview}).
+These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic history.
+Genomic regions that are outliers from this distribution are presumed to be under selection.
+In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.
 
-One possible approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
-(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but, as with other ``exact'' inference methods, it does not scale to genomic data.)
+One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
 \cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
@@
 
-Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby sites---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
 We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
@@
 
-\section*{Definitions and previously known results}
+[STREAMLINE THE FOLLOWING:]
 
-\subsection*{The SFS and 2-SFS}
 Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
@@
 
+\section*{Methods}
+
 \subsection*{Frequency pointwise mutual information}
+
+[DEFINITION FIRST, THEN EXPLANATION]
 
@@
 One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
+
+\subsection*{Computing branch-length moments}
+
+We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$ using the \texttt{python} numerical package \texttt{numpy} \autocite{numpy}.
+For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
+To calculate the second moments of the coalescent time distribution for exponentially growing populations, we used Gaussian quadrature.
+For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.
+
+Functions for compute the branch-length moments are available [GITHUB].
+The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
+The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
+They are therefore only practical for sample sizes up to $n \approx 40$.
+The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ [CHECK THIS] and is also only practical for samples up to $n \approx 50$.
+
+\subsection*{Coalescent simulations}
+
+Our beta coalescent simulation code is based on modifications to the \texttt{msprime} package made by Joe Zhu [link to github] to allow for multiple mergers.
+This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
+
+[Possibly move some of the description here]
+
+\subsection*{Forward-time simulations of selective sweeps}
+
+We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
+In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
+Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
+We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
+With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
+Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
+Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.
+
+In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
+For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.
+
+\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].
+
+\subsection*{Analysis of \textit{D. melanogaster} data}
+
+The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
+The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
+We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
+These files contain the sequence alignments of all flies in the sample on all chromosome arms.
+We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
+For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
+(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)
+
+We calculated the average pairwise diversity, $\pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
+Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
+Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
+Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\pi$ for further analysis.
+The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
+
+In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites where fewer than 90 of the 100 samples have genotype calls.
+This leaves over 90\% of all sites and does not substantially alter the fraction of sites that are polymorphic in the sample (Table~\ref{tab:called_sites})
+
+We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
+We fit a three-epoch piecewise constant--$N$ model:
+\begin{align}
+    N(t) = \begin{cases}
+                N_1 & 0 \leq t \leq t_1 \\
+                N_2 & t_1 <  t \leq t_2 \\
+                N_{\text{anc}} & t > t_2,
+            \end{cases}
+    \label{eq:piecewise}
+\end{align}
+estimating both change-points ($t_1$ < $t_2$) and both population sizes ($N_1$, $N_2$).
+We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
+For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
+Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
+We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.
+
+In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
+Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
+For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
+We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
+We estimated $\E{T_2}$ by $\pi / 2 \mu$.
+For $\pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
+For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
+These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.
+
+A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available [GITHUB].
 
@@
 
-\section*{Methods}
+\printbibliography
 
-\subsection*{Computing branch-length moments}
-
-We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$ using the \texttt{python} numerical package \texttt{numpy} \autocite{numpy}.
-For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
-To calculate the second moments of the coalescent time distribution for exponentially growing populations, we used Gaussian quadrature.
-For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.
-
-Functions for compute the branch-length moments are available [GITHUB].
-The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
-The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
-They are therefore only practical for sample sizes up to $n \approx 40$.
-The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ [CHECK THIS] and is also only practical for samples up to $n \approx 50$.
-
-\subsection*{Coalescent simulations}
-
-Our beta coalescent simulation code is based on modifications to the \texttt{msprime} package made by Joe Zhu [link to github] to allow for multiple mergers.
-This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
-
-[Possibly move some of the description here]
-
-\subsection*{Forward-time simulations of selective sweeps}
-
-We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
-In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
-Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
-We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
-With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
-Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
-Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.
-
-In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
-For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.
-
-\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].
-
-\subsection*{Analysis of \textit{D. melanogaster} data}
-
-The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
-The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
-We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
-These files contain the sequence alignments of all flies in the sample on all chromosome arms.
-We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
-For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
-(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)
-
-We calculated the average pairwise diversity, $\pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
-Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
-Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
-Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\pi$ for further analysis.
-The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
-
-In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites where fewer than 90 of the 100 samples have genotype calls.
-This leaves over 90\% of all sites and does not substantially alter the fraction of sites that are polymorphic in the sample (Table~\ref{tab:called_sites})
-
-We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
-We fit a three-epoch piecewise constant--$N$ model:
-\begin{align}
-    N(t) = \begin{cases}
-                N_1 & 0 \leq t \leq t_1 \\
-                N_2 & t_1 <  t \leq t_2 \\
-                N_{\text{anc}} & t > t_2,
-            \end{cases}
-    \label{eq:piecewise}
-\end{align}
-estimating both change-points ($t_1$ < $t_2$) and both population sizes ($N_1$, $N_2$).
-We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
-For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
-Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
-We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.
-
-In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
-Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
-For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
-We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
-We estimated $\E{T_2}$ by $\pi / 2 \mu$.
-For $\pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
-For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
-These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.
-
-A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available [GITHUB].
-
-\printbibliography
+\section*{Supplementary Methods}
+[IF NECESSARY]
 
@@
 % Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.
+
+% With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
+% One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
+% The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
+% Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
+% Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
+% Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
+% (Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)

### 2018-08-15 manuscript/manuscript.tex
Edit the introduction.

@@
 Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
-The fundamental tools of the trade are simplified mathematical models, which connect historical quantities such as the population size to observable features of genetic data, and thus form the basis of parametric inference methods.
+The fundamental tools of the trade are simplified mathematical models, which connect unobserved quantities such as the population size to observable features of genetic data.
 However, populations are complicated and, moreover, varied in their complications.
@@
 
-One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
+One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
 The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
-Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
+Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
 The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
@@
 
-The Kingman coalescent underlies a variety of methods for inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
+An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
 In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
-Under a variety of conditions, the coalescent rate is inversely proportional to the population size, $N$~\autocite{Kingman1982b}.
+Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982b}.
 Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
-Observable properties of genetic data depend on the ratio of the coalescent rate to other evolutionary rate parameters.
+Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
 For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
-Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains some features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or pattern of heterozygocity along the chromosome (e.g., \cite{LiDurbin2011}).
+Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygocity along a chromosome (e.g., \cite{LiDurbin2011}).
 
-A serious problem for this class of inference procedures is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
+A serious problem for this class of inference methods is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
 One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
 When this assumption is violated, Kingman-based inference methods are misspecified.
-For example, when a beneficial mutation increases rapidly in frequency, it distorts the gene genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
-When such selective sweeps are common, a large fraction of coalescence events may occur during sweeps.
-In this case, the average coalescent rate is proportional to the beneficial mutation supply rate, which is itself \emph{directly} rather than inversely proportional to $N$.
-Thus, the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.
+For example, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
+If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
+In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly} rather than inversely proportional to the population size.
+It follows that the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.
 
@@
 In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
-Moreover, genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
+Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
 
@@
 Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
-The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
+The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
 These models are relevant for species with
@@
 For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
-In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
+In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.
 
 It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
-This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{SelectionScanReview}).
-These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic history.
+This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{VittiEtAl2013}).
+These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
 Genomic regions that are outliers from this distribution are presumed to be under selection.
@@
 Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
 We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
@@
 Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
-Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
+We demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
 
-[STREAMLINE THE FOLLOWING:]
-
-Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
-Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
+Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
 In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
@@
 %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
+The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
 
-The SFS and folded SFS are single-site statistics.
-As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
-We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
-$\left\{ \xi_{ij}(d) : 1 \leq i, j < n\right\}$,
+In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
+We define the 2-SFS,
+$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
 as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
@@
 This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
-By analogy to the folded SFS, we define the folded 2-SFS, as
-\begin{align*}
-    \eta_{ij}(d) &= \xi_{ij}(d) \\
-                 &+ (1-\delta_{i,n-i}) \xi_{n-i,j}(d) \\
-                 &+ (1-\delta_{j,n-j}) \xi_{i,n-j}(d) \\
-                 &+ (1-\delta_{i,n-i}) (1-\delta_{j,n-j}) \xi_{n-i,n-j}(d).
-\end{align*}
-In this section, we will be considering non-recombining sites and so will suppress the distance $d$.
+We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.
 
-In the limit of low per-site mutation rate ($\mu\to0$), all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
+In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
 \begin{align}
@@
 where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
+(We suppress the distance, $d$, for non-recombining sites.)
 Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.
@@
 \cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
-He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
-This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
-(There are positive correlations between mutations at complimentary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)
+He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \not\in \{i, (n-i)\}$.
+This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \not\in \{i, (n-i)\}$.
+(There are positive correlations between mutations at complementary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)
 
@@
 
-Our goal is to demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
+In the following, we demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
 \begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
@@
 \end{enumerate*}.
-To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following useful transformation.
+
+% To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following transformation.
 

### 2018-08-16 manuscript/manuscript.tex
Edit Methods (moving to front of the paper).

@@
 
-% To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following transformation.
-
 \section*{Methods}
@@
 
-[DEFINITION FIRST, THEN EXPLANATION]
+We are interested in quantifying the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
+To do so, we define the \textit{frequency pointwise mutual information} as
+\begin{align}
+    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
+    \label{eq:fpmi}
+\end{align}
+\Fig{fig:schematic} shows the steps to compute $\fpmi$ from a sample of sequences.
 
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{figures/schematic.pdf}
+\caption{Summarizing population genetic data with frequency pointwise information.
+First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations.
+Then, for a given distance, $d$, the 2-SFS gives the number of pairs of sites with each combination of minor allele frequencies.
+Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
+[OPTIONS: Could add (1) weighted fpmi, and/or (2) binned 2-SFS] \label{fig:schematic}}
+\end{figure}
+
+In information theory, the pointwise mutual information of a pair of random variables (PMI), $X$ and $Y$, is a transformation of the probability mass function, $p(x,y)$, given by
+\begin{align}
+    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)}
+    \label{eq:pmi}
+\end{align}
+where $p(x) = \sum_y p(x,y)$ and vice versa \autocite{ChurchHanks1990}.
+PMI measures the change in the probability that $X=x$ given knowledge that $Y=y$.
+When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
+On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
+(The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.)
+
+In our setting, the 2-SFS may be interpreted as a joint probability mass function over allele frequencies.
 Given a coalescent model, the minor allele count at an arbitrary site is a random variable with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
 Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
-We are interested in quantifying the dependence between minor allele counts at a given pair of sites, particularly for $i\neq j$.
+Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of allele frequencies.
 
-A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
-The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
-\begin{align}
-    I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
-    \label{eq:mutual_information}
-\end{align}
-where, with slight abuse of notation, $p(x) = \sum_y p(x,y)$ and vice versa.
-Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of two independent random variables with the same marginal distributions as $X$ and $Y$ \autocite{CoverThomas1991}.
-Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.
-
-We would like to measure the association between alleles at particular frequencies, rather than the overall dependence between allele frequencies at nearby sites (which would be a measure of linkage disequilibrium).
-\Eq{eq:mutual_information} shows that $I[p(x,y)]$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
-\begin{align}
-    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)} = \log \frac{p(y|x)}{p(y)}
-\end{align}
-\autocite{ChurchHanks1990}.
-Pointwise mutual information measures the change in the probability that $X=x$ given knowledge that $Y=y$.
-When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
-On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
-
-As we have shown above, the expected folded 2-SFS is a joint probability mass function over minor allele counts.
-We may thus define the \textit{frequency pointwise mutual information} as
-\begin{align}
-    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
-\end{align}
 This transformation of the 2-SFS has several useful properties.
@@
 These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
-Our hope is that the 2-SFS contains additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
+Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
+That is, we expect $\fpmi$ to be sensitive to multiple mergers, but invariant under changes in the coalescent rate.
 
-For plotting purposes, we will also use a weighted version of the $\fpmi$:
-\begin{align}
-    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
-\end{align}
-where $T_2$ is the coalescence time for a sample of size two.
-This weighting emphasizes the most common pairs of minor allele counts, while maintaining invariance to the mutation and pairwise coalescence rates.
+% A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
+% The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
+% \begin{align}
+%     I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
+%     \label{eq:mutual_information}
+% \end{align}
+% where $p(x) = \sum_y p(x,y)$ and vice versa.
+% Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of two independent random variables with the same marginal distributions as $X$ and $Y$ \autocite{CoverThomas1991}.
+% Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.
 
-\subsection*{Coarse-grained frequencies}
+% We would like to measure the association between alleles at particular frequencies, rather than the overall dependence between allele frequencies at nearby sites (which would be a measure of linkage disequilibrium).
+% \Eq{eq:mutual_information} shows that $I[p(x,y)]$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
+% \begin{align}
+%     \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)} = \log \frac{p(y|x)}{p(y)}
+% \end{align}
+% \autocite{ChurchHanks1990}.
+% Pointwise mutual information measures the change in the probability that $X=x$ given knowledge that $Y=y$.
+% When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
+% On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
+
+\subsection*{Binned allele frequencies}
 
@@
 We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
-This suggests a coarse-grained SFS and 2-SFS:
+This suggests a binned SFS and 2-SFS:
 \begin{align}
@@
 where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
-Coarse-graining allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
+Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
 
@@
 \end{align}
-One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
+One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
+Also, it is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.
+
+\subsection*{Weighted $\fpmi$}
+For plotting purposes, we will also use a weighted version of the $\fpmi$:
+\begin{align}
+    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
+    \label{eq:wfpmi}
+\end{align}
+where $T_2$ is the coalescence time for a sample of size two.
+The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
+The denominator, which is proportional to the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.
 
@@
 
-Functions for compute the branch-length moments are available [GITHUB].
+Functions for computing the branch-length moments are available [GITHUB].
 The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
@@
 They are therefore only practical for sample sizes up to $n \approx 40$.
-The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ [CHECK THIS] and is also only practical for samples up to $n \approx 50$.
+The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.
 
@@
 
-Our beta coalescent simulation code is based on modifications to the \texttt{msprime} package made by Joe Zhu [link to github] to allow for multiple mergers.
+We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2014} capable of multiple mergers, based on modifications made by Joe Zhu [link to github].
 This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
 
-[Possibly move some of the description here]
+For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
+We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
+For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
+We then calculated the expectations $\{\E{\tau_i\}}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
+These expectations allow us to calculate all of the 2-SFS statistics defined above.
 
@@
 
-We calculated the average pairwise diversity, $\pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
+We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
 Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
 Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
-Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\pi$ for further analysis.
+Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\Pi$ for further analysis.
 The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
@@
 We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
-We estimated $\E{T_2}$ by $\pi / 2 \mu$.
-For $\pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
+We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
+For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
 For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
@@
 
-\subsection*{Population growth vs. the beta coalescent in non-recombining loci}
+\subsection*{Population growth versus the beta coalescent in non-recombining loci}
 
 In this section, we compare the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
-We wrote a \texttt{python} package to compute the SFS, 2-SFS, and $\fpmi$ numerically.
-This package uses the results of \cite{Fu1995} for the time-homogeneous Kingman coalescent, \cite{ZivkovicWiehe2008} for the time-inhomogeneous Kingman coalescent, and \cite{BirknerEtAl2013} for the beta coalescent.
-(See Methods for details.)
+We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
+To this end, we compute a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
+Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
+All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure1.pdf}
+\includegraphics[width=0.5\textwidth]{figures/wfpmi_growth_beta.pdf}
 \caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$).
@@
 \Fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
-We find that while exponential growth with $g=4$ has a substantial effect on the SFS (Tajima's D = $-0.39 \pi$, where $\pi$ is the average pairwise diversity), it does not qualitatively change the $\fpmi$.
+We find that while exponential growth with $g=4$ has a substantial effect on the SFS ($D = -0.39$), it does not qualitatively change the $\fpmi$.
 In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, just as in the constant-$N$ Kingman coalescent.
 
-On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, Tajima's D = $-0.42 \pi$) also generates positive $\fpmi$.
+On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, $D = -0.42$) also generates positive $\fpmi$.
 The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
@@
 Accordingly, we computed the coarse-grained $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
-For the same parameters, we also calculated a normalized Tajima's D statistic:
-$(\hat{\theta}_W - \pi) / \pi$, where $\hat{\theta}_W$ is Watterson's estimator of the coalescent scaled mutation rate \autocite{Watterson19??}.
-The latter is a measure of the distortions in the SFS induced by population growth or multiple mergers and equals zero for the constant-$N$ Kingman coalescent.
 \Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
@@
 
-\subsection*{Samples of four chromosomes}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/trees.png}
-\caption{Genealogies with a sample size of $n=4$.
-Opportunities for singleton/tripleton mutations are in orange.
-Opportunities for doubletons are in blue.
-(Note: there is a third possible multiple-merger topology, not shown.) \label{fig:trees}}
-\end{figure}
-
-We can get an intuitive understanding for the result above by considering a sample four chromosomes.
-In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
-Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
-As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
-Genealogies with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
-Conditional on observing a doubleton at a site, it is thus more likely that the genealogy has topology (B) and so the expected number of singletons at sites with the same genealogy is lower than average.
-In terms of the 2-SFS, we have $\E{\eta_{12}} < \E{\eta_{1}} \E{\eta_{2}}$.
-
-On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
-For example, topology (C) has less opportunity for singletons \emph{and} less opportunity for doubletons than (A) or (B), even though the expected proportion of singletons is higher.
-Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies higher.
-If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) so that $\E{\eta_{12}} > \E{\eta_{1}} \E{\eta_{2}}$.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
-\caption{Distortions in $\fpmi$ vs.
-distortions in the SFS for $n=4$.
-Parameters for beta coalescents and exponential growth are as in \fig{fig:fpmi_vs_tajimasd}.
-For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^{-4}$ to 2 in coalescent time units.
-\label{fig:sdpmi_vs_sdratio}}
-\end{figure}
-
-As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
-\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
-\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
-In the latter model, we vary both the fold-change in $N$ and the time of the change.
-Like Tajima's D in the previous section, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
-(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
-As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
-Moreover, the two different models of growth collapse onto one another.
-This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
-
 \subsection*{Pointwise mutual information between recombining sites}
@@
 
-To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KehellerEtAl201?} modified to allow for multiple mergers.
-In particular, we simulated a model where coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
+To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KehellerEtAl201?} modified to allow for multiple mergers (See Methods).
+In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
 In this model, marginal genealogies will follow the beta coalescent distribution, and the average SFS will be given by the formula in \cite{BirknerEtAl2013}.
 We also simulated data from two models of population growth: exponential growth and a two-epoch piecewise-constant model.
-
-For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
-We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
-For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
-We then calculated the expectations $\{\E{\tau_i\}}$ and $\{\E{\tau_ij}\}$ by averaging over independent loci.
-These expectations allow us to calculate all of the 2-SFS statistics defined above.
 
@@
 Lines are colored by the distortion in the SFS, as measured by Tajima's D.
-\label{fig:hilo_vs_d}}
+[TODO: SPLIT SWEEPS PANELS OFF INTO THEIR OWN FIGURE.]\label{fig:hilo_vs_d}}
 \end{figure}
@@
 The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
-In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
+In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
 Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
@@
 \section*{Discussion}
+
+\subsection*{Samples of four chromosomes}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/trees.png}
+\caption{Genealogies with a sample size of $n=4$.
+Opportunities for singleton/tripleton mutations are in orange.
+Opportunities for doubletons are in blue.
+(Note: there is a third possible multiple-merger topology, not shown.) \label{fig:trees}}
+\end{figure}
+
+We can get an intuitive understanding for the result above by considering a sample four chromosomes.
+In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
+Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
+As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
+Genealogies with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
+Conditional on observing a doubleton at a site, it is thus more likely that the genealogy has topology (B) and so the expected number of singletons at sites with the same genealogy is lower than average.
+In terms of the 2-SFS, we have $\E{\eta_{12}} < \E{\eta_{1}} \E{\eta_{2}}$.
+
+On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
+For example, topology (C) has less opportunity for singletons \emph{and} less opportunity for doubletons than (A) or (B), even though the expected proportion of singletons is higher.
+Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies higher.
+If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) so that $\E{\eta_{12}} > \E{\eta_{1}} \E{\eta_{2}}$.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
+\caption{Distortions in $\fpmi$ vs.
+distortions in the SFS for $n=4$.
+Parameters for beta coalescents and exponential growth are as in \fig{fig:fpmi_vs_tajimasd}.
+For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^{-4}$ to 2 in coalescent time units.
+\label{fig:sdpmi_vs_sdratio}}
+\end{figure}
+
+As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
+\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
+\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
+In the latter model, we vary both the fold-change in $N$ and the time of the change.
+Like Tajima's D, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
+(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
+As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
+Moreover, the two different models of growth collapse onto one another.
+This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
+
 \begin{itemize}
@@
 \includegraphics[width=0.8\textwidth]{figures/pi_vs_position.pdf}
-\caption{Average pairwise diversity, $\pi$, of 100 haploid samples from DPGP3 panel.
+\caption{Average pairwise diversity, $\Pi$, of 100 haploid samples from DPGP3 panel.
 Each point represents the average over a 10~KB window.

### 2018-08-16 manuscript/manuscript.tex
Finish Results.

@@
 This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
-These positive correlations, however, do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
+These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
 On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
+Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.
 
@@
 As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model then in models of population growth than generate similar distortions in the SFS.
-[IDEA: A plot of $\hilopmi$ at a fixed r vs Tajima's D for all the models together. Could make several plots, varying $r$ and $i_c$. This might be more convincing than the current figure.]
 This is true across recombination rates and high/low cutoff minor allele counts.
 These results demonstrate that $\hilopmi$ is capable of discriminating between coalescent models even when there is recombination between sites.
+
+\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
+While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $r \E{T_2} \sim 1$ apart.
+This holds for exponential as well as piecewise constant growth.
+In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
+This result is robust to the choice of cutoff minor allele count for binning.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD.pdf}
+\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination.
+All values are for $r \E{T_2} \sim 1$ and $n=100$.
+Orange points show Kingman coalescent simulations with population growth.
+Blue points show beta coalescent simulations.
+[Q: Should this be the bottom row of the previous figure?] \label{fig:hilo_vs_tajD}}
+\end{figure}
 
@@
 \begin{enumerate}[label=(\roman*), before=\unskip{: }]
-    \item Fit a demographic model to data using any relevant method.
+    \item Fit a demographic model to data.
     \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
@@
 
-We can get an intuitive understanding for the result above by considering a sample four chromosomes.
+We can get an intuitive understanding for these results by considering a sample four chromosomes.
 In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
@@
 As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
-Moreover, the two different models of growth collapse onto one another.
+Moreover, the results for two different models of growth coincide.
 This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.

### 2018-08-16 manuscript/manuscript.tex
Remove references to "coarse graining"

@@
 Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
+It is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.
 
-We can also compute the pointwise mutual information in this coarse-grained distribution as
+We can also compute the pointwise mutual information in this binned distribution as
 \begin{align}
@@
 One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
-Also, it is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.
 
@@
 The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
-This justifies coarse-graining the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.
+This justifies binning the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.
 
-Accordingly, we computed the coarse-grained $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
+Accordingly, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
 \Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
@@
 This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
-These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
+These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
 On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
@@
 
-Our results above show that $\fpmi$ and its coarse-grained analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
+Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
 This is true even when the population growth generates similar distortions in the SFS.

### 2018-08-16 manuscript/manuscript.tex
Write Discussion.

@@
 
-\subsection*{Samples of four chromosomes}
+We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant under population growth in the Kingman coalescent.
+These properties make them well-suited for coalescent model checking.
+We've demonstrated a model-checking procedure on data from \textit{D.
+melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth can not explain the positive associations between high and low frequency mutations.
 
@@
 
-We can get an intuitive understanding for these results by considering a sample four chromosomes.
+We can get an intuitive understanding for why $\fpmi$ distinguishes among coalescent models by considering a sample four chromosomes.
 In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
@@
 
-\begin{itemize}
-    \item Review results.
-    \item Availability of code: models for computing moments numerically, scripts for running simulations. Notebooks for data analysis.
-    \item We focus here on coalescent-based methods, but also relevant for diffusion methods (because they imply a coalescent model)
-    \item Mention methods to fit multiple mergers coalescent to data (computationally difficult).
-Not really what we're trying to do here.
-Our goal is that you can exploit this extra dimension of the data you already have to check your model and make sure you're not fitting a signal of selection, etc with a demographic model.
-    \item Could also develop a formal hypothesis testing procedure, using the bootstrap, etc.
-    \item Compare with: Single-locus tests that rely on building trees (Seger whale lice, Neher with viruses)
-    \item Compare with: Multi-locus tests that compare a particular region to genomic background. (What if the background isn't neutral/Kingman?)
-    \item But: we could extend our method to a local method that looks for variation in ``multiple-merger-ness''
-    \item Multiple-mergers coalescent and recombination: open area of research
-    \item $\fpmi(d)$ should be a standard posterior check for demographic models. Possibly incorporate into model fitting packages.
-    \item Interesting to compare across species.
-\end{itemize}
+We focus here on demographic inference methods that explicitly use the Kingman coalescent model for calculations.
+However, another popular class of methods are based on forward-time models, such as the Wright-Fisher model (e.g., \cite{GutenkunstEtAl2009}, \cite{dical}).
+We believe that $\fpmi$ is useful for model-checking with these methods as well.
+This is because forward-time neutral models each have an associated dual coalescent model \cite{etheridge}, which can be used to compute the expected $\fpmi$ as outlined in this paper.
+Furthermore, non-neutral models also generate predictions about the 2-SFS and thus $\fpmi$.
+In principle, any fitted model that allows calculation or simulation of the 2-SFS may be checked using $\fpmi$.
+However, the suitability of $\fpmi$ for discriminating among arbitrary families of models is unknown.
+
+In this paper, we have outlined a graphical model-checking procedure in the spirit of \cite{Anscombe} and \cite{GelmanBDA}.
+One could extend our work to implement a formal hypothesis-testing framework for rejecting the best-fit Kingman coalescent model according to some inference method.
+Numerical calculations would require higher moments of the branch-length distribution, which may be computationally intractable for large samples.
+On the other hand, it would be straightforward to develop a bootstrap-style procedure, using coalescent simulations to estimate the variance in test statistics under the fit model.
+In any case, we believe that it would be useful for simulation packages such as \texttt{msprime} and population genetics analysis libraries to include standard functions for computing the 2-SFS, $\fpmi$, and $\hilopmi$.
+
+Our coalescent simulations implement a particular version of the multiple mergers coalescent with recombination.
+However, the precise correspondence between this model and any particular forward-time process is unclear.
+In order to compare with our numerical calculations based on \cite{BirknerEtAl2013}, we use the beta coalescent for the distribution of marginal genealogies, but the explicit forward-time models that have been studied by others generate simultaneous multiple mergers.
+(See e.g., \cite{DurrettSchweinsberg2005} in a model with selective sweeps, but note that our forward-time simulations suggest that this distinction is not important to our main results.)
+Moreover, the long-range correlations observed in \fig{fig:hilo_vs_d} likely depend on an implicit choice regarding the scaling of recombination, coalescent, and multiple merger rates \autocite{EldonWakeley2006}.
+These issues are poorly understood, and more theoretical work is required to understand the interactions between multiple mergers and recombination.
+
+There are two interesting potential empirical applications of our work: assessing the evidence for variation in multiple mergers coalescence within genomes and between species.
+For example, one could compute the $\fpmi$ in different regions of a large genome and look for a relationship between the strength of non-Kingman coalescence and genomic properties such as the recombination rate and functional density.
+Alternatively, one could survey multiple species using a data set such as the diversity data compiled by \cite{Corbett-DetigEtAl2015}.
+Either would reveal new information about the suitability of population genetic models, and the forces that determine genetic diversity.
 
@@
 \section*{Supplementary Methods}
-[IF NECESSARY]
+[MOVE STUFF HERE IF NECESSARY]
 
@@
 
+Left out of Discussion:
+\begin{itemize}
+    \item Mention methods to fit multiple mergers coalescent to data (computationally difficult).
+Not really what we're trying to do here.
+Our goal is that you can exploit this extra dimension of the data you already have to check your model and make sure you're not fitting a signal of selection, etc with a demographic model.
+    \item Compare with: Single-locus tests that rely on building trees (Seger whale lice, Neher with viruses)
+\end{itemize}
+
 Things that were done but not included in this ms:

### 2018-08-16 manuscript/manuscript.tex
Line edits to manuscript.

@@
 \cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
-\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.
+\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.
 
@@
 Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
-In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
+In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
 Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta.
@@
 As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
-However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
+However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$.
 Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.
@@
 
-In our setting, the 2-SFS may be interpreted as a joint probability mass function over allele frequencies.
-Given a coalescent model, the minor allele count at an arbitrary site is a random variable with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
+In our setting, the 2-SFS may be interpreted as a joint probability mass function over minor allele counts.
+Given a coalescent model, the minor allele count at an arbitrary site is a random variable over $\{0,\ldots,\floor{n/2}\}$ with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
 Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
-Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of allele frequencies.
+Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of minor allele counts.
 
@@
 First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
-Moreover, the results of \cite{Fu1995} show that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
+Moreover, the results of \cite{Fu1995} imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
 
@@
 The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
-The denominator, which is proportional to the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.
+The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.
 
@@
 For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
-We then calculated the expectations $\{\E{\tau_i\}}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
+We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
 These expectations allow us to calculate all of the 2-SFS statistics defined above.
@@
 Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
-Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.
+Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.
 
@@
 
-The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
+We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
 The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
@@
 For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
-These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.
+These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.
 
@@
 
-In this section, we compare the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
+We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
 We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
-To this end, we compute a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
@@
 However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
-In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy exactly and calculate tree imbalance statistics directly (see e.g., \cite{Seger2010}).
+In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy explicitly and calculate tree imbalance statistics (see e.g., \cite{Seger2010}).
 The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
@@
 Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
-We do not expect this behavior to be universal to multiple mergers coalescents with recombination.
+We do not expect this behavior to be universal in multiple mergers coalescents with recombination.
 
@@
 \end{enumerate}
-This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
+This procedure checks whether the model is consistent with a dimension the data that was not used in fitting the model.
 Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.
@@
 These properties make them well-suited for coalescent model checking.
-We've demonstrated a model-checking procedure on data from \textit{D.
+We have demonstrated a model-checking procedure on data from \textit{D.
 melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth can not explain the positive associations between high and low frequency mutations.

### 2018-08-16 manuscript/manuscript.tex
Rename some figures.
Rename more figures.

@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/wfpmi_growth_beta.pdf}
+\includegraphics[width=0.5\textwidth]{figures/wfpmi_no_recombination.pdf}
 \caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$).
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure2.pdf}
+\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD.pdf}
 \caption{Pointwise mutual information between singletons and non-singletons for Kingman coalescents with exponential growth and beta coalescents.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure4.pdf}
+\includegraphics[width=0.8\textwidth]{figures/wfpmi_with_recombination.pdf}
 \caption{Weighted frequency pointwise information for three recombination rates with sample size $n=100$.
 As in \Fig{fig:nonrecombining_pmi}, the diagonal elements are masked, and the growth and multiple-merger parameters were chosen to generate average SFS with similar Tajima's D.
-\label{fig:wpmi_recombination}}
+\label{fig:wfpmi_recombination}}
 \end{figure}
 
-\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
+\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
 As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
+\includegraphics[width=0.8\textwidth]{figures/singleton-doubleton.pdf}
 \caption{Distortions in $\fpmi$ vs.

### 2018-08-16 manuscript/manuscript.tex
Fix collision in figure names.
Update To-Do and delete a misplaced figure.

@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD.pdf}
+\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_no_recombination.pdf}
 \caption{Pointwise mutual information between singletons and non-singletons for Kingman coalescents with exponential growth and beta coalescents.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD.pdf}
+\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_recombination.pdf}
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination.
@@
 TODO:
+- Text
     - Abstract
-- Discussion
-- Edit figures
     - Consistency of nomenclature, labeling, etc.
-- Check drosophila numbers. Get citations or remake figure.
-- Check references to Methods. (Should all be there and all have corresponding methods section.)
+    - Insert links to github
     - Get rest of citations
-- Code:
-    - organize
-    - clean
-    - comment
-    - post on Github
+    - Fix bibtex issues
+    - Copy edit text
+    - Copy edit bibliography
+- Figures
+    - Organize panels
+    - Remake figure 5 that got deleted
+    - Rename files
+    - Figure sizes (1-column vs 2-column)
+    - Typeface and font sizes
+    - Color schemes and color maps
+    - Consistency of plotted values (esp. pmi vs cov, base of log scales)
+- Notebooks
+    - Single notebook to make all non-data figures.
+    - Single notebook to do all DPGP3 exploratory and final analysis
+    - Reduce redundancy. Decide which helper functions belong in notebooks and which go in modules.
+    - Remove old/unused notebooks from repo.
+- Simulation output
+    - Figure out which files are necessary to reproduce figures
+    - Make tarball with simulations and include in repo (if reasonable size)
+- DPGP3 data
+    - Decide whether we need to post any processed data.
+- Code
+    - Conda environment and environment file
+    - Clean up snakefiles
+    - Compute 2-epoch moments exactly
+    - Organize modules, reduce redundancy
+    - Refactor functions into a consistent style
+    - Document
+    - Figure out what to do about msprime code
+    - Post on Github
 
 Left out of Discussion:
-\begin{itemize}
-    \item Mention methods to fit multiple mergers coalescent to data (computationally difficult).
+Mention methods to fit multiple mergers coalescent to data (computationally difficult).
 Not really what we're trying to do here.
 Our goal is that you can exploit this extra dimension of the data you already have to check your model and make sure you're not fitting a signal of selection, etc with a demographic model.
-    \item Compare with: Single-locus tests that rely on building trees (Seger whale lice, Neher with viruses)
-\end{itemize}
 

### 2018-08-27 manuscript/manuscript.tex
Minor changes.

@@
 \usepackage[inline]{enumitem}
-\usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
+\usepackage[
+    backend=biber,
+    citestyle=authoryear,
+    style=authoryear,
+    autocite=inline,
+    isbn=false, doi=false, url=false, eprint=false,
+    uniquename=false,
+    maxcitenames=2
+]{biblatex}
 % Fix biblatex behavior that writes "In: " before journal name.
@@
 
-\bibliography{references}
+\addbibresource{references.bib}
 
@@
 \abstract{[NOTE: OLD]
-The genetic diversity of a population reflects its demographic and
-evolutionary history. Methods for inferring this history typically
-assume that the ancestry of a sample can be modeled by the Kingman
-coalescent process. A defining feature of the Kingman coalescent is
-that it generates genealogies that are binary trees: no more than two
-ancestral lineages may coalesce at the same time. However, this
-assumption breaks down under several scenarios. For example, pervasive
-natural selection, rapid spatial range expansion, and extreme
-variation in offspring number can all generate genealogies with
-``multiple-merger'' events in which more than two lineages coalesce
-instantaneously. Therefore, detecting multiple mergers is important
-both for understanding which forces have shaped the diversity of a
-population and for avoiding fitting misspecified models to data.
-Current methods to detect multiple mergers rely on the average site
-frequency spectrum (SFS). However, the signatures of multiple
-mergers in the average SFS are also consistent with a Kingman
-coalescent process with a time-varying population size. Here, I
-present a new method for detecting multiple mergers based on
-the mutual information of the joint allele frequency spectrum at pairs of linked sites. Unlike the average SFS,
-the mutual information depends mostly on the topologies of genealogies
-rather than their branch lengths and is therefore robust to most
-demographic effects.}
+The genetic diversity of a population reflects its demographic and evolutionary history.
+Methods for inferring this history typically assume that the ancestry of a sample can be modeled by the Kingman coalescent process.
+A defining feature of the Kingman coalescent is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
+However, this assumption breaks down under several scenarios.
+For example, pervasive natural selection and extreme variation in offspring number can both generate genealogies with ``multiple-merger'' events in which more than two lineages coalesce instantaneously.
+Therefore, detecting multiple mergers is important both for understanding which forces have shaped the diversity of a population and for avoiding fitting misspecified models to data.
+Current methods to detect multiple mergers rely on the site frequency spectrum (SFS).
+However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent process with a time-varying population size.
+Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
+Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population size change.}
 
@@
 
-Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
 In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
@@
 
-We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
+We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
 The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
@@
 We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
-To this end, we computed a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
@@
 Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
-We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtal2015}.
+We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtAl2015}.
 We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.

### 2018-08-27 manuscript/manuscript.tex
Incorporate JN line edits.

@@
 
-\DeclareMathOperator{\pmi}{pmi}
-\DeclareMathOperator{\fpmi}{fpmi}
-\DeclareMathOperator{\wfpmi}{wfpmi}
-\DeclareMathOperator{\hilopmi}{hilopmi}
+\DeclareMathOperator{\pmi}{PMI}
+\DeclareMathOperator{\fpmi}{fPMI}
+\DeclareMathOperator{\wfpmi}{wfPMI}
+\DeclareMathOperator{\hilopmi}{hiloPMI}
 
@@
 One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
-The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
+The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
 Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
 The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
-Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
+Moreover, the model can be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
 
@@
 A serious problem for this class of inference methods is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
-One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
+For example, one of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
 When this assumption is violated, Kingman-based inference methods are misspecified.
-For example, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
+For instance, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
 If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
@@
 Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
-In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
+In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant to time-varying coalescent rates.
 We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
-These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
+These statistics may be calculated efficiently from genomic single nucleotide variant data.
 Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
@@
 We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.
+(For non-recombining sites, the 2-SFS is independent of the distance and so we will suppress the $d$ in our notation.)
 
-In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
+In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the genealogical branch length distribution by
 \begin{align}
@@
 where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
-(We suppress the distance, $d$, for non-recombining sites.)
 Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.
@@
 
-In information theory, the pointwise mutual information of a pair of random variables (PMI), $X$ and $Y$, is a transformation of the probability mass function, $p(x,y)$, given by
+In information theory, the pointwise mutual information (PMI) of a pair of random variables, $X$ and $Y$, is a transformation of the probability mass function, $p(x,y)$, given by
 \begin{align}
@@
 
-The second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the SFS.
+We claim that the second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
 Population-size variation distorts the SFS by changing the distribution of waiting times.
@@
 We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
-This suggests a binned SFS and 2-SFS:
+This suggests using a binned form of the SFS and 2-SFS:
 \begin{align}
@@
 
-We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$ using the \texttt{python} numerical package \texttt{numpy} \autocite{numpy}.
+We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$.
 For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
-To calculate the second moments of the coalescent time distribution for exponentially growing populations, we used Gaussian quadrature.
 For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.
 
-Functions for computing the branch-length moments are available [GITHUB].
+Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} \autocite{numpy} and available [GITHUB].
 The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
@@
 For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
-We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
+We chose values of $r$ so that the realized values of $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
 For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
@@
 Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
-Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous $\Pi$ for further analysis.
+Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
 The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
@@
 We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
-We fit a three-epoch piecewise constant--$N$ model:
+Following \cite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
 \begin{align}
@@
 We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
-We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
+We are interested in whether $\fpmi$ can distinguish beta from Kingman coalescent models that produce similar distortions in the SFS.
 To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
@@
 
-\Fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
-We find that while exponential growth with $g=4$ has a substantial effect on the SFS ($D = -0.39$), it does not qualitatively change the $\fpmi$.
+As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
+Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS $D = -0.4$.
+However, exponential growth does not qualitatively change the $\fpmi$.
 In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, just as in the constant-$N$ Kingman coalescent.
-
-On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, $D = -0.42$) also generates positive $\fpmi$.
+On the other hand, the beta coalescent generates positive $\fpmi$.
 The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
@@
 
-Accordingly, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
+To generalize this finding, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
 \Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
@@
 However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
-In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy explicitly and calculate tree imbalance statistics (see e.g., \cite{Seger2010}).
 The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
@@
 In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
-This result is robust to the choice of cutoff minor allele count for binning.
+The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.
 
@@
 By varying the recombination rate, mutation rates, population size, and selection coefficient, we were able to vary the rate of selective sweeps linked to neutral sites.
-After the populations reached steady-state, we sampled individuals and calculated $\{\eta_i\}$ and $\{\eta_{ij}(d)\}$ for the neutral mutations.
-(See Methods for details.)
 
@@
 
-\subsection*{Coalescent model checking: application to \textit{Drosophila melanogaster}}
+\subsection*{Application to \textit{Drosophila melanogaster}: Coalescent model checking}
 
@@
 In this section, we illustrate the procedure just outlined by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
-The DPGP3 data consists of haploid consensus sequences from $\sim 200$ wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
+The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
 Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
@@
 \section*{Discussion}
+[DEAL WITH THIS: In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy explicitly and calculate tree imbalance statistics (see e.g., \cite{Seger2010}).
+HERE WE DONT NEED A TREE]
 
@@
 
-We can get an intuitive understanding for why $\fpmi$ distinguishes among coalescent models by considering a sample four chromosomes.
+We can get an intuitive understanding for why $\fpmi$ distinguishes among coalescent models by considering a sample of four chromosomes.
 In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
@@
 \Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
-\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
+\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi(1,2)$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
 In the latter model, we vary both the fold-change in $N$ and the time of the change.
@@
 Moreover, the results for two different models of growth coincide.
-This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
+This suggests that for the Kingman coalescent with growth, $\fpmi(1,2) \approx f(\eta_1/\eta_2)$ for some function $f$.
 
@@
 In any case, we believe that it would be useful for simulation packages such as \texttt{msprime} and population genetics analysis libraries to include standard functions for computing the 2-SFS, $\fpmi$, and $\hilopmi$.
+We provide such functions in the [GITHUB] repository associated with this paper.
 

### 2018-08-28 manuscript/manuscript.tex
Title and author updates. Add ascertainment effects.

@@
 \usepackage{url}
+% Author affiliations
+\usepackage{authblk}
 % Allow inline lists
@@
 ]{biblatex}
+
+% Don't print month and day in bibliography
+\AtEveryBibitem{\clearfield{month}}
+\AtEveryBibitem{\clearfield{day}}
 % Fix biblatex behavior that writes "In: " before journal name.
@@
 
-\title{Distinguishing among coalescent models using two-site allele frequency spectra}
-\author{Daniel P. Rice}
+\title{Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra}
+\author[1]{Daniel P. Rice}
+\author[1,*]{John Novembre}
+\author[2,*]{Michael M. Desai}
+\affil[1]{Department of Human Genetics, University of Chicago, Chicago, IL}
+\affil[2]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
+\affil[*]{These authors contributed equally to this work.}
 \date{\today}
@@
 Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
-In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant to time-varying coalescent rates.
+In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fPMI) is sensitive to multiple mergers but relatively invariant to time-varying coalescent rates.
 We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
@@
 Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
-[OPTIONS: Could add (1) weighted fpmi, and/or (2) binned 2-SFS] \label{fig:schematic}}
+[OPTIONS: Could add (1) weighted fPMI, and/or (2) binned 2-SFS] \label{fig:schematic}}
 \end{figure}
@@
 
+Furthermore, the same normalization renders $\fpmi$ insensitive to ascertainment bias.
+Variant detection methods typically compare sampled sequences to a reference sequence and ``call'' a variant site when there is sufficient evidence that at least one sample differs from the reference at that site.
+As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$.
+This effect distorts the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
+However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$, which cancel.
+Thus, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
 % A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.

### 2018-08-28 manuscript/manuscript.tex
Update abstract.

@@
 
-\abstract{[NOTE: OLD]
-The genetic diversity of a population reflects its demographic and evolutionary history.
-Methods for inferring this history typically assume that the ancestry of a sample can be modeled by the Kingman coalescent process.
+\abstract{
+Population genetics methods for demographic inference typically assume that the ancestry of a sample can be modeled by the Kingman coalescent process.
 A defining feature of the Kingman coalescent is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
@@
 Therefore, detecting multiple mergers is important both for understanding which forces have shaped the diversity of a population and for avoiding fitting misspecified models to data.
-Current methods to detect multiple mergers rely on the site frequency spectrum (SFS).
+Current methods to detect multiple mergers in genomic data rely on the site frequency spectrum (SFS).
 However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent process with a time-varying population size.
 Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
-Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population size change.}
+Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population size change.
+This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
 
@@
 This effect distorts the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
-However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$, which cancel.
-Thus, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
+However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$.
+Because these factors cancel, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
 % A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
@@
 
-We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2014} capable of multiple mergers, based on modifications made by Joe Zhu [link to github].
+We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2014} capable of multiple mergers, based on modifications made by Joe Zhu [GITHUB].
 This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].

### 2018-08-28 manuscript/manuscript.tex
Edit end of introduction.

@@
 Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
-In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fPMI) is sensitive to multiple mergers but relatively invariant to time-varying coalescent rates.
-We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
 These statistics may be calculated efficiently from genomic single nucleotide variant data.
@@
 Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
-We demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
 
@@
 
-In the following, we demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
+In the following, we demonstrate that the positive associations between mutations at different frequencies distinguish the multiple-merger from the Kingman coalescent and that this distinction
 \begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
@@
 \end{enumerate*}.
+To do so, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
+We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
+Finally, we demonstrate the use of $\fpmi$ in a coalescent model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
 

### 2018-08-29 manuscript/manuscript.tex
Make genetic distance notation consistent.

@@
 
-For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
-We chose values of $r$ so that the realized values of $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
+For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d r$.
+(See SI for parameter combinations.)
+We chose values of $d r$ so that the realized values of $d r \E{T_2}$ varied over several orders of magnitude.
 For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
@@
 As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
-In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
+In both models, $\fpmi > 0$ for $i\approx j$ for $d r \E{T_2} \sim 1$.
 This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
 These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
-On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
+On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d r \E{T_2} > 1$.
 Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.
@@
 
-\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d=r\E{T_2}$ and three different choices of $i_c$.
+\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d r \E{T_2}$ and three different choices of $i_c$.
 Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
 At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
-When $r\E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
-In all Kingman models, $\hilopmi$ decays to zero for $r\E{T_2}\gg 1$.
+When $d r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
+In all Kingman models, $\hilopmi$ decays to zero for $d r \E{T_2}\gg 1$.
 
@@
 \Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
-While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $r \E{T_2} \sim 1$ apart.
+While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d r \E{T_2} \sim 1$ apart.
 This holds for exponential as well as piecewise constant growth.
@@
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination.
-All values are for $r \E{T_2} \sim 1$ and $n=100$.
+All values are for $d r \E{T_2} \sim 1$ and $n=100$.
 Orange points show Kingman coalescent simulations with population growth.
@@
 The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
-In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
+In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $d_c r \E{T_2} = 1$.
 Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.

### 2018-08-29 manuscript/manuscript.tex
Add missing citations.

@@
 
-One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
+One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982, Kingman1982a, Hudson1983, Tajima1983}.
 The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
@@
 An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
-In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
-Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982b}.
+In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
+Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
 Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
@@
 recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
-rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
-and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
+rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
+and purifying selection at sufficiently dense sites \autocite{SegerEtAl2010, NicolaisenDesai2012}.
 
@@
 \cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
-\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.
+\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
 
@@
 (Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
-This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
+This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011a} in a model with selection.
 We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.
@@
 This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
-Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
+Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatschek2013}.
 Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
@@
 
-We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2014} capable of multiple mergers, based on modifications made by Joe Zhu [GITHUB].
+We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2016} capable of multiple mergers, based on modifications made by Joe Zhu [GITHUB].
 This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
@@
 
-We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
+We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
 In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
@@
 For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
-For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
+For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightleyEtAl2014}.
 These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.
@@
 We are interested in whether $\fpmi$ can distinguish beta from Kingman coalescent models that produce similar distortions in the SFS.
-To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
@@
 There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
-This is related to the results of \cite{EldonWakeley20??}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
+This is related to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
 They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
@@
 
-Various authors have shown that natural selection can generate multiple merger coalescents at linked neutral sites (e.g., \cite{DurrettSchweinsberg2005, CoopRalph, NeherHallatscheck2013, DesaiEtAl, Seger}).
+Various authors have shown that natural selection can generate multiple merger coalescents at linked neutral sites (e.g., \cite{DurrettSchweinsberg2005, CoopRalph2012, NeherHallatschek2013, DesaiEtAl2013, SegerEtAl2010}).
 However, our simulations of the beta coalescent with recombination are of an explicitly neutral model.
@@
 
-To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
+To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
 In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with fixed selection coefficient $s$.
@@
 Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
-On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{Elyashiv??201?, GarudPetrov2016, others?})
+On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{ElyashivEtAl2016, GarudPetrov2016})
 Thus, this data is a good candidate for demonstrating the utility of $\fpmi$ for assessing coalescent model fit.
@@
 \section*{Discussion}
-[DEAL WITH THIS: In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy explicitly and calculate tree imbalance statistics (see e.g., \cite{Seger2010}).
-HERE WE DONT NEED A TREE]
 
@@
 We focus here on demographic inference methods that explicitly use the Kingman coalescent model for calculations.
-However, another popular class of methods are based on forward-time models, such as the Wright-Fisher model (e.g., \cite{GutenkunstEtAl2009}, \cite{dical}).
+However, another popular class of methods are based on forward-time models, such as the Wright-Fisher model (e.g., \cite{GutenkunstEtAl2009}, \cite{SheehanEtAl2013}).
 We believe that $\fpmi$ is useful for model-checking with these methods as well.
-This is because forward-time neutral models each have an associated dual coalescent model \cite{etheridge}, which can be used to compute the expected $\fpmi$ as outlined in this paper.
+This is because forward-time neutral models each have an associated dual coalescent model \cite{Etheridge2011e}, which can be used to compute the expected $\fpmi$ as outlined in this paper.
 Furthermore, non-neutral models also generate predictions about the 2-SFS and thus $\fpmi$.
@@
 
-In this paper, we have outlined a graphical model-checking procedure in the spirit of \cite{Anscombe} and \cite{GelmanBDA}.
+In this paper, we have outlined a graphical model-checking procedure in the spirit of \cite{Anscombe1973}.
 One could extend our work to implement a formal hypothesis-testing framework for rejecting the best-fit Kingman coalescent model according to some inference method.
@@
 \printbibliography
-
-\section*{Supplementary Methods}
-[MOVE STUFF HERE IF NECESSARY]
 
@@
 
+We also don't mention single locus methods.
+
 Things that were done but not included in this ms:
@@
 OLD TEXT:
+
 % Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).

### 2018-08-30 manuscript/manuscript.tex
Revise figures.

@@
 \abstract{
-Population genetics methods for demographic inference typically assume that the ancestry of a sample can be modeled by the Kingman coalescent process.
-A defining feature of the Kingman coalescent is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
+Demographic inference methods in population genetics typically assume that the ancestry of a sample can be modeled by the Kingman coalescent.
+A defining feature of this stochstic process is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
 However, this assumption breaks down under several scenarios.
@@
 Current methods to detect multiple mergers in genomic data rely on the site frequency spectrum (SFS).
-However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent process with a time-varying population size.
+However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent with a time-varying population size.
 Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
-Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population size change.
+Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population -ize change.
 This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/schematic.pdf}
+\includegraphics[scale=1]{figures/schematic.pdf}
 \caption{Summarizing population genetic data with frequency pointwise information.
@@
 Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
-[OPTIONS: Could add (1) weighted fPMI, and/or (2) binned 2-SFS] \label{fig:schematic}}
+ \label{fig:schematic}}
 \end{figure}
@@
 Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
-That is, we expect $\fpmi$ to be sensitive to multiple mergers, but invariant under changes in the coalescent rate.
+That is, we expect $\fpmi$ to be sensitive to multiple mergers, but invariant to changes in the coalescent rate.
 
@@
 The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
-The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.
+The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the overall mutation and coalescence rates.
 
@@
 We are interested in whether $\fpmi$ can distinguish beta from Kingman coalescent models that produce similar distortions in the SFS.
-To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant to changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/wfpmi_no_recombination.pdf}
+\includegraphics[scale=1]{figures/wfpmi_no_recombination.pdf}
 \caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$).
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_no_recombination.pdf}
+\includegraphics[scale=1]{figures/hilopmi_vs_tajimasD_no_recombination.pdf}
 \caption{Pointwise mutual information between singletons and non-singletons for Kingman coalescents with exponential growth and beta coalescents.
@@
 
-To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KehellerEtAl201?} modified to allow for multiple mergers (See Methods).
+To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
 In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/wfpmi_with_recombination.pdf}
+\includegraphics[scale=1]{figures/wfpmi_with_recombination.pdf}
 \caption{Weighted frequency pointwise information for three recombination rates with sample size $n=100$.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure5.pdf}
+\includegraphics[scale=1]{figures/hilopmi_vs_distance.pdf}
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus genetic distance.
@@
 Lines are colored by the distortion in the SFS, as measured by Tajima's D.
-[TODO: SPLIT SWEEPS PANELS OFF INTO THEIR OWN FIGURE.]\label{fig:hilo_vs_d}}
+\label{fig:hilo_vs_d}}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD_recombination.pdf}
+\includegraphics[scale=1]{figures/hilopmi_vs_tajimasD_recombination.pdf}
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination.
@@
 Blue points show beta coalescent simulations.
-[Q: Should this be the bottom row of the previous figure?] \label{fig:hilo_vs_tajD}}
+\label{fig:hilo_vs_tajD}}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/figure6.pdf}
+\includegraphics[scale=1]{figures/dpgp3_analysis.pdf}
 \caption{DPGP3 data and coalescent model predictions.
@@
 Dotted lines show the expectation under the constant-$N$ Kingman coalescent.
-[Middle row needs colorbar] \label{fig:dpgp3}}
+\label{fig:dpgp3}}
 \end{figure}
@@
 
-We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant under population growth in the Kingman coalescent.
+We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant to population growth in the Kingman coalescent.
 These properties make them well-suited for coalescent model checking.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/trees.png}
+\includegraphics[width=3.5in]{figures/trees.png}
 \caption{Genealogies with a sample size of $n=4$.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/singleton-doubleton.pdf}
+\includegraphics[scale=1]{figures/singleton-doubleton.pdf}
 \caption{Distortions in $\fpmi$ vs.
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
 \centering
-\includegraphics[width=0.8\textwidth]{figures/pi_vs_position.pdf}
+\includegraphics[scale=1]{figures/pi_vs_position.pdf}
 \caption{Average pairwise diversity, $\Pi$, of 100 haploid samples from DPGP3 panel.
@@
 
-\begin{table}[h!]
+\begin{table}[ht!]
   \begin{center}

### 2018-10-05 manuscript/manuscript.tex
Ivana and Michael suggestions (part 1).

@@
 For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
-Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygocity along a chromosome (e.g., \cite{LiDurbin2011}).
+Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygosity along a chromosome (e.g., \cite{LiDurbin2011}).
 
@@
 For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
-In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
+In a similar vein, \cite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
 Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
@@
 rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
-and purifying selection at sufficiently dense sites \autocite{SegerEtAl2010, NicolaisenDesai2012}.
+and purifying selection at sufficiently many sites \autocite{SegerEtAl2010, NicolaisenDesai2012, GoodEtAl2014}.
 
@@
 This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
-Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatschek2013}.
-Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
-As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
-However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$.
+Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
+The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
 Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.
@@
 Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
-That is, we expect $\fpmi$ to be sensitive to multiple mergers, but invariant to changes in the coalescent rate.
+That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.
 
@@
 
-For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d r$.
+For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
 (See SI for parameter combinations.)
-We chose values of $d r$ so that the realized values of $d r \E{T_2}$ varied over several orders of magnitude.
+We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
 For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
@@
 In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
-Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
+Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and per basepair mutation rate $\mu = 10^{-7}$.
 We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
@@
 
-The previous sections have shown that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
+We have shown in the previous sections that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
 However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
@@
 As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
-In both models, $\fpmi > 0$ for $i\approx j$ for $d r \E{T_2} \sim 1$.
+In both models, $\fpmi > 0$ for $i\approx j$ for $d\,r \E{T_2} \sim 1$.
 This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
 These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
-On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d r \E{T_2} > 1$.
+On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d\,r \E{T_2} > 1$.
 Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.
@@
 
-\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d r \E{T_2}$ and three different choices of $i_c$.
+\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d\,r \E{T_2}$ and three different choices of $i_c$.
 Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
 At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
-When $d r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
-In all Kingman models, $\hilopmi$ decays to zero for $d r \E{T_2}\gg 1$.
+When $d\,r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
+In all Kingman models, $\hilopmi$ decays to zero for $d\,r \E{T_2}\gg 1$.
 
@@
 \Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
-While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d r \E{T_2} \sim 1$ apart.
+While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d\,r \E{T_2} \sim 1$ apart.
 This holds for exponential as well as piecewise constant growth.
@@
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination.
-All values are for $d r \E{T_2} \sim 1$ and $n=100$.
+All values are for $d\,r \E{T_2} \sim 1$ and $n=100$.
 Orange points show Kingman coalescent simulations with population growth.
@@
 The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
-For intermediate genetic distances, $d r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
+For intermediate genetic distances, $d\,r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
 That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
-The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d r \E{T_2} \gg 1$.
+The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
 This is reasonable because the effect of a particular sweep on genealogies should be localized around the beneficial mutation.
@@
 
+\subsection*{Acknowledgements}
+TODO
+
 \printbibliography

### 2018-10-06 manuscript/manuscript.tex
Ivana and Michael suggestions (part 2).

@@
 The fundamental tools of the trade are simplified mathematical models, which connect unobserved quantities such as the population size to observable features of genetic data.
-However, populations are complicated and, moreover, varied in their complications.
-No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
+However, populations are complicated and, moreover, vary in their complications.
+A simple model cannot capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
 It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.
@@
 The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
-The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the overall mutation and coalescence rates.
+The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the mutation rate and average pairwise coalescence time.
 
@@
 
-Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} \autocite{numpy} and available [GITHUB].
+Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
 The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
@@
 We are interested in whether $\fpmi$ can distinguish beta from Kingman coalescent models that produce similar distortions in the SFS.
-To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant to changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant to changes in the average pairwise coalescence time: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
@@
 \caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus genetic distance.
+Each curve represents a different combination of coalescent parameters.
 The first three rows show coalescent simulations in \texttt{msprime}.
-Exponential growth rate $g$ ranges from 0.01 to 8.0 in coalescent time units.
+Exponential growth rate $g$ varies logarithmically from 0.01 to 8.0 in coalescent time units.
 Piecewise constant change-points, range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2.
@@
 In contrast, the beta coalescent generates $\hilopmi$ that is consistently greater than the constant-$N$ Kingman.
-As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model then in models of population growth than generate similar distortions in the SFS.
+As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model than in models of population growth than generate similar distortions in the SFS.
 This is true across recombination rates and high/low cutoff minor allele counts.
+
 These results demonstrate that $\hilopmi$ is capable of discriminating between coalescent models even when there is recombination between sites.
+In particular, population growth has very little effect on $\hilopmi$, especially for $d\,r \E{T_2} \gtrsim 1$, while multiple mergers increase $\hilopmi$ across the range of $d\,r \E{T_2}$.
+\Fig{fig:hilo_vs_d} also suggests that the most informative genomic distance for distinguishing multiple mergers from population growth is $d\,r \E{T_2} \sim 1$.
+In a later section, we will demonstrate how to use these results to implement a graphical model checking procedure based on plotting $\hilopmi$ as a function of genomic distance.
 
-\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
+\Fig{fig:hilo_vs_tajD} highlights the relative invariance of $\hilopmi$ at intermediate genomic distances to population growth.
 While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d\,r \E{T_2} \sim 1$ apart.
@@
 In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
-The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.
+The figure also shows that this result is robust to the choice of cutoff minor allele count for binning.
 
@@
 The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
-This is reasonable because the effect of a particular sweep on genealogies should be localized around the beneficial mutation.
+This decay is reasonable because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.
 
@@
 In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
-Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
+Furthermore, the total branch length is independent of the topology \autocite{Wakeley2009}.
 As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
@@
 As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
-Moreover, the results for two different models of growth coincide.
-This suggests that for the Kingman coalescent with growth, $\fpmi(1,2) \approx f(\eta_1/\eta_2)$ for some function $f$.
+Moreover, the results for two different models of growth coincide, suggesting that the functional form of the population growth does not strongly influence $\fpmi$.
 
@@
 We believe that $\fpmi$ is useful for model-checking with these methods as well.
-This is because forward-time neutral models each have an associated dual coalescent model \cite{Etheridge2011e}, which can be used to compute the expected $\fpmi$ as outlined in this paper.
+This is because forward-time neutral models each have an associated dual coalescent model \autocite{Etheridge2011e}, which can be used to compute the expected $\fpmi$ as outlined in this paper.
 Furthermore, non-neutral models also generate predictions about the 2-SFS and thus $\fpmi$.

### 2018-10-06 manuscript/manuscript.tex
Maryn suggestions (part 1).

@@
 However, populations are complicated and, moreover, vary in their complications.
-A simple model cannot capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
-It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.
+A simple model cannot capture the processes governing every species' evolution, and will generate misleading inferences if misspecified.
+It is therefore crucial to understand the limits of our models and to assess when a model is appropriate for a particular data set.
 
@@
 Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
-The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
+The Kingman coalescent has a number of convenient properties that facilitate both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
 Moreover, the model can be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).
@@
 In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
-Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
+Under many conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
 Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
@@
 If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
-In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly} rather than inversely proportional to the population size.
+In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly}, rather than inversely, proportional to the population size.
 It follows that the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.
@@
 
-An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
+An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} (\cite{Pitman1999, Sagitov1999, DonnellyKurtz1999,}; reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
 Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
-The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
-These models are relevant for species with
+The more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
+Multiple merger and simultaneous multiple merger models are relevant for species with
 ``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
@@
 In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
-For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
+For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by the rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
 In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.
@@
 This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{VittiEtAl2013}).
-These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
+Selection scan methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
 Genomic regions that are outliers from this distribution are presumed to be under selection.
@@
 One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
-\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
+To this end, \cite{BirknerEtAl2013, BlathEtAl2016} and \cite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
 \cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
-\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
+\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
 
-However, existing methods have limitations for distinguishing multiple mergers from general models of population-size change.
+However, existing methods are limited in their ability to distinguish multiple mergers from general models of population-size change.
 The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
 \cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
-The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but detecting it in data requires knowledge of the ancestral allele at each site.
-High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS at high frequencies.
+The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowledge of the ancestral allele at each site.
+High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.
 
@@
 %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
-The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
+The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowledge of the physical locations of the SNPs.
 
@@
 (Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
-This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011a} in a model with selection.
+The 2-SFS has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and by \cite{Xie2011a} in a model with selection.
 We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.
@@
 
-In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the genealogical branch length distribution by
+In the limit of low per-site mutation rate ($\mu\to0$) and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the genealogical branch length distribution by
 \begin{align}
@@
 \cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
-This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
+This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
 Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
@@
 \end{enumerate*}.
-To do so, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
+To demonstrate these four properties, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
 We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
@@
 
-We are interested in quantifying the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
+We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
 To do so, we define the \textit{frequency pointwise mutual information} as
@@
 \includegraphics[scale=1]{figures/schematic.pdf}
-\caption{Summarizing population genetic data with frequency pointwise information.
+\caption{Summarizing population genetic data with frequency pointwise information ($\fpmi$).
 First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations.
@@
 \end{align}
-where $p(x) = \sum_y p(x,y)$ and vice versa \autocite{ChurchHanks1990}.
+where $p(x) = \sum_y p(x,y)$ and similarly for $p(y)$ \autocite{ChurchHanks1990}.
 PMI measures the change in the probability that $X=x$ given knowledge that $Y=y$.
 When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
-On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
+When $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and $\pmi(x,y) < 0$ implies the converse.
 (The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.)
 
-In our setting, the 2-SFS may be interpreted as a joint probability mass function over minor allele counts.
+We can interpret the 2-SFS as a joint probability mass function over minor allele counts.
 Given a coalescent model, the minor allele count at an arbitrary site is a random variable over $\{0,\ldots,\floor{n/2}\}$ with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
-Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
+Similarly, the minor allele counts at two sites separated by $d$ bases are random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
 Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of minor allele counts.
@@
 
-We claim that the second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
+The primary benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
 Population-size variation distorts the SFS by changing the distribution of waiting times.
-However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
-On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
+However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous coalescent.
+In contrast, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
 These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.

### 2018-10-07 manuscript/manuscript.tex
Maryn suggestions (part 2).

@@
 This transformation of the 2-SFS has several useful properties.
-First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
-Moreover, the results of \cite{Fu1995} imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
-
-The primary benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
-Population-size variation distorts the SFS by changing the distribution of waiting times.
-However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous coalescent.
-In contrast, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
-These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
-Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
-That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.
+First, because $\fpmi$ is based on the minor allele frequencies, we may compute it without knowing the ancestral allele.
+Second, the results of \cite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
+Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
+In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
+We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is relatively insensitive to the distribution of coalescence times.
+Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.
 
@@
 Variant detection methods typically compare sampled sequences to a reference sequence and ``call'' a variant site when there is sufficient evidence that at least one sample differs from the reference at that site.
-As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$.
-This effect distorts the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
+As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$, skewing the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
 However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$.
@@
 We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
-This suggests using a binned form of the SFS and 2-SFS:
+Thus, to detect multiple mergers, we use a binned form of the SFS and 2-SFS:
 \begin{align}
@@
 where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
-Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
-It is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.
+Binning provides a stable estimate of the 2-SFS for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
+We note that a less coarse-grained binning scheme may strike a balance between sampling noise and preserving more detailed information about allele frequencies.
 
-We can also compute the pointwise mutual information in this binned distribution as
+We compute the pointwise mutual information of the binned distribution as
 \begin{align}
@@
 \end{align}
-One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
+While we will focus on $\hilopmi$, one could similarly calculate five other pointwise mutual information statistics from the binned 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts).
 
 \subsection*{Weighted $\fpmi$}
-For plotting purposes, we will also use a weighted version of the $\fpmi$:
+For plotting purposes, we use a weighted version of the $\fpmi$:
 \begin{align}
@@
 where $T_2$ is the coalescence time for a sample of size two.
-The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
+The numerator of the weighting factor in \eq{eq:wfpmi} emphasizes the most common pairs of minor allele counts.
 The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the mutation rate and average pairwise coalescence time.
@@
 Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
-The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
-The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
-They are therefore only practical for sample sizes up to $n \approx 40$.
+Our Kingman coalescent code can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
+The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \cite{GriffithsTavare1994}.
+Thus, these formulas are only practical for sample sizes up to $n \approx 40$.
 The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.
@@
 We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
-For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
+For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at the ends of the locus.
 We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
@@
 We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
-The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
+The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \cite{LackEtAl2015}.
 We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
@@
 We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
-For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
-(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)
+For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by selecting the first 100 remaining samples in alphanumeric order by sample name.
+As a result, the data for each chromosome arm is from a slightly different subset of the individuals.
 
@@
 Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
-Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
+Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
 Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
@@
 
-In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites where fewer than 90 of the 100 samples have genotype calls.
-This leaves over 90\% of all sites and does not substantially alter the fraction of sites that are polymorphic in the sample (Table~\ref{tab:called_sites})
+In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites with fewer than 90 of the 100 genotypes called.
+This leaves over 90\% of all sites and without substantially altering the fraction of polymorphic sites (Table~\ref{tab:called_sites}).
 
@@
 \end{align}
-estimating both change-points ($t_1$ < $t_2$) and both population sizes ($N_1$, $N_2$).
+estimating both change-points ($t_1 < t_2$) and population sizes ($N_1$, $N_2$).
 We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
@@
 In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
-Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
+Because we are using fourfold degenerate sites, we only computed the 2-SFS for distances that are multiples of three.
 For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
@@
 Growth and $\alpha$ parameters were chosen to generate average SFS with similar Tajima's D.
-The diagonal $i=j$ is masked.
+The diagonal, $i=j$, is masked.
 \label{fig:nonrecombining_pmi}}
@@
 As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
-Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS $D = -0.4$.
+Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS, $D = -0.4$.
 However, exponential growth does not qualitatively change the $\fpmi$.
-In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, just as in the constant-$N$ Kingman coalescent.
+In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, as in the constant-$N$ Kingman coalescent.
 On the other hand, the beta coalescent generates positive $\fpmi$.
@@
 On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
-Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.
+Thus, the 2-SFS, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.
 
@@
 However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
-The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
 Therefore, it is important to assess the robustness of our approach to recombination between sites.
 
-To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
+To measure the effect of recombination on $\fpmi$, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
 In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
@@
 Lines are colored by the distortion in the SFS, as measured by Tajima's D.
+The dashed line shows the constant-$N$ Kingman coalescent.
 \label{fig:hilo_vs_d}}
@@
 \Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d\,r \E{T_2}$ and three different choices of $i_c$.
-Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
+Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS (Tajima's D), which is independent of the recombination rate.
 At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
@@
 There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
-This is related to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
-They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
+This behavior relates to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
+Further, \cite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
 Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
@@
 To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
-In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with fixed selection coefficient $s$.
-Both types of mutations occurred at random, uniformly distributed across the length of the chromosome.
-By varying the recombination rate, mutation rates, population size, and selection coefficient, we were able to vary the rate of selective sweeps linked to neutral sites.
+In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with a fixed selection coefficient, $s$.
+Both types of mutations occurred at random, uniformly distributed along the chromosome.
+By varying the recombination rate, neutral and beneficial mutation rates, population size, and selection coefficient, we varied the rate of selective sweeps linked to neutral sites.
 
@@
 The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
-This decay is reasonable because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.
+This decay is expected because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.
 
@@
 
-Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
-This is true even when the population growth generates similar distortions in the SFS.
+Our results above show that $\fpmi$ and its binned analog, $\hilopmi$, are useful for distinguishing population growth from the effects of multiple mergers, even when population growth generates similar distortions in the SFS.
 We therefore propose the following model-checking procedure for demographic inference methods
@@
 \end{enumerate}
-This procedure checks whether the model is consistent with a dimension the data that was not used in fitting the model.
+This procedure checks whether the demographic model is consistent with a feature of the data that was not used in fitting the model.
 Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.
 
-In this section, we illustrate the procedure just outlined by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
+In this section, we illustrate the procedure outlined above by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
 The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
 Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
-On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{ElyashivEtAl2016, GarudPetrov2016})
+On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{ElyashivEtAl2016, GarudPetrov2016}).
 Thus, this data is a good candidate for demonstrating the utility of $\fpmi$ for assessing coalescent model fit.
 
-After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
+After filtering for missing genotypes, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome characterized by consistent high diversity (Methods).
 Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
-We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtAl2015}.
+We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
 We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
 We report our fitted parameters in Table~\ref{tab:DPGP_params}.
-We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}, See methods for details.).
+We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}, Methods).
 
@@
 In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $d_c r \E{T_2} = 1$.
-Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
+Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except adjacent to the diagonal.
 This pattern extends across a range of genomic distances, $d/d_c \in (10^{-1}, 10^2)$ (\fig{fig:dpgp3}, third row).
@@
 
-Note that the $\hilopmi$ decays toward zero at large distances.
-This matches the expectation from simulations with selective sweeps rather than the beta coalescent.
+Note that the $\hilopmi$ decays toward zero at large distances, matching the expectation from simulations with selective sweeps rather than the beta coalescent.
 However, we caution against concluding that sweeps are necessarily responsible for the deviations from the Kingman expectation.
@@
 These properties make them well-suited for coalescent model checking.
-We have demonstrated a model-checking procedure on data from \textit{D.
-melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth can not explain the positive associations between high and low frequency mutations.
+We demonstrated a model-checking procedure on data from \textit{D.
+melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth alone can not explain the positive associations between high and low frequency mutations.
 
@@
 (In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
-As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
+As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), multiple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
 Moreover, the results for two different models of growth coincide, suggesting that the functional form of the population growth does not strongly influence $\fpmi$.
@@
 In order to compare with our numerical calculations based on \cite{BirknerEtAl2013}, we use the beta coalescent for the distribution of marginal genealogies, but the explicit forward-time models that have been studied by others generate simultaneous multiple mergers.
-(See e.g., \cite{DurrettSchweinsberg2005} in a model with selective sweeps, but note that our forward-time simulations suggest that this distinction is not important to our main results.)
+(See e.g., the \cite{DurrettSchweinsberg2005} model of selective sweeps, but note that our forward-time simulations suggest that this distinction is not important to our main results.)
 Moreover, the long-range correlations observed in \fig{fig:hilo_vs_d} likely depend on an implicit choice regarding the scaling of recombination, coalescent, and multiple merger rates \autocite{EldonWakeley2006}.

### 2018-10-07 manuscript/manuscript.tex
Joe suggestions.

@@
 A simple model cannot capture the processes governing every species' evolution, and will generate misleading inferences if misspecified.
-It is therefore crucial to understand the limits of our models and to assess when a model is appropriate for a particular data set.
+It is therefore crucial to understand the limits of population genetics models and to assess when a model is appropriate for a particular data set.
 
@@
 We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
-Finally, we demonstrate the use of $\fpmi$ in a coalescent model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
+Finally, we demonstrate the use of $\fpmi$ in a graphical model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
 
@@
 Variant detection methods typically compare sampled sequences to a reference sequence and ``call'' a variant site when there is sufficient evidence that at least one sample differs from the reference at that site.
-As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$, skewing the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
-However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$.
+As a result, the ascertainment probability, $q_i$, of a mutation is an increasing function of its true allele count $i$, skewing the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} q_i$) toward high-frequency mutations.
+However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $q_i q_j$.
 Because these factors cancel, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
@@
 Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
-The boundaries of these central regions are given in Table~\ref{tab:central_regions}.
+The boundary positions of these central regions are given in Table~\ref{tab:central_regions}.
 
@@
 
-In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
+In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances between 3~bp and 5~Kb.
 Because we are using fourfold degenerate sites, we only computed the 2-SFS for distances that are multiples of three.

### 2018-10-07 manuscript/manuscript.tex
Evan suggestions.

@@
 In contrast, the beta coalescent generates $\hilopmi$ that is consistently greater than the constant-$N$ Kingman.
-As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model than in models of population growth than generate similar distortions in the SFS.
+As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model than in models of population growth that generate similar distortions in the SFS.
 This is true across recombination rates and high/low cutoff minor allele counts.

### 2018-10-08 manuscript/manuscript.tex
Polish manuscript.

@@
 Demographic inference methods in population genetics typically assume that the ancestry of a sample can be modeled by the Kingman coalescent.
-A defining feature of this stochstic process is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
+A defining feature of this stochastic process is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
 However, this assumption breaks down under several scenarios.
@@
 Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
-Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population -ize change.
+Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than on their branch lengths and is therefore largely insensitive to population size change.
 This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
@@
 However, populations are complicated and, moreover, vary in their complications.
-A simple model cannot capture the processes governing every species' evolution, and will generate misleading inferences if misspecified.
+No simple model can capture the processes governing every species' evolution, and a misspecified model will generate misleading inferences.
 It is therefore crucial to understand the limits of population genetics models and to assess when a model is appropriate for a particular data set.
@@
 The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
-\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
-The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowledge of the ancestral allele at each site.
+\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
+The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
 High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.
@@
 %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
-The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowledge of the physical locations of the SNPs.
+The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowing the physical locations of the SNPs.
 
@@
 When $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and $\pmi(x,y) < 0$ implies the converse.
-(The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.)
+The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.
 
@@
 In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
-We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is relatively insensitive to the distribution of coalescence times.
+We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is largely insensitive to the distribution of coalescence times.
 Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.
@@
 where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
-Binning provides a stable estimate of the 2-SFS for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
+Binning provides a stable estimate of the 2-SFS for large sample sizes because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
 We note that a less coarse-grained binning scheme may strike a balance between sampling noise and preserving more detailed information about allele frequencies.
@@
 
-We have shown in the previous sections that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
+We have shown in the previous section that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
 However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
@@
 In this section, we illustrate the procedure outlined above by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
-The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
+The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be mostly free of cosmopolitan admixture.
 Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
@@
 
-We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant to population growth in the Kingman coalescent.
+We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but largely invariant to population growth in the Kingman coalescent.
 These properties make them well-suited for coalescent model checking.
@@
 In any case, we believe that it would be useful for simulation packages such as \texttt{msprime} and population genetics analysis libraries to include standard functions for computing the 2-SFS, $\fpmi$, and $\hilopmi$.
-We provide such functions in the [GITHUB] repository associated with this paper.
+We provide such functions in the GitHub repository associated with this paper.
 
@@
 \subsection*{Acknowledgements}
-TODO
+DPR was supported by the Chicago Fellows Program of the University of Chicago.
+[OTHER FUNDING].
+We thank
+Arjun Biddanda,
+Maryn Carlson,
+Ivana Cvijovi\'c,
+Ben Good,
+Dick Hudson,
+Evan Koch,
+Joe Marcus,
+Richard Neher,
+Matthias Steinr\"ucken,
+and Aleksandra Walczak for helpful discussions and comments on the manuscript.
 

### 2018-10-08 manuscript/manuscript.tex
Update acknowledgements.

@@
 [OTHER FUNDING].
+This work was completed in part with resources provided by the University of Chicago Research Computing Center and Harvard Faculty of Arts and Sciences Research Computing Center.
 We thank
@@
 Matthias Steinr\"ucken,
+John Wakeley,
 and Aleksandra Walczak for helpful discussions and comments on the manuscript.
+
+\subsection*{Competing interests}
+The authors have no competing interests.
 

### 2018-10-31 src/alignment2maf.py
Remove old source files.

@@ -1,27 +1,0 @@
-import sys
-import numpy as np
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument("nSamples", type=int, help="number of samples to subsample to. Must be less than or equal to the number of samples in the file")
-args = parser.parse_args()
-
-alleles = np.zeros(args.nSamples, dtype='S1')
-for i_line, line in enumerate(sys.stdin):
-    data = list(line.strip()[:args.nSamples])
-    # Masked site or site with uncalled genotypes
-    if 'N' in data:
-        maf = 'NaN'
-    # Homozygous site
-    elif len(data) == 1:
-        maf = '0'
-    else:
-        alleles[:] = data
-        bases, counts = np.unique(alleles, return_counts=True)
-        # Multi-allelic site
-        if len(counts) > 2:
-            maf = 'NaN'
-        # Polymorphic site
-        else:
-            maf = str(np.min(counts)%args.nSamples)
-    sys.stdout.write(maf + '\n')

### 2018-10-31 src/compute_beta_coalescent_moments.py

@@ -1,30 +1,0 @@
-import numpy as np
-import coalescentmoments as cm
-import argparse
-import sys
-
-parser = argparse.ArgumentParser()
-parser.add_argument("nSamples", type=int, help="number of samples to simulate")
-parser.add_argument("alpha", type=float, help='beta coalescent parameter alpha')
-parser.add_argument('--folded_sfs', dest='folded', action='store_true',
-        help='Record folded site frequency spectrum (DEFAULT=True)')
-parser.add_argument('--unfolded_sfs', dest='folded', action='store_false',
-        help='Record unfolded site frequency spectrum (DEFAULT=False)')
-parser.set_defaults(folded=True)
-args = parser.parse_args()
-
-
-M1, M2 = cm.sfs_moments(args.nSamples, alpha=args.alpha)
-
-# Print in a format compatible with other scripts
-print('#N_SAMPLES={}'.format(args.nSamples))
-print('#ALPHA={}'.format(args.alpha))
-print('#SFS_FOLDED={}'.format(args.folded))
-if args.folded:
-    M1_folded, M2_folded = cm.fold_sfs_moments(args.nSamples, M1, M2)
-    print(' '.join([str(x) for x in M1_folded]))
-    # Print second moment matrix in condensed form (only the upper triangular portion)
-    print(' '.join([str(x) for x in M2_folded[np.triu_indices(M2_folded.shape[0])]]))
-else:
-    print(' '.join([str(x) for x in M1]))
-    print(' '.join([str(x) for x in M2[np.triu_indices(M2.shape[0])]]))

### 2018-10-31 src/denoise.py

@@ -1,26 +1,0 @@
-import numpy as np
-import pywt
-
-def anscombe_transform(X):
-        return 2.0*np.sqrt(X)
-
-def anscombe_transform_inverse(X):
-        return (X/2.0)**2
-
-def f_thresh(x, t):
-    return np.sign(x)*np.maximum(0, np.abs(x) - t)
-
-def soft_threshold(wavelet_coeffs, t=1.0):
-    return [wavelet_coeffs[0]] + [f_thresh(wc,t) for wc in wavelet_coeffs[1:]]
-
-def denoise(X, wavelet='db1', extension_mode='constant', t=1.0, anscombe=True):
-    if anscombe:
-        X = anscombe_transform(X)
-    wc = pywt.wavedec(X, wavelet, extension_mode)
-    wc_denoised = soft_threshold(wc, t=t)
-    inv = pywt.waverec(wc_denoised, wavelet, extension_mode)
-    if len(inv) > len(X):
-        inv = inv[:-(len(inv)-len(X))]
-    if anscombe:
-        inv = anscombe_transform_inverse(inv)
-    return inv

### 2018-10-31 src/filter_missing_data.py

@@ -1,32 +1,0 @@
-import sys
-import numpy as np
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument("nSamples", type=int, help="number of samples in file")
-parser.add_argument("maxN", type=int, help='Maximum number of non-genotyped samples allowed to include a site')
-parser.add_argument("-p", "--positions_in_file", action="store_true", help="input file contains position numbers")
-parser.add_argument("-a", "--maxAlleles", type=int, default=2, help="maximum number of alleles allowed (default=2)")
-args = parser.parse_args()
-
-alleles = np.zeros(args.nSamples, dtype='S1')
-for i_line, line in enumerate(sys.stdin):
-    if args.positions_in_file:
-        data = list(line.split()[1])
-        outline = line
-    else:
-        data = list(line.strip())
-        outline = str(i_line+1) + '\t' + line
-
-    # Skip masked sites and sites with more than maxN ungenotyped samples
-    if data == ['N'] or data.count('N') > args.maxN:
-        continue
-    # Always write monomorphic sites
-    elif len(data) == 1:
-        sys.stdout.write(outline)
-    # Make sure there are no more than maxAlleles alleles at the site
-    else:
-        alleles[:] = data
-        n_alleles = np.unique(alleles[alleles != 'N']).size
-        if n_alleles <= args.maxAlleles:
-            sys.stdout.write(outline)

### 2018-10-31 src/msprime_like_slim.py

@@ -1,38 +1,0 @@
-import sys
-import numpy as np
-import argparse
-
-#FIXME: this is a hack to use my local version of msprime
-# sys.path.insert(1, '/users/danielrice/msprime-lambda/')
-# sys.path.insert(1, '/home/dpr/mmc_genomics/src/msprime-lambda/')
-sys.path.insert(1, '/Users/dpr/mmc_genomics/src/msprime-lambda/')
-import msprime
-# import jsfs
-
-# FIXME: take these from a config file
-n_samples = 100
-N = 500
-L = int(1e8)
-mu_n = 1e-7
-r = 1e-8
-
-# Run simulation
-treeseq = msprime.simulate(
-        sample_size=n_samples,
-        length=L,
-        recombination_rate=r,
-        Ne=N,
-        mutation_rate=mu_n)
-
-# Strings to format the output like SLiM
-header_str = '#OUT: {} . . {}\nMutations:\n'
-line_str = '. . m0 {} . . . . {}\n'
-sys.stdout.write(header_str.format(N, n_samples))
-# Loop over trees in the tree sequence, getting the positions and allele counts
-for tree in treeseq.trees():
-    for mutation in tree.mutations():
-        position = int(np.floor(mutation.position))
-        allele_count = tree.get_num_leaves(mutation.node)
-        sys.stdout.write(line_str.format(position, allele_count))
-# This is so the parsing script knows to stop
-sys.stdout.write('Genomes:\n')

### 2018-10-31 src/twosfstheory.py

@@ -1,138 +1,0 @@
-import numpy as np
-import fu
-from scipy.special import binom, expi, loggamma
-
-class TwoSFS():
-    '''
-    Calculate the two-site frequency spectrum for perfectly-linked sites.
-    Uses formulas from Fu (1995), Eriksson et al. (2010), and Rafajlovic et al. (2014).
-    Currently only implemented for exponential growth.
-    '''
-
-    # Class variable that stores parameters p_ki and p_kimj
-    # These are independent of the demographic model.
-    fu_tensors = {}
-
-    def __init__(self, sample_size, growth_rate=None, alpha=None):
-        self.n = sample_size
-
-        if growth_rate and alpha:
-            raise ValueError('May not specify both growth rate and alpha.')
-
-        # For exponential growth, use Rafajlovic eqns B.1-B.3
-        elif growth_rate is not None:
-            # If already calculated, retreive fu tensors
-            # If not, calculate them and store them.
-            try:
-                p_ki, p_kimj = TwoSFS.fu_tensors[self.n]
-            except KeyError:
-                print('Calculating...')
-                p_ki = fu.marginal_leaf_prob(self.n)
-                p_kimj = fu.joint_leaf_prob(self.n)
-                TwoSFS.fu_tensors[self.n] = (p_ki, p_kimj)
-            # Calculate first and second moments of coalescence times.
-            t_k = time_first_moments(self.n, growth_rate)
-            t_km = time_second_moments(self.n, growth_rate)
-
-            # Rafajovic B.1
-            k = np.arange(2, self.n+1)
-            self.sfs = np.dot(p_ki, k * t_k) / 2
-            # Rafajovic B.2
-            count_factor = k[:,None]*k[None,:] - np.diag(k)
-            # TODO: Check order of dot product
-            self.two_sfs = np.dot(p_kimj, count_factor * t_km) / 4
-
-        # For beta coalescent, use Birkner eqns ???
-        elif alpha is not None:
-            # TODO:
-            raise ValueError('Beta-coalescent not implemented yet.')
-        else:
-            raise ValueError('Must specify either growth_rate or alpha.')
-
-    def get_sfs(self, folded=False):
-        if not folded:
[…]

### 2018-10-31 src/write_fastNeutrino_input.py

@@ -1,32 +1,0 @@
-import numpy as np
-import argparse
-
-parser = argparse.ArgumentParser(description="Write fastNeutrino input files with constant-N epochs")
-parser.add_argument('ancestral_size', metavar='N_anc', type=float,
-                    help='Ancestral population size')
-parser.add_argument('num_epochs', metavar='n', type=int,
-                    help='Number of constant-size epochs (default=1)')
-parser.add_argument('--spacing', type=str, default='free',
-                    help='Spacing of the epoch boundaries')
-
-args = parser.parse_args()
-
-# One epoch is the ancestral state.
-n = args.num_epochs-1
-
-# TODO: Make these flexible
-start = 10
-end = 10*args.ancestral_size
-
-if args.spacing == 'free':
-    times = ['?t{}'.format(i) for i in range(n)]
-elif args.spacing == 'fixed':
-    times = np.logspace(np.log10(start), np.log10(end), n)
-
-# First line of the input file is the ancestral size.
-print(args.ancestral_size)
-for i, t in enumerate(times):
-    print('c', '?n{}'.format(i), t)
-# There aren't any constrained parameters.
-print(0)
-print(0)

### 2018-11-01 src/helpers.py
Clean up figure-making notebooks.

@@ -1,6 +1,10 @@
 import numpy as np
 import gzip
 
+def tajimasD(n, eta):
+    theta_pi = sfs2pi(eta, n)
+    theta_w  = np.sum(eta) / np.sum(1/np.arange(1,n))
+    return (theta_pi - theta_w) / theta_pi
 
 def pairwise_diversity(allele_counts, sample_size):
     f = allele_counts / sample_size

### 2018-11-01 src/mac2sfs.py
Zip 4Dsites file, include it, and handle that it's zipped.

@@ -1,5 +1,6 @@
 import argparse
 import numpy as np
+import gzip
 import helpers as h
 from scipy.stats import hypergeom
 
@@ -45,7 +46,7 @@ n_ds = args.n_downsample
 # Strip off the "chr" from chromosme names.
 chrom = args.chromosome[3:]
 # Import 4-fold degenerate sites
-with open(args.fourfold_file) as infile:
+with gzip.open(args.fourfold_file) as infile:
     # Convert 4d sites from one-indexed to zero-indexed.
     fourDsites = [int(line.split()[1]) - 1 for line in infile
                     if line.startswith(chrom)]

### 2018-11-01 src/helpers.py
Remove one commented line from helpers.py

@@ -6,6 +6,7 @@ def tajimasD(n, eta):
     theta_w  = np.sum(eta) / np.sum(1/np.arange(1,n))
     return (theta_pi - theta_w) / theta_pi
 
+
 def pairwise_diversity(allele_counts, sample_size):
     f = allele_counts / sample_size
     return 2*f*(1-f) / (1 - (1/sample_size))
@@ -23,7 +24,6 @@ def sfs2pi(sfs, sample_size):
 
 
 def cross_correlation(X, Y, lim):
-    # return np.correlate(X, Y[:-lim], 'valid') / (X.shape[0] - lim)
     n_comps = X.shape[0] - lim
     xy = np.correlate(X, Y[:-lim], 'valid') / n_comps
     x = np.convolve(X, np.ones(n_comps), 'valid') / n_comps

### 2018-11-01 src/jsfs.py
Use installed msprime.

@@ -1,7 +1,5 @@
 import sys
 import numpy as np
-#FIXME: this is a hack to use my local version of msprime
-sys.path.insert(1, '/users/danielrice/msprime-lambda/')
 import msprime
 from collections import deque
 
@@ -146,7 +144,7 @@ if __name__ == "__main__":
 
     population_configurations = [msprime.PopulationConfiguration(sample_size=nSamples, multiple_merger_para=alpha)]
 
-    simulations = msprime.simulate(num_replicates=nLoci, recombination_rate=r, 
+    simulations = msprime.simulate(num_replicates=nLoci, recombination_rate=r,
                                     population_configurations=population_configurations)
     t_test = 0
     t_naive = 0

### 2018-11-01 src/simulate_joint_sfs.py

@@ -1,15 +1,6 @@
 import sys
 import numpy as np
 import argparse
-
-#FIXME: this is a hack to use my local version of msprime
-# sys.path.insert(1, '/users/danielrice/msprime-lambda/')
-# sys.path.insert(1, '/home/dpr/mmc_genomics/src/msprime-lambda/')
-import os
-module_path = os.path.abspath(os.path.join('src/msprime-lambda/'))
-if module_path not in sys.path:
-    sys.path.append(module_path)
-
 import msprime
 import jsfs
 

### 2018-11-01 src/alignment2mac.py
Comment SLiM-parsing scripts.

@@ -2,6 +2,10 @@ import sys
 import argparse
 from collections import Counter
 
+'''
+Calculate the minor allele count from DPGP3 sequence alignments.
+'''
+
 parser = argparse.ArgumentParser()
 parser.add_argument("nSamples", type=int, help="number of samples to subsample to. Must be less than or equal to the number of samples in the file")
 args = parser.parse_args()

### 2018-11-01 src/alleles2correlations.py

@@ -2,6 +2,12 @@ import sys
 import numpy as np
 import helpers as h
 
+'''
+Process SLiM output:
+Take windowed site frequency spectra and
+calculate the correlations between high- and low-frequency alleles
+'''
+
 filename = sys.argv[1]
 n_samples = int(sys.argv[2])
 # The total length of the simulated chromosome

### 2018-11-01 src/merge_corr.py

@@ -2,6 +2,10 @@ import sys
 import gzip
 import numpy as np
 
+'''
+Merge correlations calculated from different SLiM runs into one file.
+'''
+
 n_samples = int(sys.argv[1])
 # The total length of the simulated chromosome
 L_str = sys.argv[2]
@@ -58,7 +62,7 @@ for fn in filenames:
             hihi_corr[i,:] += y
 
 sfs /= n_files
-pi_corr /= n_files 
+pi_corr /= n_files
 lolo_corr /= n_files
 lohi_corr /= n_files
 hihi_corr /= n_files

### 2018-11-01 src/parse_slim.py

@@ -2,6 +2,11 @@ import sys
 import numpy as np
 import helpers as h
 
+'''
+Parse SLiM output by calculating the SFS in windows.
+Windowing greatly reduces the time to compute correlations because mutations are sparse.
+'''
+
 infn = sys.argv[1]
 L_str = sys.argv[2]
 w_str = sys.argv[3]

### 2018-11-01 src/fastNeutrino2params.py
Comment simulation scripts.

@@ -5,6 +5,10 @@ if module_path not in sys.path:
     sys.path.append(module_path)
 from demographicmodel import DemographicModel
 
+'''
+Read a fastNeutrino model file and output the flags to run simulate_joint_sfs.py
+'''
+
 script, fn = sys.argv
 model = DemographicModel(fn)
 model.rescale()

### 2018-11-01 src/simulate_joint_sfs.py

@@ -4,6 +4,11 @@ import argparse
 import msprime
 import jsfs
 
+'''
+Run msprime simulations from the command line.
+Output the average SFS and the 2-SFS at the extreme ends of the locus.
+'''
+
 # Constant for the length of time for a bottleneck to last
 BOTTLENECK_DURATION = 1e-6
 

### 2018-11-01 src/fu.py
Clean up and comment source files.

@@ -1,27 +1,0 @@
-import numpy as np
-from scipy.special import binom
-
-'''
-Implement formulas for the moments of the number of leaves subtended by branches
-at different levels from Fu (1995) as reproduced in Rafajlovic et al. (2014).
-'''
-
-def marginal_leaf_prob(n):
-    '''
-    Calculate the probability that a branch at level k has i leaves.
-    Return as an (n-1)x(n-1) array.
-    '''
-    k = np.arange(2, n+1)
-    i = np.arange(1, n)
-    num = (k[None,:]-1) * binom(n-k[None,:], i[:,None]-1)
-    den = i * binom(n-1, i)
-    return num / den[:,None]
-
-def joint_leaf_prob(n):
-    '''
-    Calculate the joint probability that a branches at level k and m
-    have i and j leaves respectively.
-    Return as an (n-1)^4 array.
-    '''
-    p_kimj = np.zeros((n-1,n-1,n-1,n-1))
-    return p_kimj

### 2018-11-01 src/merge_seq.py

@@ -1,8 +1,12 @@
 import sys
 import tarfile
 
+'''
+    Combine the consensus sequences into a single file for one chromosome
+'''
+
 # We will read the files in chunks of size CHUNKSIZE bytes
-CHUNKSIZE=4096 
+CHUNKSIZE=4096
 
 #TODO: argparse
 tar_fn = sys.argv[1]

### 2018-11-01 src/subsample_slim_output.py

@@ -1,5 +1,9 @@
 import sys
 
+'''
+Take SLiM output with n_old individuals and randomly sample n_new individuals.
+'''
+
 script, n_old, n_new = sys.argv
 n_old = int(n_old)
 n_new = int(n_new)
@@ -14,7 +18,7 @@ if n_old == n_new:
     exit()
 
 
-line = infile.readline() 
+line = infile.readline()
 # Re-print header
 while not line.startswith('#OUT'):
     outfile.write(line)

### 2018-11-01 src/zivkovic.py
Comment zivkovic.py

@@ -4,6 +4,9 @@ from numpy.polynomial.laguerre import laggauss
 from math import factorial
 from functools import partial
 
+'''
+Impliments the formulas in Zivkovic et al. for calculating second moments of coalescent times
+'''
 LAGDEGREE = 40
 
 def zivkovic_alpha(n):

### 2018-11-02 manuscript/manuscript.tex
Polish the manuscript.

@@
 ]{biblatex}
+
+\newcommand{\beginsupplement}{%
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
 
@@
 \title{Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra}
-\author[1]{Daniel P. Rice}
-\author[1,*]{John Novembre}
-\author[2,*]{Michael M. Desai}
+\author[1,2]{Daniel P. Rice}
+\author[1,2,*]{John Novembre}
+\author[3,4,*]{Michael M. Desai}
 \affil[1]{Department of Human Genetics, University of Chicago, Chicago, IL}
-\affil[2]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
+\affil[2]{Department of Ecology and Evolution, University of Chicago, Chicago, IL}
+\affil[3]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
+\affil[4]{Department of Physics, Harvard University, Cambridge, MA}
 \affil[*]{These authors contributed equally to this work.}
@@
 This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
+Finally, we demonstrate a graphical model-checking procedure based on the pointwise mutual information using genomic diversity data from \textit{Drosophila melanogaster}.
 
@@
 
-While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
-For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
-In a similar vein, \cite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
+While the example above is extreme, it is well established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
+For example, \textcite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
+In a similar vein, \textcite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
 Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.
@@
 One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
-To this end, \cite{BirknerEtAl2013, BlathEtAl2016} and \cite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
-\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
-\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
+To this end, \textcite{BirknerEtAl2013, BlathEtAl2016} and \textcite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
+Further, \textcite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
+In a related approach, \textcite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
 
@@
 The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
-\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
-The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
+\textcite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
+The non-mononotic SFS identified by \textcite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
 High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.
 
-Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
-These statistics may be calculated efficiently from genomic single nucleotide variant data.
+Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018}---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+These statistics may be calculated efficiently from genomic single-nucleotide-variant data.
 Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
@@
 
-Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
+Following the notation of \textcite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
 In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
@@
 (Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
-The 2-SFS has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and by \cite{Xie2011a} in a model with selection.
+The 2-SFS has been studied for a non-recombining sites by \textcite{FerrettiEtAl2018} in a neutral model and by \textcite{Xie2011a} in a model with selection.
 We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.
@@
 
-\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
+\textcite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite-sites locus under the standard time-homogeneous Kingman coalescent.
 He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \not\in \{i, (n-i)\}$.
@@
 
-\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
+\textcite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
 This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
 Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
-The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
+The calculations of \textcite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6 of \cite{BirknerEtal2013}).
 Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.
@@
 We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
-Finally, we demonstrate the use of $\fpmi$ in a graphical model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
+Finally, we demonstrate a graphical model-checking procedure based on $\fpmi$ using genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.
 
@@
 
-We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
+We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for sites with different counts ($i\neq j$).
 To do so, we define the \textit{frequency pointwise mutual information} as
@@
 \includegraphics[scale=1]{figures/schematic.pdf}
-\caption{Summarizing population genetic data with frequency pointwise information ($\fpmi$).
+\caption{Summarizing population genetic data with frequency pointwise mutual information ($\fpmi$).
 First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations.
@@
 First, because $\fpmi$ is based on the minor allele frequencies, we may compute it without knowing the ancestral allele.
-Second, the results of \cite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
+Second, the results of \textcite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
 Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
@@
 With finite data, estimates of the 2-SFS will be noisy.
-This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
+This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman coalescent \autocite{Fu1995} and faster than this in models with population size growth or multiple mergers.
 We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
@@
 We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$.
-For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
-For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.
+For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \textcite{ZivkovicWiehe2008}.
+For the beta coalescent, we implemented the recursion described by \textcite{BirknerEtAl2013}.
 
-Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
+Functions for computing the branch-length moments are implemented in \texttt{python} and available in the \texttt{git} repository for this project (\url{https://github.com/dp-rice/multiplemergers}).
 Our Kingman coalescent code can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
-The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \cite{GriffithsTavare1994}.
+The formulas of \textcite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \textcite{GriffithsTavare1994}.
 Thus, these formulas are only practical for sample sizes up to $n \approx 40$.
-The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.
+The recursion of \textcite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.
 
@@
 
-We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2016} capable of multiple mergers, based on modifications made by Joe Zhu [GITHUB].
-This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].
+We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2016} capable of multiple mergers, based on modifications made by Joe Zhu (\url{https://github.com/shajoezhu}).
+This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available at \url{https://github.com/dp-rice/multiplemergers}.
 
-For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
+For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per-basepair recombination rate $r$ resulting in a per-locus total recombination rate $d\,r$.
 (See SI for parameter combinations.)
@@
 In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
-Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and per basepair mutation rate $\mu = 10^{-7}$.
+Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per-basepair $r = 10^{-8}$ and per-basepair mutation rate $\mu = 10^{-7}$.
 We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
@@
 
-In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
+In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial, $f_{\text{sel}}$, over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
 For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.
 
-\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].
+\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available at \url{https://github.com/dp-rice/multiplemergers}.
 
@@
 
-We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
-The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \cite{LackEtAl2015}.
+We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \textcite{LangleyEtAl2011}.
+The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \textcite{LackEtAl2015}.
 We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
@@
 We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
-Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
+Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres, in agreement with calculations by \textcite{Corbett-DetigEtAl2015}.
 Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
@@
 In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites with fewer than 90 of the 100 genotypes called.
-This leaves over 90\% of all sites and without substantially altering the fraction of polymorphic sites (Table~\ref{tab:called_sites}).
+This leaves over 90\% of all sites and does not substantially alter the fraction of polymorphic sites (Table~\ref{tab:called_sites}).
 
 We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
-Following \cite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
+Following \textcite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
 \begin{align}
@@
 estimating both change-points ($t_1 < t_2$) and population sizes ($N_1$, $N_2$).
-We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
+We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \textcite{RagsdaleGutenkunst2017}.
 For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
@@
 In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances between 3~bp and 5~Kb.
-Because we are using fourfold degenerate sites, we only computed the 2-SFS for distances that are multiples of three.
+Because we are using fourfold degenerate sites, we only computed the 2-SFS for distances that are multiples of three basepairs.
 For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
-We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
+We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per-basepair per generation \autocite{ComeronEtAl2012}.
 We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
 For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
-For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightleyEtAl2014}.
+For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per-basepair per-generation \autocite{KeightleyEtAl2014}.
 These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.
 
-A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available [GITHUB].
+A \texttt{snakemake} pipeline, \texttt{python} scripts, and \texttt{jupyter} notebooks to replicate our data processing, model fitting, simulations, and analysis of the DPGP3 data are available at \url{https://github.com/dp-rice/multiplemergers}.
 
@@
 
-We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
+We first compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
 We are interested in whether $\fpmi$ can distinguish beta from Kingman coalescent models that produce similar distortions in the SFS.
-To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant to changes in the average pairwise coalescence time: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson1975}.
+To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant to changes in the average pairwise coalescence time: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta \autocite{Watterson1975}.
 Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
-All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).
+All results are computed numerically using the results of \textcite{Fu1995}, \textcite{ZivkovicWiehe2008}, and \textcite{BirknerEtAl2013} (see Methods).
 
@@
 
-As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
+As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for the constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
 Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS, $D = -0.4$.
@@
 To generalize this finding, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
-\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
+\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but remains negative.
 On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
@@
 
-To measure the effect of recombination on $\fpmi$, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
-In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
+To measure the effect of recombination on $\fpmi$, we ran coalescent simulations using a version of the program \texttt{msprime} \autocite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
+In particular, we simulated a model where the marginal trees are generated by the beta coalescent process, as in our numerical calculations above.
 In this model, marginal genealogies will follow the beta coalescent distribution, and the average SFS will be given by the formula in \cite{BirknerEtAl2013}.
@@
 
-\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
+\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances between sites in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
 As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
@@
 The first three rows show coalescent simulations in \texttt{msprime}.
-Exponential growth rate $g$ varies logarithmically from 0.01 to 8.0 in coalescent time units.
-Piecewise constant change-points, range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2.
+Exponential growth rate $g$ varies logarithmically from 1 to 8.0 in coalescent time units.
+Piecewise constant change-points range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2.
 The fourth row shows forward-time \texttt{SLiM} simulations with selective sweeps at linked sites (beneficial mutation fraction $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$, see Methods).
@@
 \Fig{fig:hilo_vs_d} also suggests that the most informative genomic distance for distinguishing multiple mergers from population growth is $d\,r \E{T_2} \sim 1$.
-In a later section, we will demonstrate how to use these results to implement a graphical model checking procedure based on plotting $\hilopmi$ as a function of genomic distance.
+In a later section, we will demonstrate how to use these results to implement a model checking procedure based on plotting $\hilopmi$ as a function of genomic distance.
 
@@
 There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
-This behavior relates to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
-Further, \cite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
+This behavior relates to the results of \textcite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
+Further, \textcite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
 Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
@@
 That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
-The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
+The primary difference between the forward-time sweeps and beta coalescent simulations is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
 This decay is expected because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.
@@
     \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
-    \item Compare true to predicted statistics. Evaluate model fit.
+    \item Compare true to predicted statistics to evaluate model fit.
 \end{enumerate}
@@
 
-In this section, we illustrate the procedure outlined above by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
+In this section, we illustrate the procedure outlined above by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \autocite{LackEtAl2015}.
 The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be mostly free of cosmopolitan admixture.
@@
 The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
-In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\Pi \mu)$ is the distance scale corresponding to $d_c r \E{T_2} = 1$.
+In contrast, the second row shows the expected and observed weighted $\fpmi$ averaged over distances less than 200~bp apart, which corresponds to to $d/d_c < 15$.
 Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except adjacent to the diagonal.
@@
 We demonstrated a model-checking procedure on data from \textit{D.
-melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth alone can not explain the positive associations between high and low frequency mutations.
+melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth alone cannot explain the positive associations between high and low frequency mutations.
 
@@
 In the latter model, we vary both the fold-change in $N$ and the time of the change.
-Like Tajima's D, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
+Like Tajima's D, the singleton/doubleton ratio measures distortion in the SFS relative to the constant-$N$ Kingman coalescent.
 (In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
@@
 We focus here on demographic inference methods that explicitly use the Kingman coalescent model for calculations.
-However, another popular class of methods are based on forward-time models, such as the Wright-Fisher model (e.g., \cite{GutenkunstEtAl2009}, \cite{SheehanEtAl2013}).
+However, another popular class of methods are based on forward-time models, such as the Wright-Fisher model (e.g., \cite{GutenkunstEtAl2009, SheehanEtAl2013}).
 We believe that $\fpmi$ is useful for model-checking with these methods as well.
@@
 
-In this paper, we have outlined a graphical model-checking procedure in the spirit of \cite{Anscombe1973}.
-One could extend our work to implement a formal hypothesis-testing framework for rejecting the best-fit Kingman coalescent model according to some inference method.
+In this paper, we have outlined a graphical model-checking procedure in the spirit of \textcite{Anscombe1973}.
+One could extend our work to implement a formal hypothesis-testing framework for rejecting the Kingman coalescent model best-fit according to some inference method.
 Numerical calculations would require higher moments of the branch-length distribution, which may be computationally intractable for large samples.
@@
 However, the precise correspondence between this model and any particular forward-time process is unclear.
-In order to compare with our numerical calculations based on \cite{BirknerEtAl2013}, we use the beta coalescent for the distribution of marginal genealogies, but the explicit forward-time models that have been studied by others generate simultaneous multiple mergers.
+In order to compare with our numerical calculations based on \textcite{BirknerEtAl2013}, we use the beta coalescent for the distribution of marginal genealogies, but the explicit forward-time models that have been studied by others generate simultaneous multiple mergers.
 (See e.g., the \cite{DurrettSchweinsberg2005} model of selective sweeps, but note that our forward-time simulations suggest that this distinction is not important to our main results.)
@@
 
-There are two interesting potential empirical applications of our work: assessing the evidence for variation in multiple mergers coalescence within genomes and between species.
+An interesteing potential future empircal application of our work would be to use $\fpmi$ to assess the evidence for variation in multiple mergers coalescence within genomes and between species.
 For example, one could compute the $\fpmi$ in different regions of a large genome and look for a relationship between the strength of non-Kingman coalescence and genomic properties such as the recombination rate and functional density.
-Alternatively, one could survey multiple species using a data set such as the diversity data compiled by \cite{Corbett-DetigEtAl2015}.
+Alternatively, one could survey multiple species using a data set such as the diversity data compiled by \textcite{Corbett-DetigEtAl2015}.
 Either would reveal new information about the suitability of population genetic models, and the forces that determine genetic diversity.
@@
 DPR was supported by the Chicago Fellows Program of the University of Chicago.
-[OTHER FUNDING].
+J.N. acknowledges support for this work from NIH grants GM108805 and HG007089.
+M.M.D. acknowledges support from the Simons Foundation (Grant 376196), grant DEB-1655960 from the NSF, and grant GM104239 from the NIH.
 This work was completed in part with resources provided by the University of Chicago Research Computing Center and Harvard Faculty of Arts and Sciences Research Computing Center.
@@
 \clearpage
-
+\beginsupplement
 \section*{Supplementary Figures and Tables}
@@
 \end{document}
-
-TODO:
-- Text
-    - Abstract
-    - Consistency of nomenclature, labeling, etc.
-    - Insert links to github
-    - Get rest of citations
-    - Fix bibtex issues
-    - Copy edit text
-    - Copy edit bibliography
-- Figures
-    - Organize panels
-    - Remake figure 5 that got deleted
-    - Rename files
-    - Figure sizes (1-column vs 2-column)
-    - Typeface and font sizes
-    - Color schemes and color maps
-    - Consistency of plotted values (esp. pmi vs cov, base of log scales)
-- Notebooks
-    - Single notebook to make all non-data figures.
-    - Single notebook to do all DPGP3 exploratory and final analysis
-    - Reduce redundancy. Decide which helper functions belong in notebooks and which go in modules.
-    - Remove old/unused notebooks from repo.
-- Simulation output
-    - Figure out which files are necessary to reproduce figures
-    - Make tarball with simulations and include in repo (if reasonable size)
-- DPGP3 data
-    - Decide whether we need to post any processed data.
-- Code
-    - Conda environment and environment file
-    - Clean up snakefiles
-    - Compute 2-epoch moments exactly
-    - Organize modules, reduce redundancy
-    - Refactor functions into a consistent style
-    - Document
-    - Figure out what to do about msprime code
-    - Post on Github
-
-Left out of Discussion:
-Mention methods to fit multiple mergers coalescent to data (computationally difficult).
-Not really what we're trying to do here.
-Our goal is that you can exploit this extra dimension of the data you already have to check your model and make sure you're not fitting a signal of selection, etc with a demographic model.
-
-We also don't mention single locus methods.
-
-Things that were done but not included in this ms:
-- Background selection (didn't run the right parameters to see the effect. Might be computationally challenging.)
-- 2-Deme model with migration. (Could include in SI.)
-- 1-D stepping-stone.
-
-OLD TEXT:
-
-% Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
-
-% There are two limitations of these existing results.
-% First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
-% Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.
-
-% With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
-% One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
-% The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
-% Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
-% Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
-% Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
-% (Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)

### 2018-11-02 src/zivkovic.py
Add warning when using 2-epoch growth

@@ -250,6 +250,8 @@ def sigma_ij3(n, Ett_kpk):
     return np.fliplr(ret)
 
 def sigma_ij(n, mode, **params):
+    if n > 4 and mode=="two-epoch":
+        raise ValueError('Two-epoch growth has not been tested for n>4!')
     Sigma_i = sigma_i(n, mode, **params)
     Ett_kpk = time_second_moments(n, mode, **params)
     Sigma_ij1 = sigma_ij1(n, Ett_kpk)

### 2018-11-02 src/zivkovic.py
Comment zivkovic functions.

@@ -4,10 +4,78 @@ from numpy.polynomial.laguerre import laggauss
 from math import factorial
 from functools import partial
 
+# Degree of laguerre polynomials for numerical integration
+LAGDEGREE = 40
+
 '''
 Impliments the formulas in Zivkovic et al. for calculating second moments of coalescent times
 '''
-LAGDEGREE = 40
+
+def sigma_i(n, mode, **params):
+    '''
+    Calculate the expected site frequency spectrum.
+
+    Parameters
+    ----------
+    n : int
+        Sample size
+    mode : string
+        Type of model. For now may be "constant", "exponential", or "two-epoch".
+    params : {str : float}
+        Dictionary of parameters, which depend on mode.
+        "constant" takes no parameters.
+        "exponential" takes a growth rate parameter 'g'.
+        "two-epoch" takes two parameters: growth time "tau" and inverse growth factor "f".
+    '''
+    alpha = zivkovic_alpha(n)
+    p_nki = marginal_leaf_prob(n)
+    K = np.arange(n+1)
+    B = binom(K,2)
+    if mode == 'constant':
+        I1_j = 1 / B
+    else:
+        I1_j = laguerre_integral(lambda_inv(mode, **params), B)
+
+    I1_j[np.isinf(I1_j)] = 0
+    I1_j[np.isnan(I1_j)] = 0
+    sign_jk = diagonal_signs(n)
+    alpha_jk = alpha[n]
+    T_k = I1_j @ (sign_jk*alpha_jk)
+    T_k[:2] = 0
+    p_ki = p_nki[n]
+    return (K*T_k) @ p_ki
+
+def sigma_ij(n, mode, **params):
+    '''
+    Calculate the covariance matrix site frequency spectrum.
+
+    Parameters
+    ----------
+    n : int
+        Sample size
+    mode : string
+        Type of model. For now may be "constant", "exponential", or "two-epoch".
+    params : {str : float}
+        Dictionary of parameters, which depend on mode.
+        "constant" takes no parameters.
+        "exponential" takes a growth rate parameter 'g'.
[…]

### 2018-11-02 src/coalescentmoments.py
Comment coalescentmoments.py

@@ -2,6 +2,72 @@ import sys
 import numpy as np
 from scipy.special import beta, binom
 
+'''
+Calculate the site frequncy spectrum first and second moments
+according to Fu 1995 and Eldon et al 2015.
+'''
+
+def sfs_moments(N, alpha, m2=True):
+    '''
+    Calculate first and second moments of the site frequency spectrum.
+
+    Parameters
+    ----------
+    N : int
+        Sample size
+    alpha : float
+        Beta coalescent parameter alpha
+    m2 : Bool
+        If True, return first and second moments.
+        If False, return first moments only
+
+    '''
+    # If Kingman coalescent, use explicit Fu formulas to save time
+    if alpha == 2.0:
+        moments = fu_moments(N)
+        if m2:
+            return moments
+        else:
+            return moments[0]
+    else:
+        Pij = pij(N, alpha)
+        G = g(N, Pij, alpha)
+        Pnkb = pnkb(N, Pij, G)
+        if m2:
+            Peq = peq(N, Pij, G)
+            Pun = pun(N, Pij, G, Pnkb)
+            Pne = pne(N, Pij, G, Pnkb)
+            return sfs_m1(N, G, Pnkb), sfs_m2(N, G, Peq, Pun, Pne)
+        else:
+            return sfs_m1(N, G, Pnkb)
+
+def fold_sfs_moments(N, M1, M2):
+    '''
+    Calculate first and second moments of the folded site frequency spectrum.
+
+    Parameters
+    ----------
+    N : int
+        Sample size
+    M1 : np.array(dtype=float)
+        First moments of the unfolded site frequency spectrum
+    M2 : np.array(dtype=float)
+        Second moments of the unfolded site frequency spectrum
+
+    '''
+
+    M1_folded = np.zeros(N//2)
+    for i in range(N//2):
+         M1_folded[i] = (M1[i] + M1[N-i-2]) / (1.0 + (i==N-i-2))
+
+    M2_folded = np.zeros((N//2, N//2))
+    for i in range(N//2):
+        for j in range(i):
[…]

### 2018-11-03 manuscript/manuscript.tex
Fix biblatex in-text citation issue.

@@
     uniquename=false,
-    maxcitenames=2
+    uniquelist=false,
+    maxcitenames=2,
+    mincitenames=1,
+    maxbibnames=9
 ]{biblatex}
@@
   \begin{center}
-    \caption{Boundaries of the homogeneous central regions of chromosomes (shown in \ref{fig:dpgp_pi}).}
+    \caption{Boundaries of the homogeneous central regions of chromosomes (shown in \fig{fig:dpgp_pi}).}
     \label{tab:central_regions}

### 2018-11-03 manuscript/manuscript.tex
Fix typos in manuscript and make multiple-merger punctuation consistent.

@@
 
-An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} (\cite{Pitman1999, Sagitov1999, DonnellyKurtz1999,}; reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
-Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
-The more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
-Multiple merger and simultaneous multiple merger models are relevant for species with
+An important extension of the Kingman coalescent is a family of models known as \textit{multiple-merger coalescents} (\cite{Pitman1999, Sagitov1999, DonnellyKurtz1999,}; reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
+Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple-merger coalescents permit more than two lineages to coalesce in a single event.
+The more general class of simultaneous-multiple-merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple-merger event at the same time.
+multiple-merger and simultaneous-multiple-merger models are relevant for species with
 ``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
@@
 For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by the rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
-In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.
+In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading, and inferences based on the Kingman coalescent may be qualitatively incorrect.
 
@@
 One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
-To this end, \textcite{BirknerEtAl2013, BlathEtAl2016} and \textcite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
-Further, \textcite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
+To this end, \textcite{BirknerEtAl2013, BlathEtAl2016} and \textcite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous) multiple-merger coalescents.
+Further, \textcite{EldonEtAl2015} showed that it is possible to distinguish between a multiple-merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
 In a related approach, \textcite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.
@@
 The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
-\textcite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
+\textcite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple-merger SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
 The non-mononotic SFS identified by \textcite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
@@
 
-Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018}---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
+Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018}---are useful for distinguishing between the Kingman coalescent with population growth and multiple-merger coalescents.
 These statistics may be calculated efficiently from genomic single-nucleotide-variant data.
@@
 
-\textcite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
+\textcite{BirknerEtAl2013} extended Fu's calculation to a family of multiple-merger coalescents called beta coalescents.
 This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
 Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
-The calculations of \textcite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6 of \cite{BirknerEtal2013}).
+The calculations of \textcite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6 of \cite{BirknerEtAl2013}).
 Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.
@@
 Second, the results of \textcite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
-Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
+Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single-site SFS.
 In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
-We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is largely insensitive to the distribution of coalescence times.
-Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.
+We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single-site spectra, it is largely insensitive to the distribution of coalescence times.
+Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes the multiple-merger coalescent from the Kingman coalescent.
 
@@
 With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
-Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
+Such sweeps will affect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
 Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.
@@
 
-We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \textcite{LangleyEtAl2011}.
+We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from ${\sim}200$ flies, obtained via the haploid embryo method of \textcite{LangleyEtAl2011}.
 The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \textcite{LackEtAl2015}.
@@
 Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
-Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
+Therefore, we selected a 13-16~Mb ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
 The boundary positions of these central regions are given in Table~\ref{tab:central_regions}.
@@
 This behavior relates to the results of \textcite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
-Further, \textcite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
-Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
-We do not expect this behavior to be universal in multiple mergers coalescents with recombination.
+Further, \textcite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple-merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
+Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple-merger events, even at infinite genetic distances.
+We do not expect this behavior to be universal in multiple-merger coalescents with recombination.
 
@@
 
-Various authors have shown that natural selection can generate multiple merger coalescents at linked neutral sites (e.g., \cite{DurrettSchweinsberg2005, CoopRalph2012, NeherHallatschek2013, DesaiEtAl2013, SegerEtAl2010}).
+Various authors have shown that natural selection can generate multiple-merger coalescents at linked neutral sites (e.g., \cite{DurrettSchweinsberg2005, CoopRalph2012, NeherHallatschek2013, DesaiEtAl2013, SegerEtAl2010}).
 However, our simulations of the beta coalescent with recombination are of an explicitly neutral model.
@@
 Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies higher.
-If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) so that $\E{\eta_{12}} > \E{\eta_{1}} \E{\eta_{2}}$.
+If multiple-merger events are frequent enough, this effect may dominate the tradeoff between (A) and (B) so that $\E{\eta_{12}} > \E{\eta_{1}} \E{\eta_{2}}$.
 
@@
 
-Our coalescent simulations implement a particular version of the multiple mergers coalescent with recombination.
+Our coalescent simulations implement a particular version of the multiple-merger coalescent with recombination.
 However, the precise correspondence between this model and any particular forward-time process is unclear.
@@
 (See e.g., the \cite{DurrettSchweinsberg2005} model of selective sweeps, but note that our forward-time simulations suggest that this distinction is not important to our main results.)
-Moreover, the long-range correlations observed in \fig{fig:hilo_vs_d} likely depend on an implicit choice regarding the scaling of recombination, coalescent, and multiple merger rates \autocite{EldonWakeley2006}.
+Moreover, the long-range correlations observed in \fig{fig:hilo_vs_d} likely depend on an implicit choice regarding the scaling of recombination, coalescent, and multiple-merger rates \autocite{EldonWakeley2006}.
 These issues are poorly understood, and more theoretical work is required to understand the interactions between multiple mergers and recombination.
 
-An interesteing potential future empircal application of our work would be to use $\fpmi$ to assess the evidence for variation in multiple mergers coalescence within genomes and between species.
+An interesteing potential future empircal application of our work would be to use $\fpmi$ to assess the evidence for variation in multiple-merger coalescence within genomes and between species.
 For example, one could compute the $\fpmi$ in different regions of a large genome and look for a relationship between the strength of non-Kingman coalescence and genomic properties such as the recombination rate and functional density.
@@
 \subsection*{Acknowledgements}
-DPR was supported by the Chicago Fellows Program of the University of Chicago.
+D.P.R. was supported by the Chicago Fellows Program of the University of Chicago.
 J.N. acknowledges support for this work from NIH grants GM108805 and HG007089.
@@
 \caption{Average pairwise diversity, $\Pi$, of 100 haploid samples from DPGP3 panel.
-Each point represents the average over a 10~KB window.
+Each point represents the average over a 10~Kb window.
 Vertical lines show the boundaries of the central regions defined in Table~\ref{tab:central_regions}.
@@
     \begin{tabular}{l|r|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      \textbf{Chr.} & \textbf{Start position (MB)} & \textbf{End position (MB)}\\
+      \textbf{Chr.} & \textbf{Start position (Mb)} & \textbf{End position (Mb)}\\
       \hline

### 2018-11-03 manuscript/manuscript.tex
Fix missing last sentence of abstract.

@@
 Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than on their branch lengths and is therefore largely insensitive to population size change.
-This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
-Finally, we demonstrate a graphical model-checking procedure based on the pointwise mutual information using genomic diversity data from \textit{Drosophila melanogaster}.
+This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.
+Finally, we demonstrate a graphical model-checking procedure based on the pointwise mutual information using genomic diversity data from \textit{Drosophila melanogaster}.}
 

### 2019-03-11 manuscript/manuscript.tex

@@
 \title{Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra}
-\author[1,2]{Daniel P. Rice}
-\author[1,2,*]{John Novembre}
-\author[3,4,*]{Michael M. Desai}
-\affil[1]{Department of Human Genetics, University of Chicago, Chicago, IL}
-\affil[2]{Department of Ecology and Evolution, University of Chicago, Chicago, IL}
-\affil[3]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
-\affil[4]{Department of Physics, Harvard University, Cambridge, MA}
-\affil[*]{These authors contributed equally to this work.}
+\author[$\ast$,$\dagger$,1]{Daniel P. Rice}
+\author[$\ast$,$\dagger$,2]{John Novembre}
+\author[$\ddager$,$\S$,2]{Michael M. Desai}
+\affil[$\ast$]{Department of Human Genetics, University of Chicago, Chicago, IL}
+\affil[$\dagger$]{Department of Ecology and Evolution, University of Chicago, Chicago, IL}
+\affil[$\ddager$]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
+\affil[$\S$]{Department of Physics, Harvard University, Cambridge, MA}
+\affil[1]{Corresponding author: 920 E 58th St, Chicago, IL 60637, \url{dpr@uchicago.edu}}
+\affil[2]{These authors contributed equally to this work.}
 \date{\today}
@@
 Finally, we demonstrate a graphical model-checking procedure based on the pointwise mutual information using genomic diversity data from \textit{Drosophila melanogaster}.}
+
+\textbf{Keywords:} multiple-mergers coalescent, demographic inference, model checking, site frequency spectrum, mutual information
+
+\textbf{Short Title:} Detecting multiple mergers using 2-SFS
 
